### 2016-08-11 doc/paper.tex
Some lifting changes

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Automatically Lifting Software Analyses to Product Lines}
\author{Ramy Shahin}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

\section{First section}

Your text goes here.

\subsection{A subsection}

More text.

\end{document}

### 2016-08-11 LTS.hs

@@ -4,18 +4,20 @@
 -- Ramy Shahin - July 10th 2016
 -------------------------------------------------------------------------------
 module LTS where
-import Data.Graph
 import Data.List
-import Variability
 
 -- abstract State type
-type State = Vertex
-
--- abstract Transition type
-type Transition = Edge
+type State = Int
 
 -- abstract Action type
 type Act = Int
+
+-- abstract Transition type
+data Transition = Transition {
+    source :: State,
+    target :: State,
+    action :: Act
+    } deriving (Show)
 
 -- abstract abstract proposition type
 type AP = Int
@@ -28,39 +30,34 @@ type AP = Int
 --      TODO: AP is a set of atomic propositions
 --      TODO: L is a a labeling function, mapping states to sets of propositions (AP)
 data LTS = LTS {
-    getGraph    :: Graph,
-    getActions  :: [Act],
-    getInitStates :: [State] 
+    getStates       :: [State],
+    getTransitions  :: [Transition],
+    getActions      :: [Act],
+    getInitStates   :: [State] 
     -- TODO: [AP] 
     -- TODO: (Table [AP])
     } deriving (Show)
-
--- fine-grained accessors
-getStates :: LTS -> [State]
-getStates = vertices . getGraph
-
-getTransitions :: LTS -> [Transition]
-getTransitions = edges . getGraph
-
+ 
 -------------------------------------------------------------------------------
 -- LTS Algorithms
 -------------------------------------------------------------------------------
 
+neighbors :: [Transition] -> State -> [State]
+neighbors ts s = [(target t) | t <- ts, (source t) == s]
+
 -- Depth-first Search
-dfs_ :: [Transition]    ->      -- graph edges
-        --[State]         ->      -- path prefix so far from source (recursively computed)
+dfs ::  [Transition]    ->      -- graph edges
         State           ->      -- target node
         State           ->      -- source node 
         [State]                 -- returns the path from source to target
         
-dfs_ edges target src =
-    let neighbors   t = [x | (s, x) <- edges, s == t]
-        dfs_rec     pathPrefix st = 
+dfs edges target src =
+    let dfs_rec     pathPrefix st = 
             if st == target 
             then 
                 pathPrefix ++ [target] 
             else 
-                let unvisitedNeighbors = (neighbors st) \\ pathPrefix
+                let unvisitedNeighbors = (neighbors edges st) \\ pathPrefix
                     paths = map (dfs_rec (pathPrefix ++ [st])) unvisitedNeighbors 
[…]

### 2016-08-11 LTS_test.hs

@@ -8,17 +8,16 @@
 module LTS_test where
 import Test.QuickCheck
 import LTS
-import Data.Graph
-import Graph_test
-
+import Data.List
+{-
 -- Generators
 instance Arbitrary LTS where
     arbitrary = do
-       graph <- arbitrary
-       actions <- arbitrary
-       let states = vertices graph
-       initStates <- sublistOf states
-       return (LTS graph actions initStates)
+       states       <- arbitrary
+       transitions  <- [Gen (x,y) | x <- sublistOf states, y <- sublistOf states]
+       actions      <- arbitrary
+       initStates   <- sublistOf states
+       return (LTS (nub states) (nub transitions) (nub actions) (nub initStates))
       
 -- helper functions      
 notEmpty = not . null
@@ -64,3 +63,4 @@ return [] -- need this for GHC 7.8
 
 
 runTests = $quickCheckAll
+-}

### 2016-08-11 LiftedLTS.hs

@@ -1,0 +1,31 @@
+-------------------------------------------------------------------------------
+-- LiftedLTS.hs
+-- Lifted Labeled Transition System (LTS) library
+-- Ramy Shahin - July 20th 2016
+-------------------------------------------------------------------------------
+module LiftedLTS where
+import LTS
+import Data.List
+import Data.Tree
+import Variability
+
+-- abstract State type
+type StateLifted = Lifted State
+
+type TransitionLifted = Lifted Transition
+
+consLiftedTransition :: StateLifted -> StateLifted -> Act -> PresenceCondition -> TransitionLifted
+consLiftedTransition s0 s1 a pc = apply3 pc Transition s0 s1 (liftValue a)
+
+{-
+getSource :: TransitionLifted -> StateLifted
+getSource t = [(s0,pc) | ((s0,_,_),pc) <- t]
+
+getTarget :: TransitionLifted -> StateLifted
+getSource t = [(s1,pc) | ((_,s1,_),pc) <- t]
+
+getAction :: TransitionLifted -> StateLifted
+-}
+
+neighborsLifted :: [TransitionLifted] -> StateLifted -> [StateLifted]
+neighborsLifted ts s = [(target t) | t <- ts, apply2 (==) (source t) s]

### 2016-08-11 LiftedLTS_test.hs

@@ -1,0 +1,12 @@
+-------------------------------------------------------------------------------
+-- LiftedLTS_test.hs
+-- Lifted Labeled Transition System (LTS) unit tests
+-- Ramy Shahin - August 10th 2016
+-------------------------------------------------------------------------------
+{-# LANGUAGE TemplateHaskell #-}
+
+module LiftedLTS_test where
+import Test.QuickCheck
+import LiftedLTS
+import Data.List
+

### 2016-08-11 Variability.hs

@@ -9,15 +9,22 @@ module Variability where
 type Expr = Bool
 
 type PresenceCondition  = Expr
-type SPLContext         = Expr
+--type SPLContext         = Expr
+
+conj :: [PresenceCondition] -> PresenceCondition
+conj a = True     -- TODO
+
+disj :: Expr -> Expr -> Expr
+disj a b = True     -- TODO
 
 sat :: Expr -> Bool
 sat ex = True
 
-data SPLVariable a = 
-    SPLVariable a PresenceCondition
-    deriving (Eq) 
+type SPLOption a = (a, PresenceCondition)
 
+--instance Eq SPLOption a where
+--    (==) a b = (getValue a == getValue b) && sat(getPresenceCondition a && getPresenceCondition b)
+    
 -- when lifting a product value to a product line value, we might end up with
 -- different values for each product in the product line. This is why a value is
 -- lifted into a set of values, each with a path condition. An important
@@ -27,11 +34,51 @@ data SPLVariable a =
 -- to the same set of products). This does not affect correctness, but severely
 -- affects performance as we are now degenerating into brute force analysis
 -- across all possible products.
-type SPLLifted a = [SPLVariable a]
+type Lifted a = [SPLOption a]
+    
+-- lift a value
+liftValue :: a -> Lifted a
+liftValue x = [(x, True)]
 
-get :: SPLContext -> SPLVariable a -> Maybe a
-get cntxt v = case v of
-    SPLVariable val pc -> if sat (cntxt && pc) then Just val else Nothing
+-- join (Lifted-Lifted) - join 2 lifted values
+join2 :: PresenceCondition -> Lifted a -> Lifted b -> Lifted (a,b)
+join2 pc a b = 
+    let xProduct = [((aVal, bVal), (conj [pc, aPC, bPC])) | (aVal, aPC) <- a, (bVal, bPC) <- b]
+    in  filter (\(_, pc) -> sat pc) xProduct
+    
+join3 :: PresenceCondition -> Lifted a -> Lifted b -> Lifted c -> Lifted (a,b,c)
+join3 pc a b c = 
[…]

### 2016-11-25 doc/paper.tex
Variability unit test

@@
 \usepackage{graphicx} % support the \includegraphics command and options
+\usepackage{amsthm}
 
@@
 % These packages are all incorporated in the memoir class to one degree or another...
+
+\usepackage{minted}
 
@@
 
+%%\usepackage{natbib}
+
 %%% END Article customizations
@@
 
-\section{First section}
+\theoremstyle{definition}
+\newtheorem{exmp}{Example}[section]
 
-Your text goes here.
+\section{Introduction}
 
-\subsection{A subsection}
+%Software Product Lines (SPLs) are configurable sets of software artifacts from which different software products can be generated.
+SPLs, pretty much like classic manufacturing product lines, are much more economic in terms of time and effort, compared to developing and maintaining different software product variations separately.
+This is the reason why SPLs are becoming commonly adopted in many software domains, including operating systems, embedded systems, compilers and many others.
 
-More text.
+%SPLs capture product variability in several ways, including annotative mechanisms available in conventional programming languages.
+One of the most commonly used annotative mechanisms is C preprocessor macro definitions and conditional compilation.
+Product features are abstracted in terms of preprocessor macros, and feature selection is done using build-time macro definitions.
+This approach, although ad-hoc compared to more structural approaches provided by SPL language (e.g.
+Featured-oriented languages), has the benefit of not requiring any tools beyond the conventional C toolchain.
+
+\section{Background}
+
+\subsection{Software Product Lines}
+
+\subsubsection{Product Variability}
+
+\subsubsection{Features}
+
+\subsubsection{Presence Conditions}
+Software Product Lines (SPLs) are built on top of both common and variable software artifacts.
+Common artifacts exist in all products of the product line, while variable artifacts exist only in a non-empty subset of valid products.
+That subset is defined using a \emph{Presence Condition}: a propositional formula over product line features.
+
+\subsubsection{Product Generation}
+
+\subsection{Software Analyses}
+
+\subsection{Typed Lambda Calculus}
+
+\section{Lifting Typed Lambda Calculus}
+
+By lifting Typed Lambda Calculus (TLC), we mean providing syntactic transformation rules that lift any program written in TLC by compositionally lifting TLC syntactic constructs.
+Those transformation rules can be applied to any software analysis written in TLC, resulting in an equivalent analysis that can be applied to a Software Product Line.
+
+In the following description of the transformation rules we use Haskell syntax.
+This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as the construction of those constructs on top of TLC is standard \cite{Pierce2002}.
+
+When lifting an analysis from products to product lines, we need to appropriately tag variable artifacts with their corresponding presence conditions.
+We use the \emph{PresenceCondition} abstract data type for this purpose.
+
+\subsection{Lifting Types}
+Given a type \textbf{t}, we lift it into type \textbf{Lifted t} which is:
+
+\begin{minted}{haskell}
+type Lifted t = [(t, PresenceCondition)]
+\end{minted}
+
+where:
+\begin{itemize}
+\item \textbf{t} is a type parameter that can be instantiated with any type
+\item \textbf{(t, PresenceCondition)} is a pair of a value of type \textbf{t} together with a \textbf{PresenceCondition}
+\item \textbf{[(t, PresenceCondition)]} is a list of such pairs
+\item \textbf{Lifted t $\equiv$ [(t, PresenceCondition)]} is a parametric data type, instantiated by passing the type argument \textbf{t} to the type constructor \textbf{Lifted}
+\end{itemize}
+
+A variable of type \textbf{t} in a product line might actually have different values in different products, hence the multiplicity of items in a list.
+The \textbf{PresenceCondition} tags are used to keep track of which values correspond to which products.
+For this to be semantically correct we need to make sure that the sets of products represented by those presence conditions do not overlap, otherwise we might end up with a product variable having different values at the same time.
+This \textbf{disjointness} requirement is formalized in the following invariant:
+
+\newtheorem*{disjInv*}{Disjointness Invariant}
+\begin{disjInv*}
+Given a lifted value $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i \neq j: {unsat (pc_i \wedge pc_j)}$
+\end{disjInv*}
+
+The invariant simply states that a lifted value can not have different values with presence condition that overlap.
+The \emph{unsat} function takes a proposition and returns true if and only if that proposition is unsatisfiable.
+The conjunction of two presence conditions is unsatisfiable if and only if they represent non-overlapping sets of products.
+
+This invariant is a precondition on inputs into the program being lifted, and in the following sections we will show how its validity is maintained by the lifting transformations.
+
+\subsection{Lifting Values}
+
+To lift a value of type \textbf{t}, all we need to is to provide a presence condition for it, and then encapsulate both into a value of type \textbf{Lifted t}.
+This is what the following \textbf{lift} operator does:
+
+\begin{minted}{haskell}
+lift :: PresenceCondition -> t -> Lifted t
+lift pc x = [(x, pc)]
+\end{minted}
+
+This operator can be applied to values of any type, including functions. The importance of lifting function values will be clear in the following subsection.
+
+In many cases we need to provide the most general presence condition, which is \textbf{True} (the set of all products). For this we provide the following \textbf{liftT} operator:
+
+\begin{minted}{haskell}
+liftT :: t -> Lifted t
+liftT x = lift True x
+\end{minted}
+
+Input values to analyses from product lines will typically be of lifted types. For example, an integer can have one value in a set of products, and a different value in another set.
+
+\begin{exmp}[Lifted Values]
+
+Assuming we have two features A and B, variables x and y of type \textbf{Lifted Int} and variable z of type \textbf{Lifted String} can have the following values: \\
+
+x = [(1,A),(-2, $\neg A \wedge B$),(3, $\neg A \wedge \neg B$)]
+
+y = [(5, $A \wedge \neg B$), (4,B), (-3, $\neg A \wedge \neg B$)]
+
+z = [("hello", $\neg A \wedge B$),("helloooo", A)] \\
+
+Both x and y satisfy the disjointness invariant.
+It is worth mentioning that a lifted value does not have to be defined in each product in the product line.
+For example, z is not defined in the product identified by the presence condition $(\neg A \wedge \neg B)$.
+
+\end{exmp}
+
+\subsection{Applying Lifted Functions}
+
+Now that we have lifted values (both data and functions), we need to provide a way to apply lifted functions to lifted operands.
+This is how data propagates through a functional program.
+Starting with the simplest case, where we have a lifted unary function of type \textbf{Lifted (a $\to$ b)} and an argument of type \textbf{Lifted a} we need to provide a way to apply that function to its argument resulting in a \textbf{Lifted b} value.
+Recall that a lifted value of type \textbf{Lifted t} is just a collection of values of type \textbf{t}, each with a corresponding presence condition, and the whole lifted value has to satisfy the disjointness Invariant.
+So to apply a collection of functions to a collection of values, all we need to is to apply each of the functions to each of the values.
+Since the result has to be a lifted value, then we also need a presence condition for each of the computed values.
+This presence condition is simply the conjunction of the presence conditions of the function and the argument used to compute it.
+If the conjunction of the presence conditions is unsatisfiable (i.e.
+a contradiction), it means that the corresponding value does not belong to any product, so it can be safely filtered out.
+This is how the apply function looks in Haskell:
+
+\begin{minted}{haskell}
+apply :: Lifted (a -> b) -> Lifted a -> Lifted b
+apply fn x = [(fnVal xVal, conj [fnPC,xPC]) 
+              | (fnVal,fnPC) <- fn, (xVal,xPC) <- x, sat(conj[fnPC,aPC])] 
+\end{minted}
+
+Here we are using Haskell's list comprehension to build the lifted result of applying a lifted function to a single lifted argument.
+The function \textbf{conj} takes a list of presence conditions are returns their conjunction, and the function \textbf{sat} takes a presence condition (a symbolic proposition in general) and returns true if its argument is satisfiable, otherwise it returns false.
+
+\begin{exmp}[Unary function application]
+
+Assume that we need to be apply the absolute value function (\emph{abs}) to the lifted variable \emph{x} from the previous example.
+Since \emph{x} is lifted, we need a lifted version of \emph{abs}, which we can simply get by using the liftT function:
+
+\begin{minted}{haskell}
+liftedAbs = liftT abs
+\end{minted}
+
+Now we can apply \emph{liftedAbs} to \emph{x}:
+
+\begin{minted}{haskell}
+abs_x = apply liftedAbs x
+\end{minted}
+
+The \emph{apply} function applies a lifted function to a lifted value (of matching type), resulting in a lifted value. In this example, the result \emph{abs\_x} will have the value: \\
+
+[(1,A),(2, $\neg A \wedge B$), (3, $\neg A\wedge \neg B$)].
+
+\end{exmp}
+
+\begin{exmp}[Binary function application]
+
+Now assume that we would like to add \emph{x} and \emph{y}. The first step is to get a lifted version of the (+) operator:
+
+\begin{minted}{haskell}
+plus = (+)
+liftedPlus = liftT plus
+\end{minted}
+
+The function \emph{apply} takes a function and a single argument, so how do we extend it beyond unary functions?
+
+Actually this is no different from functions in TLC, where all functions are unary, and functions of higher arities are just Curried higher-order functions \cite{Pierce2002}.
+If we pass \emph{liftedPlus} and \emph{x} to apply, it will apply the binary lifted operator \emph{liftedPlus} to a single lifted argument \emph{x}:
+
+\begin{minted}{haskell}
+intermediateVal = apply liftedPlus x
+\end{minted}
+
+But since multi-parameter functions are actually Curried unary functions, \emph{intermediateVal} is a lifted unary function [((plus 1), A),((plus -2), $\neg A \wedge B$),((plus 3), $\neg A \wedge \neg B$)].
+
+Now we just need to apply \emph{intermediateVal}, which is actually a lifted unary function, to \emph{y}:
+
+\begin{minted}{haskell}
+finalVal = apply intermediateVal y
+\end{minted}
+
+The result in \emph{finalVal} is: \\
+
+[(6, $A \wedge \neg B$), (5, $A \wedge B$), (2, $\neg A \wedge B$), (0, $\neg A \wedge \neg B$)] \\
+
+Remember that apply generates the cross product of its arguments (each of them a list of values labeled with presence conditions), and excludes the pairs where the conjunction of presence conditions is not satisfiable.
+This is why finalVal has only four possible values, although the cross product initially had 9.
+
+As a shortcut, we can define a binary version of \emph{apply} (let us call it \emph{apply2}) that implements this pattern of partial function application and Currying:
+
+\begin{minted}{haskell}
+apply2 :: Lifted (a -> b -> c) -> Lifted a -> Lifted b -> Lifted c
+apply2 fn a b = apply (apply fn a) b
+\end{minted}
+
+Versions of \emph{apply} for functions of other aritiies  can be defined similarly:
+
+\begin{minted}{haskell}
+apply3 :: Lifted (a -> b -> c -> d) -> Lifted a -> Lifted b -> Lifted c -> Lifted d
+apply3 fn a b c = apply (apply2 fn a b) c
+
+apply4 :: Lifted (a -> b -> c -> d -> e) -> 
+			Lifted a -> Lifted b -> Lifted c -> Lifted d -> Lifted e
+apply4 fn a b c d = apply (apply3 fn a b c) d
+
+...
+\end{minted}
+
+\end{exmp}
+
+\begin{exmp}[Conditional Expression]
+
+An \emph{if-then-else} expression is just syntactic sugar for a ternary polymorphic function, taking a boolean argument and two arguments of type \emph{t} and returning one of them depending on what the first argument evaluates to.
+We can easily de-sugar \emph{if-then-else} expressions into the following \emph{cond} function:
+
+\begin{minted}{haskell}
+cond :: Bool -> t -> t -> t
+cond p a b = if p then a else b
+\end{minted}
+
+Now lifting a conditional expression is no different from lifting any other function:
+
+\begin{minted}{haskell}
+condLifted = apply3 (liftT cond)
+\end{minted}
+
+So for example the expression
+
+\begin{minted}{haskell}
+if (x == y) then (x + y) else (x - y)
+\end{minted}
+
+Is lifted into:
+
+\begin{minted}{haskell}
+liftedEq    a b = apply2 (liftT (==)) a b
+liftedPlus  a b = apply2 (liftT (+))  a b
+liftedMinus a b = apply2 (liftT (-))  a b
+
+condLifted (liftedEq x y) (liftedPlus x y) (liftedMinus x y)
+\end{minted}
+
+Please not that Haskell's lazy evaluation makes the de-sugared \emph{cond} function semantically equivalent to a conditional expression because only one of the \emph{then} and \emph{else} expressions is evaluated after the boolean condition is evaluated based on its truth value.
+However, a language with strict evaluation semantics would evaluate both the \emph{then} and \emph{else} expressions before they are passed to \emph{cond}.
+This is still fine if both expressions are pure (with no side-effects).
+However, if any of the expression has side effects, the equivalence between \emph{cond} and the typical conditional expression is broken.
+
+\end{exmp}
+
+\subsubsection{The Disjointness Invariant and Function Application}
+
+Function application is the only way of generating new lifted values out of existing ones.
+To preserve the soundness of our lifting system, we need to make sure that function application preserves Invariant 1.
+
+\newtheorem{th1}{Theorem}
+\begin{th1}
+If fn and x satisfy Invariant 1, then the result of (apply fn x) also satisfies the invariant.
+\end{th1}
+
+\begin{proof}
+
+By contradiction:
+
+\begin{itemize}
+\item Assume that both fn and x satisfy the invariant while the result y = (apply fn x) does not
+\item Then there exists at least one pair of presence conditions in y ($pc_i$ and $pc_j$) where sat($pc_i \wedge pc_j$) = true
+\item Since presence conditions $pc_i$ and $pc_j$ where constructed by apply as conjunctions of presence conditions from fn and x, and we know presence conditions of fn are disjoint (same for those of x), then there is no truth assignment that would satisfy the conjunction of any of the pairs of presence conditions in y.
+\end{itemize}
+
+\end{proof}
+
+\subsubsection{Lifting Composite Types}
+
+\subsubsection{Function Rewriting}
+
+\subsection{Lifting Parametric Types}
+
+\section{Examples and Evaluation}
+
+\subsection{LTS Reachability with Witness}
+
+\subsection{TODO: Some other analysis}
+
+\section{Discussion}
+\subsection{Deep vs Shallow Lifting}
+
+\section{Related Work}
+
+\section{Conclusion}
+
+\bibliography{/users/ramy/bib/library} 
+\bibliographystyle{ieeetr}
 

### 2016-11-25 Variability.hs

@@ -3,9 +3,6 @@
 -- Software Product Line Variability library
 -- Ramy Shahin - July 14th 2016
 -------------------------------------------------------------------------------
-{-# LANGUAGE PolyKinds #-}
---{-# LANGUAGE #-}
-
 module Variability where
 
 -- abstract propositional expression
@@ -23,7 +20,7 @@ disj a b = True     -- TODO
 sat :: Expr -> Bool
 sat ex = True
 
-type SPLOption a = (a, PresenceCondition)
+type SPLValue a = (a, PresenceCondition)
 
 --instance Eq SPLOption a where
 --    (==) a b = (getValue a == getValue b) && sat(getPresenceCondition a && getPresenceCondition b)
@@ -32,13 +29,21 @@ type SPLOption a = (a, PresenceCondition)
 -- when lifting a product value to a product line value, we might end up with
 -- different values for each product in the product line. This is why a value is
 -- lifted into a set of values, each with a path condition. An important
--- invariant here is that the path conditions of those values should not depend
+-- inSPLVariable here is that the path conditions of those values should not depend
 -- on each other, i.e. non of them logically implies the other. Violating this
--- invariant would result in redundant values (i.e. multiple values belonging
+-- inSPLVariable would result in redundant values (i.e. multiple values belonging
 -- to the same set of products). This does not affect correctness, but severely
 -- affects performance as we are now degenerating into brute force analysis
 -- across all possible products.
-type Lifted a = [SPLOption a]
+newtype SPLVariable a = V [SPLValue a]
+
+instance Show a => Show (SPLVariable a) where
+    show v = case v of V v' -> show v'
+
+instance Functor SPLVariable where
+    fmap f (V v) = V [(f x, pc) | (x, pc) <- v]
+
+
 --type Lifted2 a b = Lifted (a (Lifted b))
[…]

### 2016-11-25 Variability_test.hs

@@ -1,0 +1,14 @@
+-- Variability unit tests
+import Variability
+import Test.HUnit
+
+x :: SPLVariable Int
+x = V [(3, True), (-6, True), (-2, True)]
+
+y :: SPLVariable Int
+y = V [(-7, True), (2, True), (-1, True)]
+
+xAbs = apply (liftT abs) x
+yAbs = apply (liftT abs) y
+
+x_plus_y = apply2 (liftT (+)) x y

### 2017-01-03 doc/paper.tex
N/A

--- 2016-08-11 doc/paper.tex
@@
 % These packages are all incorporated in the memoir class to one degree or another...
+
+\usepackage{minted}
+\usepackage{stmaryrd}
 
@@
 
+\usepackage{bussproofs}
+
 %%% ToC (table of contents) APPEARANCE
@@
 \renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!
+
+%%\usepackage{natbib}
 
@@
 
-\section{First section}
+\theoremstyle{definition}
+\newtheorem{exmp}{Example}[section]
 
-Your text goes here.
+\newcommand{\tab}{\qquad}
 
-\subsection{A subsection}
+% mathematical object definition
+\newcommand{\term}[1]		{\mathrm{#1}}
+\newcommand{\var}[1]		{\mathrm{#1}}
+\newcommand{\type}[1]		{\mathrm{#1}}
+\newcommand{\kw}[1]	{\mathrm{#1}}
+\newcommand{\ite}[3]		{\kw{if} \: #1 \: \kw{then} \: #2 \: \kw{else} \: #3}
 
-More text.
+\section{Introduction}
+
+%Software Product Lines (SPLs) are configurable sets of software artifacts from which different software products can be generated.
+SPLs, pretty much like classic manufacturing product lines, are much more economic in terms of time and effort, compared to developing and maintaining different software product variations separately.
+This is the reason why SPLs are becoming commonly adopted in many software domains, including operating systems, embedded systems, compilers and many others.
+
+%SPLs capture product variability in several ways, including annotative mechanisms available in conventional programming languages.
+One of the most commonly used annotative mechanisms is C preprocessor macro definitions and conditional compilation.
+Product features are abstracted in terms of preprocessor macros, and feature selection is done using build-time macro definitions.
+This approach, although ad-hoc compared to more structural approaches provided by SPL language (e.g.
+Featured-oriented languages), has the benefit of not requiring any tools beyond the conventional C toolchain.
+
+\section{Background}
+
+\subsection{Propositional Logic}
+TODO (syntax, semantics, sat and unsat)
+
+\subsection{Typed Lambda Calculus}
+
+We assume that analyses are written in a Typed Lambda Calculus (TLC) based language. This is a summary of the TLC syntax and semantics from\cite{Pierce2002}.
+
+\subsubsection{Syntax}
+
+\begin{center}
+\begin{tabular}{ l r}
+$\term{t} ::=$                                           	&	\tab (terms)		\\
+\tab 	$\var{x}$						&	\tab variable		\\
+\tab	$\lambda \var{v}:\type{T} .\term{t}$	&	\tab abstraction		\\	
+\tab	$\term{t} \:  \term{t}$ 				&	\tab application		\\
+\\
+$\var{v} ::=$						&	\tab (values)		\\
+\tab	$\lambda \var{v}:\type{T} .\term{t}$	&	\tab abstraction	value\\
+\\
+$T ::=$							&	\tab (types)		\\
+\tab $...$							&	\tab primitive types	\\
+\tab	$T \to T$						&	\tab type of functions \\
+\\
+$\Gamma ::=$						&	\tab (contexts)		\\
+\tab 	$\Phi$						&	\tab empty context	\\
+\tab	$\Gamma, \var{x}:\type{T}$		&	\tab term variable binding	\\
+\\	
+\end{tabular}
+\end{center}
+
+\subsubsection{Semantics}
+
+\begin{prooftree}
+\AxiomC{$\term{t}_1 \to \term{t}_1'$}
+\RightLabel{(E-App1)}
+\UnaryInfC{$\term{t}_1 \: \term{t}_2 \to \term{t}_1' \:  \term{t}_2$}	
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\term{t}_2 \to \term{t}_2'$}
+\RightLabel{(E-App2)}
+\UnaryInfC{$\var{v}_1 \: \term{t}_2 \to \var{v}_1 \:  \term{t}_2'$}	
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{E-AppAbs}
+\UnaryInfC{$(\lambda \var{x} . \term{t}) \var{v}_2 \to [\var{x} \mapsto \var{v}_2] \term{t}$}
+\end{prooftree}
+
+\subsubsection{Typing}
+
+\begin{prooftree}
+\AxiomC{$\var{x}:\type{T} \in \Gamma$}
+\RightLabel{T-Var}
+\UnaryInfC{$\Gamma \vdash \var{x}:\type{T}$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\Gamma, \var{x}:\type{T_1} \vdash \term{t_2}:\type{T_2}$}
+\RightLabel{T-Abs}
+\UnaryInfC{$\Gamma \vdash \lambda \var{x}:\type{T_1}.\term{t_2} : \type{T_1} \to \type{T_2}$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\Gamma \vdash \term{t_1}:\type{T_1} \to \type{T_2}$}
+\AxiomC{$\Gamma \vdash \term{t_2}:\type{T_1}$}
+\RightLabel{T-App}
+\BinaryInfC{$\Gamma \vdash \term{t_1} \: \term{t_2} : \type{T_2}$}
+\end{prooftree}
+
+\subsection{Software Product Lines}
+
+These formal definitions are adapted from the ones in\cite{Salay2014}:
+
+A product line $P$ it a tuple $<F,\Phi,D,\phi>$ where:
+\begin{itemize}
+	\item $F$ is a set of features
+	\item $\Phi$ (the feature model) is a propositional formula over $F$
+	\item $D$ (the domain model) is a set of domain elements
+	\item $\phi : D \to PC$ is a total function mapping domain elements to presence conditions %TODO
+\end{itemize}
+
+%\subsubsection{Product Variability}
+
+%\subsubsection{Features}
+
+%\subsubsection{Presence Conditions}
+%Software Product Lines (SPLs) are built on top of both common and variable software artifacts.
+Common artifacts exist in all products of the product line, while variable artifacts exist only in a non-empty subset of valid products.
+That subset is defined using a \emph{Presence Condition}: a propositional formula over product line features.
+
+%\subsubsection{Product Generation}
+
+\subsection{Software Analyses}
+We assume that a Software Analysis is a TLC program $G$ that analyzes a single instance of the product line $P$.
+
+%\section{Lifting Typed Lambda Calculus}
+
+%By lifting Typed Lambda Calculus (TLC), we mean providing syntactic transformation rules that lift any program written in TLC by compositionally lifting TLC syntactic constructs.
+Those transformation rules can be applied to any software analysis written in TLC, resulting in an equivalent analysis that can be applied to a Software Product Line.
+
+%In the following description of the transformation rules we use Haskell syntax.
+This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as the construction of those constructs on top of TLC is standard \cite{Pierce2002}.
+
+%When lifting an analysis from products to product lines, we need to appropriately tag variable artifacts with their corresponding presence conditions.
+We use the \emph{PresenceCondition} abstract data type for this purpose.
+
+%\subsection{Lifting Types}
+%Given a type \textbf{t}, we lift it into type \textbf{Lifted t} which is:
+
+%\begin{minted}{haskell}
+%type Lifted t = [(t, PresenceCondition)]
+%\end{minted}
+
+%where:
+%\begin{itemize}
+%\item \textbf{t} is a type parameter that can be instantiated with any type
+%\item \textbf{(t, PresenceCondition)} is a pair of a value of type \textbf{t} together with a \textbf{PresenceCondition}
+%\item \textbf{[(t, PresenceCondition)]} is a list of such pairs
+%\item \textbf{Lifted t $\equiv$ [(t, PresenceCondition)]} is a parametric data type, instantiated by passing the type argument \textbf{t} to the type constructor \textbf{Lifted}
+%\end{itemize}
+
+%A variable of type \textbf{t} in a product line might actually have different values in different products, hence the multiplicity of items in a list.
+The \textbf{PresenceCondition} tags are used to keep track of which values correspond to which products.
+For this to be semantically correct we need to make sure that the sets of products represented by those presence conditions do not overlap, otherwise we might end up with a product variable having different values at the same time.
+This \textbf{disjointness} requirement is formalized in the following invariant:
+
+%\newtheorem*{disjInv*}{Disjointness Invariant}
+%\begin{disjInv*}
+%Given a lifted value $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i \neq j: {unsat (pc_i \wedge pc_j)}$
+%\end{disjInv*}
+
+%The invariant simply states that a lifted value can not have different values with presence condition that overlap.
+The \emph{unsat} function takes a proposition and returns true if and only if that proposition is unsatisfiable.
+The conjunction of two presence conditions is unsatisfiable if and only if they represent non-overlapping sets of products.
+
+%This invariant is a precondition on inputs into the program being lifted, and in the following sections we will show how its validity is maintained by the lifting transformations.
+
+%\subsection{Lifting Values}
+
+%To lift a value of type \textbf{t}, all we need to is to provide a presence condition for it, and then encapsulate both into a value of type \textbf{Lifted t}.
+This is what the following \textbf{lift} operator does:
+
+%\begin{minted}{haskell}
+%lift :: PresenceCondition -> t -> Lifted t
+%lift pc x = [(x, pc)]
+%\end{minted}
+
+%This operator can be applied to values of any type, including functions. The importance of lifting function values will be clear in the following subsection.
+
+%In many cases we need to provide the most general presence condition, which is \textbf{True} (the set of all products). For this we provide the following \textbf{liftT} operator:
+
+%\begin{minted}{haskell}
+%liftT :: t -> Lifted t
+%liftT x = lift True x
+%\end{minted}
+
+%Input values to analyses from product lines will typically be of lifted types. For example, an integer can have one value in a set of products, and a different value in another set.
+
+%\begin{exmp}[Lifted Values]
+
+%Assuming we have two features A and B, variables x and y of type \textbf{Lifted Int} and variable z of type \textbf{Lifted String} can have the following values: \\
+
+%x = [(1,A),(-2, $\neg A \wedge B$),(3, $\neg A \wedge \neg B$)]
+
+%y = [(5, $A \wedge \neg B$), (4,B), (-3, $\neg A \wedge \neg B$)]
+
+%z = [("hello", $\neg A \wedge B$),("helloooo", A)] \\
+
+%Both x and y satisfy the disjointness invariant.
+It is worth mentioning that a lifted value does not have to be defined in each product in the product line.
+For example, z is not defined in the product identified by the presence condition $(\neg A \wedge \neg B)$.
+
+%\end{exmp}
+
+%\subsection{Applying Lifted Functions}
+
+%Now that we have lifted values (both data and functions), we need to provide a way to apply lifted functions to lifted operands.
+This is how data propagates through a functional program.
+Starting with the simplest case, where we have a lifted unary function of type \textbf{Lifted (a $\to$ b)} and an argument of type \textbf{Lifted a} we need to provide a way to apply that function to its argument resulting in a \textbf{Lifted b} value.
+
+%Recall that a lifted value of type \textbf{Lifted t} is just a collection of values of type \textbf{t}, each with a corresponding presence condition, and the whole lifted value has to satisfy the disjointness Invariant.
+So to apply a collection of functions to a collection of values, all we need to is to apply each of the functions to each of the values.
+Since the result has to be a lifted value, then we also need a presence condition for each of the computed values.
+This presence condition is simply the conjunction of the presence conditions of the function and the argument used to compute it.
+If the conjunction of the presence conditions is unsatisfiable (i.e.
+a contradiction), it means that the corresponding value does not belong to any product, so it can be safely filtered out.
+This is how the apply function looks in Haskell:
+
+%\begin{exmp}[Unary function application]
+
+%Assume that we need to be apply the absolute value function (\emph{abs}) to the lifted variable \emph{x} from the previous example.
+Since \emph{x} is lifted, we need a lifted version of \emph{abs}, which we can simply get by using the liftT function:
+
+%\begin{minted}{haskell}
+%liftedAbs = liftT abs
+%\end{minted}
+
+%Now we can apply \emph{liftedAbs} to \emph{x}:
+
+%\begin{minted}{haskell}
+%abs_x = apply liftedAbs x
+%\end{minted}
+
+%The \emph{apply} function applies a lifted function to a lifted value (of matching type), resulting in a lifted value. In this example, the result \emph{abs\_x} will have the value: \\
+
+%[(1,A),(2, $\neg A \wedge B$), (3, $\neg A\wedge \neg B$)].
+
+%\end{exmp}
+
+%\begin{exmp}[Binary function application]
+
+%Now assume that we would like to add \emph{x} and \emph{y}. The first step is to get a lifted version of the (+) operator:
+
+%\begin{minted}{haskell}
+%plus = (+)
+%liftedPlus = liftT plus
+%\end{minted}
+
+%The function \emph{apply} takes a function and a single argument, so how do we extend it beyond unary functions?
+
+%Actually this is no different from functions in TLC, where all functions are unary, and functions of higher arities are just Curried higher-order functions \cite{Pierce2002}.
+If we pass \emph{liftedPlus} and \emph{x} to apply, it will apply the binary lifted operator \emph{liftedPlus} to a single lifted argument \emph{x}:
+
+%\begin{minted}{haskell}
+%intermediateVal = apply liftedPlus x
+%\end{minted}
+
+%But since multi-parameter functions are actually Curried unary functions, \emph{intermediateVal} is a lifted unary function [((plus 1), A),((plus -2), $\neg A \wedge B$),((plus 3), $\neg A \wedge \neg B$)].
+
+%Now we just need to apply \emph{intermediateVal}, which is actually a lifted unary function, to \emph{y}:
+
+%\begin{minted}{haskell}
+%finalVal = apply intermediateVal y
+%\end{minted}
+
+%The result in \emph{finalVal} is: \\
+
+%[(6, $A \wedge \neg B$), (5, $A \wedge B$), (2, $\neg A \wedge B$), (0, $\neg A \wedge \neg B$)] \\
+
+%Remember that apply generates the cross product of its arguments (each of them a list of values labeled with presence conditions), and excludes the pairs where the conjunction of presence conditions is not satisfiable.
+This is why finalVal has only four possible values, although the cross product initially had 9.
+
+%As a shortcut, we can define a binary version of \emph{apply} (let us call it \emph{apply2}) that implements this pattern of partial function application and Currying:
+
+%\begin{minted}{haskell}
+%apply2 :: Lifted (a -> b -> c) -> Lifted a -> Lifted b -> Lifted c
+%apply2 fn a b = apply (apply fn a) b
+%\end{minted}
+
+%Versions of \emph{apply} for functions of other aritiies  can be defined similarly:
+
+%\begin{minted}{haskell}
+%apply3 :: Lifted (a -> b -> c -> d) -> Lifted a -> Lifted b -> Lifted c -> Lifted d
+%apply3 fn a b c = apply (apply2 fn a b) c
+
+%apply4 :: Lifted (a -> b -> c -> d -> e) -> 
+%			Lifted a -> Lifted b -> Lifted c -> Lifted d -> Lifted e
+%apply4 fn a b c d = apply (apply3 fn a b c) d
+
+%...
+%\end{minted}
+
+%\end{exmp}
+
+%\begin{exmp}[Conditional Expression]
+
+%An \emph{if-then-else} expression is just syntactic sugar for a ternary polymorphic function, taking a boolean argument and two arguments of type \emph{t} and returning one of them depending on what the first argument evaluates to.
+We can easily de-sugar \emph{if-then-else} expressions into the following \emph{cond} function:
+
+%\begin{minted}{haskell}
+%cond :: Bool -> t -> t -> t
+%cond p a b = if p then a else b
+%\end{minted}
+
+%Now lifting a conditional expression is no different from lifting any other function:
+
+%\begin{minted}{haskell}
+%condLifted = apply3 (liftT cond)
+%\end{minted}
+
+%So for example the expression
+
+%\begin{minted}{haskell}
+%if (x == y) then (x + y) else (x - y)
+%\end{minted}
+
+%Is lifted into:
+
+%\begin{minted}{haskell}
+%liftedEq    a b = apply2 (liftT (==)) a b
+%liftedPlus  a b = apply2 (liftT (+))  a b
+%liftedMinus a b = apply2 (liftT (-))  a b
+
+%condLifted (liftedEq x y) (liftedPlus x y) (liftedMinus x y)
+%\end{minted}
+
+%Please not that Haskell's lazy evaluation makes the de-sugared \emph{cond} function semantically equivalent to a conditional expression because only one of the \emph{then} and \emph{else} expressions is evaluated after the boolean condition is evaluated based on its truth value.
+However, a language with strict evaluation semantics would evaluate both the \emph{then} and \emph{else} expressions before they are passed to \emph{cond}.
+This is still fine if both expressions are pure (with no side-effects).
+However, if any of the expression has side effects, the equivalence between \emph{cond} and the typical conditional expression is broken.
+
+%\end{exmp}
+
+%\subsubsection{Lifting Composite Types}
+
+%\subsubsection{Function Rewriting}
+
+%\subsection{Lifting Parametric Types}
+
+\section{Formal Definitions and Theorems}
+
+\subsection{Definitions}
+\subsubsection{Product Line Values}
+Given a type $t \in T$,  a value $\var{v}$ of type $\type{t}$, and the presence condition of $\var{v}$ is $\var{pc}$, then the pair $(\var{v}, \term{pc})$ is of the polymorphic type $\type{SPLVal} \: \type{t}$.
+
+\begin{prooftree}
+\AxiomC{$\type{t} \in T$}
+\AxiomC{$\var{v}:\type{t}$}
+\AxiomC{$\phi (\var{v}) = \term{pc}$}
+\RightLabel{(T-SPLVal)}
+\TrinaryInfC{$(\var{v}, \term{pc}) : \type{(SPLVal \: \type{t})}$}	
+\end{prooftree}
+
+\begin{minted}{haskell}
+type SPLVal t = (t, PresenceCondition)
+\end{minted}
+
+\subsubsection{Product Line Variables}
+
+Given a type $t \in T$, the set $\{(x_1, pc_1),...,(x_n, pc_n)\}$ is of type $(\type{SPLVar} \: \type{t})$, where:
+\begin{itemize}
+\item $\forall i, (x_i, pc_i) : (\type{SPLVal} \: \type{t})$ (Correctness)
+\item $\forall (i \neq j), {unsat(\var{pc}_i \wedge \var{pc}_j)}$ (Disjointness)
+\end{itemize}
+
+\begin{prooftree}
+\AxiomC{$\type{t} \in T$}
+\AxiomC{$(\var{x_1}, \var{pc_1}), ... , (\var{x_n}, \var{pc_n}) : (\type{SPLVal} \: \type{t})$}
+\AxiomC{$\forall (i \neq j), {unsat(\var{pc}_i \wedge \var{pc}_j)}$}
+\RightLabel{(T-SPLVar)}
+\TrinaryInfC{$\{(x_1, pc_1),...,(x_n, pc_n)\} : (\type{SPLVar} \: \type{t})$}	
+\end{prooftree}
+
+\begin{minted}{haskell}
+type SPLVar t = [SPLVal t]
+\end{minted}
+
+%\newtheorem*{prodValid*}{Product Validity}
+%\begin{prodValid*}
+%Given an SPLVar $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i: {pc_i \Rightarrow \Phi}$
+%\end{prodValid*}
+
+%\newtheorem*{correctness*}{Correctness}
+%\begin{correctness*}
+%Given an SPLVar $v'$ that corresponds to a single-product variable $v$ in $G$, $(x,pc) \in v' \Leftrightarrow \forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket v \rrbracket = x}$
+%\end{correctness*}
+
+%\newtheorem*{disjInv*}{Disjointness}
+%\begin{disjInv*}
+%Given an SPLVar $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i \neq j: {unsat (pc_i \wedge pc_j)}$
+%\end{disjInv*}
+
+\subsubsection{Function Definition}
+Given a function $\var{f} : \type{a_1} \to ... \to \type{a_n} \to \type{b}$ where $\exists i, \type{a_i} \in T$, \\
+we define $\var{f'} : \type{SPLVar} \: (\type{a_1} \to ... \to \type{a_n} \to \type{b})$, where $\var{f'} = \{(\var{f}, True)\}$
+
+\begin{prooftree}
+\AxiomC{$\var{f} : \type{a_1} \to ... \to \type{a_n} \to \type{b}$}
+\AxiomC{$\exists i, \type{a_i} \in T$}
+\RightLabel{(T-SPLVar-f)}
+\BinaryInfC{$\var{f'} : \type{SPLVar} \: (\type{a_1} \to ... \to \type{a_n} \to \type{b}) = \{(\var{f}, True)\}$}	
+\end{prooftree}
+
+\begin{minted}{haskell}
+f' :: SPLVar (a -> b)
+f' = [(f, True)]
+\end{minted}
+
+%Here the newly defined SPL function variable $f'$ is a singleton wrapper of the original function $f$, and making it applicable to all products.
+The SPLVar $f'$ obviously satisfies the disjointness condition, and because it is a fresh variable that does not exist in $G$ the correctness condition does not apply to it.
+
+\subsubsection{Function Application}
+
+Given a function $\var{f'} : \type{SPLVar} \: (\type{a} \to \type{b})$ 
+and a variable $\var{x} : (\type{SPLVar} \: \type{a})$, 
+the result of applying $\var{f'}$ to $\var{x}$ is the set 
+$\{(\var{f''}(\var{x'}), \var{fpc} \wedge \var{xpc}) \: | \: (\var{f''}, \var{fpc}) \in \var{f'}, (\var{x'}, \var{xpc}) \in x, sat(\var{fpc} \wedge \var{xpc})\}$
+
+\begin{prooftree}
+\AxiomC{$\var{f'} : \type{SPLVar} \: (\type{a} \to \type{b})$}
+\AxiomC{$\var{x} : (\type{SPLVar} \: \type{a})$}
+\RightLabel{(E-SPL-apply)}
+\BinaryInfC{$(\var{apply} \: \var{f'} \: \var{x}) = \{(\var{f''}(\var{x'}), \var{fpc} \wedge \var{xpc}) \: | \: (\var{f''}, \var{fpc}) \in \var{f'}, (\var{x'}, \var{xpc}) \in x, sat(\var{fpc} \wedge \var{xpc})\}$}
+\end{prooftree}
+
+\begin{minted}{haskell}
+apply :: SPLVar (a -> b) -> SPLVar a -> SPLVar b
+apply fn x = [(fnVal xVal, conj [fnPC,xPC]) 
+              | (fnVal,fnPC) <- fn, (xVal,xPC) <- x, sat(conj[fnPC,aPC])] 
+\end{minted}
+
+%Here we are using Haskell's list comprehension syntax to build the (SPLVar b) result of applying a lifted function to a single argument of type (SPLVar a).
+The function \textbf{conj} takes a list of presence conditions and returns their conjunction, and the function \textbf{sat} takes a presence condition (a symbolic proposition in general) and returns true if its argument is satisfiable, otherwise it returns false.
+
+As in almost all functional programming languages, Currying is the standard mechanism to extend application of single-argument functions to multi-argument functions.
+We provide versions of apply for different arities:
+
+\begin{minted}{haskell}
+apply2 :: SPLVar (a -> b -> c) -> SPLVar a -> SPLVar b -> SPLVar c
+apply2 fn a = apply (apply fn a)
+
+apply3 :: SPLVar (a -> b -> c -> d) -> SPLVar a -> SPLVari b -> SPLVar c -> SPLVar d
+apply3 fn a b = apply (apply2 fn a b)
+\end{minted}
+
+\subsection{Theorems}
+
+\subsubsection{Application Preserves Disjointness}
+
+\newtheorem{th1}{Theorem}
+\begin{th1}
+If f and x satisfy the disjointness condition, then the result of (apply f x) also satisfies disjointness.
+\end{th1}
+
+\begin{proof}
+
+By contradiction:
+
+\begin{itemize}
+\item Assume that both f and x satisfy disjointness while the result y = (apply f x) does not
+\item Then there exists at least one pair of presence conditions in y ($pc_i$ and $pc_j$) where sat($pc_i \wedge pc_j$) = true
+\item Each of $pc_i$ and $pc_j$ is a conjunction of presence conditions from f and x (definition of apply), then for arbitrary a, b, c and d: \\
+               $pc_i = fpc_a \wedge xpc_b$ \\
+               $pc_j = fpc_c \wedge xpc_d$
+\item Then $pc_i \wedge pc_j = pc_a \wedge xpc_b \wedge fpc_c \wedge xpc_d$
+\item For this conjunction to be satisfiable, we must have both sat($fpc_a \wedge xpc_b$) and sat($fpc_c \wedge xpc_d$), which contradicts the assumption that f and x both satisfy disjointness
+\item Then y = (apply f x) must satisfy disjointness too
+\end{itemize}
+
+\end{proof}
+
+\subsubsection{Application Preserves Correctness}
+
+\newtheorem{th2}{Theorem}
+\begin{th2}
+If f' and x' satisfy the correctness condition with respect to original variables f and x respectively, then the result of (apply f' x') also satisfies correctness with respect to (f x).
+\end{th2}
+
+\begin{proof}
+Given v' = (apply f' x'), we need to prove that $(y, pc) \in v' \Leftrightarrow \forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket f(x) \rrbracket = y}$
+\begin{itemize}
+\item{$\Rightarrow$}
+	\begin{itemize}
+	\item Assume that $(y,pc) \in v'$ and for some arbitrary truth assignment $e$, $e \Rightarrow pc$
+	\item Then by definition of apply, there exists $(f'', fpc) \in f'$, $(x'', xpc) \in x$, such that y = f''(x'') and pc = $fpc \wedge xpc$
+	\item Since $e \Rightarrow pc$, then $e \Rightarrow fpc \wedge xpc$, which means that $e \Rightarrow fpc$ and $e \Rightarrow xpc$
+	\item By definition of $G(e)$ and the assumption that both f' and x' satisfy the correctness condition, $G(e) \llbracket f \rrbracket = f'$, and $G(e) \llbracket x \rrbracket = x'$
+	\item Then $G(e) \llbracket f(x) \rrbracket = y$
+	\end{itemize}
+\item{$\Leftarrow$}
+	\begin{itemize}
+	\item Assume $\forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket f(x) \rrbracket = y}$
+	\item Then given an arbitrary truth assignment $e$ (identifying a single product) where $e \Rightarrow pc$, $G(e) \llbracket f(x) \rrbracket = y$
+	\item Since f' and x' are both assumed to satisfy the disjointness and correctness conditions, then $(f,fpc) \in f'$ and $(x,xpc) \in x'$ where $pc = fpc \wedge xpc$
+	\item Then by definition of apply, $(y,pc) \in$ (apply f' x')
+	\end{itemize}
+\end{itemize}
+
+\end{proof}
+
+\section{Examples and Evaluation}
+
+\subsection{LTS Reachability with Witness}
+
+\subsection{TODO: Some other analysis}
+
+\section{Discussion}
+\subsection{Deep vs Shallow Lifting}
+
+\section{Related Work}
+
+\section{Conclusion}
+
+\bibliography{paper} 
+\bibliographystyle{ieeetr}
 

### 2017-01-03 Variability.hs

@@ -5,6 +5,8 @@
 -------------------------------------------------------------------------------
 module Variability where
 
+import Control.Monad
+
 -- abstract propositional expression
 type Expr = Bool
 
@@ -99,9 +101,10 @@ apply4 fn a b c = apply (apply3 fn a b c)
 --applyFn :: (a -> SPLVariable b) -> SPLVariable a -> SPLVariable b
 --applyFn fn (V x) = (map (\(x', pc) -> ((fn x'), pc)) x)
 
---instance Monad SPLVariable where
---    return x = liftT x
---    x >>= y  = apply y x
+instance Monad SPLVariable where
+    return x = liftT x
+    y >>= f  = case y of 
+                V y' -> V (concat (map (\(y'',pc) -> (f y'', pc)) y'))
 
 -- lifting conditional expression
 cond :: Bool -> a -> a -> a

### 2017-01-05 Prop_test.hs
SPL type families (incomplete)

@@ -1,0 +1,25 @@
+-- Propositional Logic and SAT solving
+-- Ramy Shahin
+-- Jan 3rd 2017
+module Prop where
+
+import Z3.Monad
+import qualified Data.Traversable as T
+
+script :: Z3 Result
+script = do
+    p <- mkFreshBoolVar "p"
+    q <- mkFreshBoolVar "q"
+    r <- mkFreshBoolVar "r"
+    s <- mkFreshBoolVar "s"
+
+    assert =<< mkAnd =<< T.sequence [(mkNot q), (return q)]
+    assert =<< mkOr =<< T.sequence [(mkNot p), (mkNot s)]
+
+    check
+    
+run :: IO ()
+run = do
+    result <- evalZ3 script
+    print result
+

### 2017-01-05 SPL.hs

@@ -1,9 +1,11 @@
 -------------------------------------------------------------------------------
--- Variability.hs
--- Software Product Line Variability library
+-- SPL.hs
+-- Software Product Line library
 -- Ramy Shahin - July 14th 2016
 -------------------------------------------------------------------------------
-module Variability where
+{-# LANGUAGE TypeFamilies #-}
+
+module SPL where
 
 import Control.Monad
 
@@ -22,7 +24,7 @@ disj a b = True     -- TODO
 sat :: Expr -> Bool
 sat ex = True
 
-type SPLValue a = (a, PresenceCondition)
+type Val a = (a, PresenceCondition)
 
 --instance Eq SPLOption a where
 --    (==) a b = (getValue a == getValue b) && sat(getPresenceCondition a && getPresenceCondition b)
@@ -31,19 +33,26 @@ type SPLValue a = (a, PresenceCondition)
 -- when lifting a product value to a product line value, we might end up with
 -- different values for each product in the product line. This is why a value is
 -- lifted into a set of values, each with a path condition. An important
--- inSPLVariable here is that the path conditions of those values should not depend
+-- inVar here is that the path conditions of those values should not depend
 -- on each other, i.e. non of them logically implies the other. Violating this
--- inSPLVariable would result in redundant values (i.e. multiple values belonging
+-- inVar would result in redundant values (i.e. multiple values belonging
 -- to the same set of products). This does not affect correctness, but severely
 -- affects performance as we are now degenerating into brute force analysis
 -- across all possible products.
-newtype SPLVariable a = V [SPLValue a]
+--newtype Var (a :: *) = V [Val a]
+--type Var1 (a :: * -> *) = [Val (a (Var b))]
 
-instance Show a => Show (SPLVariable a) where
-    show v = case v of V v' -> show v'
+--type Var a = [Val a]
 
-instance Functor SPLVariable where
[…]

### 2017-01-05 SPL_test.hs

@@ -1,0 +1,16 @@
+-- SPL unit tests
+import SPL
+import Test.HUnit
+
+x :: Var Int
+x = [(3, True), (-6, True), (-2, True)]
+
+y :: Var Int
+y = [(-7, True), (2, True), (-1, True)]
+
+xAbs = apply (liftT abs) x
+yAbs = apply (liftT abs) y
+
+x_plus_y = apply2 (liftT (+)) x y
+
+

### 2017-01-05 Variability_test.hs

@@ -1,14 +1,0 @@
--- Variability unit tests
-import Variability
-import Test.HUnit
-
-x :: SPLVariable Int
-x = V [(3, True), (-6, True), (-2, True)]
-
-y :: SPLVariable Int
-y = V [(-7, True), (2, True), (-1, True)]
-
-xAbs = apply (liftT abs) x
-yAbs = apply (liftT abs) y
-
-x_plus_y = apply2 (liftT (+)) x y

### 2017-01-06 Prop.hs
SAT/Z3 module with tests

@@ -4,22 +4,56 @@
 module Prop where
 
 import Z3.Monad
-import qualified Data.Traversable as T
+import Data.Vector(Vector, (!), fromList)
 
-script :: Z3 Result
-script = do
-    p <- mkFreshBoolVar "p"
-    q <- mkFreshBoolVar "q"
-    r <- mkFreshBoolVar "r"
-    s <- mkFreshBoolVar "s"
+type Universe = Vector String
 
-    assert =<< mkAnd =<< T.sequence [(mkNot q), (return q)]
-    assert =<< mkOr =<< T.sequence [(mkNot p), (mkNot s)]
+mkUniverse :: [String] -> Universe
+mkUniverse = fromList
 
+data Prop =
+    T
+  | F
+  | Atom Universe Int
+  | Not Prop
+  | Conj [Prop]
+  | Disj [Prop]
+
+showPropList :: [Prop] -> String
+showPropList ps = 
+    case ps of
+        []  -> ""
+        [p] -> (show p) 
+        p1 : p2 : ps' -> (show p1) ++ ", " ++ showPropList (p2 : ps')
+
+instance Show Prop where
+    show prop = 
+        case prop of
+            T        -> "True"
+            F        -> "False"
+            Atom u i ->  u ! i
+            Not p    -> "Not(" ++ show p ++ ")"
+            Conj ps  -> "And(" ++ showPropList ps ++ ")"
+            Disj ps  -> "Or("  ++ showPropList ps ++ ")"
+
+mkZ3Formula :: Vector AST -> Prop -> Z3 AST
+mkZ3Formula atoms p =
+    case p of
+        T        -> mkTrue
+        F        -> mkFalse
+        Atom _ i -> return (atoms ! i)
+        Not p'   -> mkNot =<< (mkZ3Formula atoms p')
+        Conj ps  -> mkAnd =<< (mapM (mkZ3Formula atoms) ps)
+        Disj ps  -> mkOr  =<< (mapM (mkZ3Formula atoms) ps)
+
+mkZ3Script :: Universe -> Prop -> Z3 Result
+mkZ3Script u p = do
+    atoms <- mapM mkFreshBoolVar u
+    assert =<< (mkZ3Formula atoms p)
     check
     
-run :: IO ()
-run = do
-    result <- evalZ3 script
-    print result
+checkSAT :: Universe -> Prop -> IO Result
+checkSAT u p = evalZ3 (mkZ3Script u p)
+
+
 

### 2017-01-06 Prop_test.hs

@@ -1,25 +1,31 @@
--- Propositional Logic and SAT solving
+-- Propositional Logic and SAT solving Tests
 -- Ramy Shahin
 -- Jan 3rd 2017
-module Prop where
+import Prop
 
-import Z3.Monad
-import qualified Data.Traversable as T
+import GHC.Arr
 
-script :: Z3 Result
-script = do
-    p <- mkFreshBoolVar "p"
-    q <- mkFreshBoolVar "q"
-    r <- mkFreshBoolVar "r"
-    s <- mkFreshBoolVar "s"
+u :: Universe
+u = mkUniverse ["P", "Q", "R", "S"]
 
-    assert =<< mkAnd =<< T.sequence [(mkNot q), (return q)]
-    assert =<< mkOr =<< T.sequence [(mkNot p), (mkNot s)]
+p, q, r, s :: Prop
+p = Atom u 0
+q = Atom u 1
+r = Atom u 2
+s = Atom u 3
 
-    check
-    
+prop1 = Conj [p, (Not q)]   -- sat
+prop2 = Disj [(Not r), s]   -- sat
+prop3 = Conj [p, (Not p)]   -- unsat
+prop4 = Disj [r]            -- sat
+prop5 = Conj [F, p, q]      -- unsat
+prop6 = Disj [T, prop1, prop3] -- sat
+
+props = [prop1, prop2, prop3, prop4, prop5, prop6]
+
 run :: IO ()
-run = do
-    result <- evalZ3 script
-    print result
+run = mapM_ (\p -> do
+                    r <- (checkSAT u p)
+                    putStrLn ((show p) ++ "------>" ++ (show r))
+            ) props
 

### 2017-01-06 Prop.hs
Adding implication and double-implication

@@ -18,6 +18,8 @@ data Prop =
   | Not Prop
   | Conj [Prop]
   | Disj [Prop]
+  | Impl Prop Prop
+  | Iff  Prop Prop
 
 showPropList :: [Prop] -> String
 showPropList ps = 
@@ -29,22 +31,32 @@ showPropList ps =
 instance Show Prop where
     show prop = 
         case prop of
-            T        -> "True"
-            F        -> "False"
-            Atom u i ->  u ! i
-            Not p    -> "Not(" ++ show p ++ ")"
-            Conj ps  -> "And(" ++ showPropList ps ++ ")"
-            Disj ps  -> "Or("  ++ showPropList ps ++ ")"
+            T           -> "True"
+            F           -> "False"
+            Atom u i    ->  u ! i
+            Not p       -> "Not(" ++ show p ++ ")"
+            Conj ps     -> "And(" ++ showPropList ps ++ ")"
+            Disj ps     -> "Or("  ++ showPropList ps ++ ")"
+            Impl p1 p2  -> "(" ++ show p1 ++ " => " ++ show p2 ++ ")"
+            Iff  p1 p2  -> "(" ++ show p1 ++ " <=> " ++ show p2 ++ ")"
 
 mkZ3Formula :: Vector AST -> Prop -> Z3 AST
 mkZ3Formula atoms p =
     case p of
-        T        -> mkTrue
-        F        -> mkFalse
-        Atom _ i -> return (atoms ! i)
-        Not p'   -> mkNot =<< (mkZ3Formula atoms p')
-        Conj ps  -> mkAnd =<< (mapM (mkZ3Formula atoms) ps)
-        Disj ps  -> mkOr  =<< (mapM (mkZ3Formula atoms) ps)
+        T           -> mkTrue
+        F           -> mkFalse
+        Atom _ i    -> return (atoms ! i)
+        Not p'      -> mkNot =<< (mkZ3Formula atoms p')
+        Conj ps     -> mkAnd =<< (mapM (mkZ3Formula atoms) ps)
+        Disj ps     -> mkOr  =<< (mapM (mkZ3Formula atoms) ps)
+        Impl p1 p2  -> do
+                        p1' <- mkZ3Formula atoms p1
+                        p2' <- mkZ3Formula atoms p2
+                        mkImplies p1' p2'
+        Iff p1 p2   -> do
[…]

### 2017-01-06 Prop_test.hs

@@ -20,8 +20,12 @@ prop3 = Conj [p, (Not p)]   -- unsat
 prop4 = Disj [r]            -- sat
 prop5 = Conj [F, p, q]      -- unsat
 prop6 = Disj [T, prop1, prop3] -- sat
+prop7 = Impl p (Not p)      -- sat
+prop8 = Iff  p (Not p)      -- unsat
+prop9 = Impl p q            -- sat
+prop10 = Iff p q            -- sat
 
-props = [prop1, prop2, prop3, prop4, prop5, prop6]
+props = [prop1, prop2, prop3, prop4, prop5, prop6, prop7, prop8, prop9, prop10]
 
 run :: IO ()
 run = mapM_ (\p -> do

### 2017-01-06 Prop.hs
Removed the universe parameter from checkSAT, sat and unsat. Instead, the universe is retrieved from within the formula.

@@ -4,7 +4,7 @@
 module Prop where
 
 import Z3.Monad
-import Data.Vector(Vector, (!), fromList)
+import Data.Vector as V(Vector, (!), fromList, empty, length)
 
 type Universe = Vector String
 
@@ -40,6 +40,25 @@ instance Show Prop where
             Impl p1 p2  -> "(" ++ show p1 ++ " => " ++ show p2 ++ ")"
             Iff  p1 p2  -> "(" ++ show p1 ++ " <=> " ++ show p2 ++ ")"
 
+maxUniverse :: Universe -> Universe -> Universe
+maxUniverse u1 u2 =
+    let l1 = V.length u1
+        l2 = V.length u2
+    in
+        if (l1 >= l2) then u1 else u2
+
+getUniverse :: Prop -> Universe
+getUniverse p = 
+    case p of
+        T -> empty
+        F -> empty
+        Atom u i -> u
+        Not p' -> getUniverse p'
+        Conj ps -> foldr (maxUniverse . getUniverse) empty ps
+        Disj ps -> foldr (maxUniverse . getUniverse) empty ps
+        Impl p1 p2 -> maxUniverse (getUniverse p1) (getUniverse p2)
+        Iff  p1 p2 -> maxUniverse (getUniverse p1) (getUniverse p2)
+
 mkZ3Formula :: Vector AST -> Prop -> Z3 AST
 mkZ3Formula atoms p =
     case p of
@@ -64,11 +83,11 @@ mkZ3Script u p = do
     assert =<< (mkZ3Formula atoms p)
     check
     
-checkSAT :: Universe -> Prop -> IO Result
-checkSAT u p = evalZ3 (mkZ3Script u p)
+checkSAT :: Prop -> IO Result
+checkSAT p = evalZ3 (mkZ3Script (getUniverse p) p)
 
-sat :: Universe -> Prop -> IO Bool
-sat u p = (return . ((==) Sat)) =<< checkSAT u p
+sat :: Prop -> IO Bool
+sat p = (return . ((==) Sat)) =<< checkSAT p
 
-unsat :: Universe -> Prop -> IO Bool
-unsat u p = (return . ((==) Unsat)) =<< checkSAT u p
+unsat :: Prop -> IO Bool
+unsat p = (return . ((==) Unsat)) =<< checkSAT p

### 2017-01-06 Prop_test.hs

@@ -29,7 +29,7 @@ props = [prop1, prop2, prop3, prop4, prop5, prop6, prop7, prop8, prop9, prop10]
 
 run :: IO ()
 run = mapM_ (\p -> do
-                    r <- (checkSAT u p)
+                    r <- (checkSAT p)
                     putStrLn ((show p) ++ "------>" ++ (show r))
             ) props
 

### 2017-03-15 doc/evaluation.tex
various late changes

\documentclass[12pt]{article}
\begin{document}
\section{Evaluation}
\subsection{Benchmarks}
\subsubsection{Parsing C Code}

\begin{itemize}
	\item Case studies from SPLVerifier (contributed to SVComp)
	\begin{itemize}
		\item email client
		\item elevator
		\item minepump
	\end{itemize}

	\item \emph{Stretch goal:} x86 Linux kernel files (about 7600 files) used in the evaluation of Typechef and SuperC
\end{itemize}

\subsubsection{C Code Typechecking}
On the same case studies used in C code parsing.
We need a C type checker written in a functional language though to lift.

\subsubsection{Simple LTS Algorithms}
\begin{itemize}
	\item State reachability
	\item State reachability with witness
\end{itemize}

\subsubsection{Model Transformations (graph rules)}
ICSE'14 paper

\subsection{Evaluation Metrics}
\subsubsection{Correctness}
Especially for the parsing case study, we can compare against Typechef and SuperC.

\subsubsection{Performance}
With respect to number of features and number of valid products.

Performance is evaluated for the baseline (black-box shallow lifting) and any optimizations that might be implemented. Possible optimizations include:
\begin{itemize}
	\item Using Binary Decision Diagrams (BDDs) instead of SAT solvers
	\item Deep-lifting of the List data structure for space optimization and data redundancy elimination
	\item Deep-lifting of other data structures
	\item Depp-lifting of algorithms (as needed, based on profiling results) 
\end{itemize}

\end{document}

### 2017-03-15 TypeChecker/src/TypeChecker.hs

@@ -1,0 +1,9 @@
+-- TypeChecker.hs
+-- C type checker lifted to product lines
+-- Ramy Shahin
+-- Feb. 22nd 2017
+module TypeChecker where
+
+import SPL
+import Language.C.Analysis.TypeCheck
+

### 2017-03-15 vcpp/src/macro.hs

@@ -1,0 +1,29 @@
+-- macro.hs
+--  a module for processing C preprocessor macros
+--  Ramy Shahin
+--  March 2nd 2017
+module Macro where
+import SPL
+import Data.Map
+
+data Macro = Macro {
+    name        :: String,
+    params      :: [String],
+    bodyTokens  :: [String]
+    expansion   :: [String] -> String
+} deriving (Show)
+
+type Macro' = Var Macro
+
+type MacroTable = String (Map Macro')
+
+addMacro :: PresenceCondition -> MacroTable -> Macro -> MacroTable
+-- TODO: #define
+
+removeMacro :: PresenceCondition -> MacroTable -> String -> MacroTable
+-- TODO: #undef
+
+
+
+
+

### 2017-03-15 vcpp/src/old_lexer.hs

@@ -1,0 +1,35 @@
+-- lexer.hs
+--  A C language lexer
+-- Ramy Shahin
+-- March 2nd 2017
+
+module Lexer where 
+
+import Text.ParserCombinators.Parsec 
+import Text.ParserCombinators.Parsec.Expr
+import Text.ParserCombinators.Parsec.Language 
+import qualified Text.ParserCombinators.Parsec.Token as Token
+
+cLangDef = emptyDef {
+    Token.commentStart      = "/*",
+    Token.commentEnd        = "*/",
+    Token.commentLine       = "//",
+    Token.nestedComments    = True,
+    Token.identStart        = letter <|> char '_',
+    Token.identLetter       = alphaNum <|> char '_',
+    Token.reservedNames     = ["auto", "break", "case", "char", "const", 
+                               "continue", "default", "do", "double", "else",
+                               "enum", "extern", "float", "for", "goto", "if",
+                               "int", "long", "register", "return", "short",
+                               "signed", {-"sizeof",-} "static", "struct", 
+                               "switch", "typedef", "union", "unsigned", 
+                               "void", "volatile", "while"],
+    Token.reservedOpNames   = ["++", "--", "sizeof", "&", "*", "+", "-", "~",
+                               "!", "/", "%", "<<", ">>", "<", "<=", ">", ">=",
+                               "==", "!=", "^", "|", "&&", "||", "?", ":", "=",
+                               "*=", "/=", "%=", "+=", "-=", "<<=", ">>=", 
+                               "&=", "^=", "|=", ","],
+    Token.caseSensitive     = True
+}
+
+cLexer = Token.makeTokenParser cLangDef

### 2017-03-15 vcpp/test/include1.c

@@ -1,0 +1,7 @@
+#include "a.h"
+int x=0;
+int y=1;
+
+void main()
+{
+}

### 2017-03-18 SPL/src/SPL.hs
lifted cond bug fix

@@ -70,13 +70,16 @@ compact (Var v) =
                         in  (head vs, disj pcs)) 
             gs)
 
-valIndex :: Var t -> PresenceCondition -> [Val t]
-valIndex (Var v) pc =
-    filter (\(_,pc') -> sat (conj[pc, pc'])) v
+valIndex :: Eq t => Var t -> t -> [Val t]
+valIndex (Var v) x =
+    filter (\(x',pc') -> case x' of 
+                            Just x'' -> x'' == x
+                            _        -> False) 
+                        v
 
 subst :: Show t => Var t -> PresenceCondition -> Var t
 subst (Var v) pc =
-    Var (filter (\(v',pc') -> sat (conj [pc,pc'])) v)
+    Var (filter (\(_,pc') -> sat (conj [pc,pc'])) v)
 
 union :: Var t -> Var t -> Var t
 union (Var a) (Var b) =
@@ -91,7 +94,6 @@ inv (Var v) = {-trace ("inv: " ++ (show (Var v))) $-}
     all (\((_, pc1),(_, pc2)) -> unsat (conj[pc1, pc2])) (pairs v)
 
 apply :: Var (a -> b) -> Var a -> Var b
-
 apply (Var fn) (Var x) =
     Var [(case (fn', x') of
         (Just fn'', Just x'') -> Just (fn'' x'')
@@ -101,17 +103,6 @@ apply (Var fn) (Var x) =
                     (x', xpc) <- x,
                     let pc = conj[fnpc, xpc],
                     sat pc]
-{-
-apply (Var fn) (Var x) = 
-    let triples = [(fn', x', c) |   (fn', fnpc) <- fn,
-                                    (x', xpc) <- x,
-                                    let c= conj[fnpc, xpc],
-                                    sat c                   ]
-    in  Var (map (\(fn', x', c) ->  (case (fn', x') of
-                                        (Just fn'', Just x'') -> fn'' `seq` Just (fn'' x'')
-                                        (_,_) -> Nothing
-    , c)) triples)
-  -}      
 
 instance Show a => Show (Var a) where
     show (Var v) = "{\n" ++ (foldr (++) "" (map (\x -> (show x) ++ "\n") v)) ++ "}" 
@@ -129,25 +120,6 @@ cond :: Bool -> a -> a -> a
 cond p a b = if p then a else b
 
[…]

### 2017-03-18 SPL/src/VList.hs

@@ -1,0 +1,15 @@
+module VList where
+
+import SPL
+
+data List' a =
+    VNull
+  | VCons (Var a) (Var (List' a))
+  deriving(Show)
+
+type VList a = Var (List' a)
+
+vlength :: VList a -> Var Int
+vlength VNull = (mkVarT 0)
+vlength (VCons h tl) = (mkVarT 1) |+| (vlength tl)
+

### 2017-03-18 SPL/test/SPL_test.hs

@@ -1,7 +1,7 @@
 -- SPL unit tests
 import SPL
 import Prop
-import Test.HUnit
+--import Test.HUnit
 import Control.Applicative
 
 u :: Universe
@@ -36,7 +36,7 @@ abs' = pure abs
 xAbs = abs' <*> x
 yAbs = abs' <*> y
 
-(|+|) = cliftV2 (+)
+(|+|) = liftV2 (+)
 
 x_plus_y0 = apply (apply (pure (+)) x) y
 x_plus_y1 = (pure (+)) <*> x <*> y
@@ -44,13 +44,13 @@ x_plus_y2 = x |+| y
 
 safeDiv :: Int -> Int -> Int
 safeDiv a b = if b == 0 then 0 else (div a b)
-div' = cliftV2 div
+div' = liftV2 div
 
 zero = (mkVarT 0)
 one = (mkVarT 1)
  
 safeDiv' :: Var Int -> Var Int -> Var Int
-safeDiv' = cliftV2 safeDiv 
+safeDiv' a b = cond' (b |==| zero) zero (div' a b)
 
 a :: Var Int
 a = mkVarT 0
@@ -62,4 +62,4 @@ xs = mkVarList [w,x,y,z]
 xs' = mkVarList' (mkVarT [w, x, y, z])
 
 length' :: Var [a] -> Var Int
-length' = cliftV length
+length' = liftV length

### 2017-03-18 SPL/test/VList_test.hs

@@ -1,0 +1,36 @@
+import VList
+import SPL
+import Prop
+
+u :: Universe
+u = mkUniverse ["P", "Q", "R", "S"]
+
+p, q, r, s :: Prop
+p = Atom u 0
+q = Atom u 1
+r = Atom u 2
+s = Atom u 3
+
+pq = conj[p,q]
+p_q = conj[p, neg q]
+_pq = conj[neg p, q]
+_p_q = conj[neg p, neg q]
+_p = neg p
+
+w :: Var Int
+w = mkVars [(12, pq), (2, p_q), (3, _p_q)]
+
+x :: Var Int
+x = mkVars [(7, pq), (-3, p_q), (-8, _pq), (0, _p_q)]
+
+y :: Var Int
+y = mkVars [(-11, _p)]
+
+z :: Var Int
+z = mkVars [(6, p)]
+
+l0 = VNull
+l1 = VCons w l0
+l2 = VCons x l1
+l3 = VCons y l2
+l4 = VCons z l3

### 2017-04-08 doc/evaluation.tex
paper definitions section

@@
 \section{Evaluation}
-\subsection{Benchmarks}
-\subsubsection{Parsing C Code}
 
+\subsection{Evaluation Metrics}
+\subsubsection{Correctness}
+Correctness validation would involve enumerating outputs for all products and comparing them with the product-by-product output.
+Practically we can only do this for relatively small case studies with a few features.
+Simple FTS examples might be enough.
+
+The rest of the paper should provide correctness proofs, so the proofs together with small validation examples should adequately cover correctness.
+
+\subsubsection{Performance}
+With respect to number of features and number of valid products.
+
+We should demonstrate black-box, white-box and gray-box lifting performance.
+
+Performance is evaluated for the baseline (black-box shallow lifting) and any optimizations that might be implemented. Possible optimizations might include:
 \begin{itemize}
-	\item Case studies from SPLVerifier (contributed to SVComp)
-	\begin{itemize}
-		\item email client
-		\item elevator
-		\item minepump
+	\item Using the feature model to eliminate infeasible feature combinations early on (more computational load for the SAT solver though)
+	\item Using Binary Decision Diagrams (BDDs) instead of SAT solvers
+	\item Deep-lifting of the List data structure for space optimization and data redundancy elimination
+	\item Deep-lifting of other data structures
+	\item Deep-lifting of algorithms (as needed, based on profiling results)
 \end{itemize}
 
-	\item \emph{Stretch goal:} x86 Linux kernel files (about 7600 files) used in the evaluation of Typechef and SuperC
-\end{itemize}
-
-\subsubsection{C Code Typechecking}
-On the same case studies used in C code parsing.
-We need a C type checker written in a functional language though to lift.
+\subsection{Benchmarks}
 
@@
 
-\subsubsection{Model Transformations (graph rules)}
-ICSE'14 paper
+\subsubsection{Parsing C Code}
+\begin{itemize}
+	\item GNU CoreUtils
+%	\item Case studies from SPLVerifier (contributed to SVComp)
+%	\begin{itemize}
+%		\item email client
+%		\item elevator
+%		\item minepump
+%	\end{itemize}
 
-\subsection{Evaluation Metrics}
-\subsubsection{Correctness}
-Especially for the parsing case study, we can compare against Typechef and SuperC.
-
-\subsubsection{Performance}
-With respect to number of features and number of valid products.
-
-Performance is evaluated for the baseline (black-box shallow lifting) and any optimizations that might be implemented. Possible optimizations include:
-\begin{itemize}
-	\item Using Binary Decision Diagrams (BDDs) instead of SAT solvers
-	\item Deep-lifting of the List data structure for space optimization and data redundancy elimination
-	\item Deep-lifting of other data structures
-	\item Depp-lifting of algorithms (as needed, based on profiling results) 
+%	\item \emph{Stretch goal:} x86 Linux kernel files (about 7600 files) used in the evaluation of Typechef and SuperC
 \end{itemize}
 
+\subsubsection{C Code Analyses}
+On the same case studies used in C code parsing.
+\begin{itemize}
+	\item Typechecking: We need a C type checker written in a functional language though to lift.
+	\item A simple workflow of LLVM analyses (gray-box)
+\end{itemize}
+
+%\subsubsection{Model Transformations (graph rules)}
+%ICSE'14 paper
+
 \end{document}

### 2017-04-08 SPL/src/SPL.hs

@@ -218,17 +218,17 @@ foldr' f' e' xs' =
 -}
 
 -- Lifted List (VList)
-type VList a = Var [a]
+--type VList a = Var [a]
 
-vnull :: VList a -> Var Bool
-vnull = liftV null
+--vnull :: VList a -> Var Bool
+--vnull = liftV null
 
-vhead = liftV head
+--vhead = liftV head
 
-vtail = liftV tail
+--vtail = liftV tail
 
-mkVList :: (Show t) => [Var t] -> Var [t]
-mkVList = foldr (|:|) e
+--mkVList :: (Show t) => [Var t] -> Var [t]
+--mkVList = foldr (|:|) e
 
 --mkVarList' :: (Show t) => Var [t] -> Var [t]
 --mkVarList' = foldr' (pure (:)) e

### 2017-04-08 FTS/examples/Wiper.hs
some FTS examples

@@ -1,0 +1,56 @@
+-- Wiper.hs
+-- Ramy Shahin
+-- March 19th 2017
+module Wiper where
+
+import LTS
+import LiftedLTS
+import Prop
+import SPL
+import Test.HUnit (Test(..), assertEqual, runTestTT)
+import Control.Applicative
+
+-- Features
+f :: Universe
+f = mkUniverse ["sL", "sH", "wL", "wH"]
+
+sL, sH, wL, wH :: Prop
+sL  = Atom f 0
+sH  = Atom f 1
+wL  = Atom f 2
+wH  = Atom f 3
+
+states :: Var [State]
+states = mkVarStates 0 2
+
+non, noRain, little, heavy, rain, heavyRain :: Action'
+non         = mkAction (mkVarT "non") e
+noRain      = mkAction (mkVarT "noRain") e
+little      = mkAction (mkVarT "little") e
+heavy       = mkAction (mkVarT "heavy") e
+rain        = mkAction (mkVarT "rain") e
+heavyRain   = mkAction (mkVarT "heavyRain") e
+
+
+actions :: Var [Action]
+actions =  mkVarList [non, noRain, little, heavy, rain, heavyRain]
+
+start2locking, locking2waiting :: Transition'
+
+--                                     from         to          actions
+s0_s1_euro          = mkTransition     (states!!0)  (states!!1)  e -- T
+locking2waiting     = mkTransition     locking      waiting  (mkVarList [heaterOn]) -- (disj[heat, delay])
+locking2washing     = mkTransition     locking      washing  (mkVarList [setDelay, washStart])
+waiting2washing     = mkTransition     waiting      washing  (mkVarList [heaterOff, washStart]) -- (disj[heat, delay])
+washing2unlocking   = mkTransition     washing      unlocking (mkVarList [quickCool]) -- (neg dry)
+washing2drying      = mkTransition     washing      drying    e -- dry
+drying2unlocking    = mkTransition     drying       unlocking (mkVarList [quickCool]) -- dry
+unlocking2finish    = mkTransition     unlocking    finish    e -- T
+
+transitions :: Var [Transition]
+
+transitions =  mkVarList [start2locking, locking2waiting, locking2washing, waiting2washing, washing2unlocking, washing2drying, drying2unlocking, unlocking2finish]
+
+washingMachine :: LTS'
+washingMachine = mkLTS  states  actions  transitions  (mkVarList [start]) --T
+

### 2017-04-08 VendingMachine.hs

@@ -1,0 +1,66 @@
+-- VendingMachine.hs
+-- Ramy Shahin
+-- March 17th 2017
+module VendingMachine where
+
+import LTS
+import LiftedLTS
+import Prop
+import SPL
+import Test.HUnit (Test(..), assertEqual, runTestTT)
+import Control.Applicative
+
+-- Features
+f :: Universe
+f = mkUniverse ["coffee", "tea", "cappuccino", "ringTone", "euro", "dollar"]
+
+coffee, tea, cappuccino, ringTone, euro, dollar :: Prop
+coffee      = Atom f 0
+tea         = Atom f 1
+cappuccino  = Atom f 2
+ringTone    = Atom f 3
+euro        = Atom f 4
+dollar      = Atom f 5
+
+states :: Var [State]
+states = mkVarStates 0 13
+
+oneEuro, oneDollar, sugar, noSugar, coffee_, tea_, cappuccino_, pourSugar, pourCoffee, pourTea, pourMilk, displayDone, skipRingTone, ringTone, cupTaken :: Action'
+oneEuro     = mkAction (mkVarT "oneEuro") e
+oneDollar   = mkAction (mkVarT "oneDollar") e
+sugar       = mkAction (mkVarT "sugar") e
+noSugar     = mkAction (mkVarT "noSugar") e
+coffee_     = mkAction (mkVarT "coffee_") e
+tea_        = mkAction (mkVarT "tea_") e
+cappuccino_ = mkAction (mkVarT "cappuccino_") e
+pourSugar   = mkAction (mkVarT "pourSugar") e
+pourCoffee  = mkAction (mkVarT "pourCoffee") e
+pourTea     = mkAction (mkVarT "pourTea") e
+pourMilk    = mkAction (mkVarT "pourMilk") e
+displayDone = mkAction (mkVarT "displayDone") e
+skipRingTone = mkAction (mkVarT "skipRingTone") e
+ringTone    = mkAction (mkVarT "ringTone") e
+cupTaken    = mkAction (mkVarT "cupTaken") e
+
+actions :: Var [Action]
+actions =  mkVarList [oneEuro, oneDollar, sugar, noSugar, coffee_, tea_, cappuccino_, pourSugar, pourCoffee, pourTea, pourMilk, displayDone, skipRingTone, ringTone, cupTaken]
+
+start2locking, locking2waiting :: Transition'
+
+--                                     from         to          actions
+s0_s1_euro          = mkTransition     (states!!0)  (states!!1)  e -- T
[…]

### 2017-04-08 doc/paper.tex

--- 2016-11-25 doc/paper.tex
@@
-% !TEX TS-program = pdflatex
-% !TEX encoding = UTF-8 Unicode
+%% bare_conf.tex
+%% V1.4b
+%% 2015/08/26
+%% by Michael Shell
+%% See:
+%% http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
+%% conference paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/pkg/ieeetran
+%% and
+%% http://www.ieee.org/
 
-% This is a simple template for a LaTeX document using the "article" class.
-% See "book", "report", "letter" for other types of document.
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE! 
+%% User assumes all risk.
+%% In no event shall the IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%*************************************************************************
 
-\documentclass[11pt]{article} % use larger type; default would be 10pt
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. The IEEE's font choices and paper sizes can   ***
+% *** trigger bugs that do not appear when using other class files.       ***                          ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
 
-\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
+\documentclass[10pt,conference]{IEEEtran}
+% Some Computer Society conferences also require the compsoc mode option,
+% but others use the standard conference format.
+%
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[conference]{../sty/IEEEtran}
 
-%%% Examples of Article customizations
-% These packages are optional, depending whether you want the features they provide.
-% See the LaTeX Companion or other references for full information.
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
 
-%%% PAGE DIMENSIONS
-\usepackage{geometry} % to change the page dimensions
-\geometry{a4paper} % or letterpaper (US) or a5paper or....
-% \geometry{margin=2in} % for example, change the margins to 2 inches all round
-% \geometry{landscape} % set up the page for landscape
-%   read geometry.pdf for detailed page layout information
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/pkg/ifpdf
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
 
-\usepackage{graphicx} % support the \includegraphics command and options
+% *** CITATION PACKAGES ***
+%
+%\usepackage{cite}
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of the IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off
+% such as if a citation ever needs to be enclosed in parenthesis.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 5.0 (2009-03-20) and later if using hyperref.sty.
+% The latest version can be obtained at:
+% http://www.ctan.org/pkg/cite
+% The documentation is contained in the cite.sty file itself.
+
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation
+% can be obtained at: 
+% http://www.ctan.org/pkg/graphicx
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found at:
+% http://www.ctan.org/pkg/epslatex
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
+
+% *** MATH PACKAGES ***
+%
+%\usepackage{amsmath}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics.
+%
+% Note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/amsmath
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/pkg/algorithms
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/pkg/algorithmicx
+
+% *** ALIGNMENT PACKAGES ***
+%
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/array
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+% *** SUBFIGURE PACKAGES ***
+%\ifCLASSOPTIONcompsoc
+%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+%\else
+%  \usepackage[caption=false,font=footnotesize]{subfig}
+%\fi
+% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
+% for subfigure.sty, the latter of which is no longer maintained and is
+% incompatible with some LaTeX packages including fixltx2e. However,
+% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
+% which will override IEEEtran.cls' handling of captions and this will result
+% in non-IEEE style figure/table captions. To prevent this problem, be sure
+% and invoke subfig.sty's "caption=false" package option (available since
+% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
+% handling of captions.
+% Note that the Computer Society format requires a larger sans serif font
+% than the serif footnote size font used in traditional IEEE formatting
+% and thus the need to invoke different subfig.sty package options depending
+% on whether compsoc mode has been enabled.
+%
+% The latest version and documentation of subfig.sty can be obtained at:
+% http://www.ctan.org/pkg/subfig
+
+% *** FLOAT PACKAGES ***
+%
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure.
+% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
+% corrections already built into the system in which case a warning will
+% be issued if an attempt is made to load fixltx2e.sty as it is no longer
+% needed.
+% The latest version and documentation can be found at:
+% http://www.ctan.org/pkg/fixltx2e
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/stfloats
+% Do not use the stfloats baselinefloat ability as the IEEE does not allow
+% \baselineskip to stretch. Authors submitting work to the IEEE should note
+% that the IEEE rarely uses double column equations and that authors should try
+% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
+% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
+% such ways.
+% Do not attempt to use stfloats with fixltx2e as they are incompatible.
+% Instead, use Morten Hogholm'a dblfloatfix which combines the features
+% of both fixltx2e and stfloats:
+%
+% \usepackage{dblfloatfix}
+% The latest version can be found at:
+% http://www.ctan.org/pkg/dblfloatfix
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+%\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/url
+% Basically, \url{my_url_here}.
+
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+\usepackage{amsmath}
 \usepackage{amsthm}
+\usepackage{bussproofs}
+\usepackage{minted}
+\usepackage{stmaryrd}
 
-% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
 
-%%% PACKAGES
-\usepackage{booktabs} % for much better looking tables
-\usepackage{array} % for better arrays (eg matrices) in maths
-\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
-\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
-\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
-% These packages are all incorporated in the memoir class to one degree or another...
-
-\usepackage{minted}
-
-%%% HEADERS & FOOTERS
-\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
-\pagestyle{fancy} % options: empty , plain , fancy
-\renewcommand{\headrulewidth}{0pt} % customise the layout...
-\lhead{}\chead{}\rhead{}
-\lfoot{}\cfoot{\thepage}\rfoot{}
-
-%%% SECTION TITLE APPEARANCE
-\usepackage{sectsty}
-\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
-% (This matches ConTeXt defaults)
-
-%%% ToC (table of contents) APPEARANCE
-\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
-\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
-\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
-\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!
-
-%%\usepackage{natbib}
-
-%%% END Article customizations
-
-%%% The "real" document content comes below...
-
-\title{Automatically Lifting Software Analyses to Product Lines}
-\author{Ramy Shahin}
-%\date{} % Activate to display a given date or no date (if empty),
-         % otherwise the current date is printed
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
 
 \begin{document}
+%
+% paper title
+% Titles are generally capitalized except for words such as a, an, and, as,
+% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
+% not capitalized unless they are the first or last word of the title.
+% Linebreaks \\ can be used within to get better formatting as desired.
+% Do not put math or special symbols in the title.
+\title{Automatically Lifting Functional Software Analyses to Product Lines}
+
+% author names and affiliations
+% use a multiple column layout for up to three different
+% affiliations
+\author{\IEEEauthorblockN{Ramy Shahin}
+\IEEEauthorblockA{Computer Science Department\\
+University of Toronto\\
+Email: ramy.shahin@mail.utoronto.ca}
+\and
+\IEEEauthorblockN{Marsha Chechik}
+\IEEEauthorblockA{Computer Science Department\\
+University of Toronto\\
+Email: chechik@cs.toronto.edu}}
+
+% conference papers do not typically use \thanks and this command
+% is locked out in conference mode. If really needed, such as for
+% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
+% after \documentclass
+
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
+% 
+%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
+%Homer Simpson\IEEEauthorrefmark{2},
+%James Kirk\IEEEauthorrefmark{3}, 
+%Montgomery Scott\IEEEauthorrefmark{3} and
+%Eldon Tyrell\IEEEauthorrefmark{4}}
+%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
+%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
+
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% make the title area
 \maketitle
+
+% As a general rule, do not put math, special symbols or citations
+% in the abstract
+\begin{abstract}
+TODO
+\end{abstract}
+
+% no keywords
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
 
@@
 \newtheorem{exmp}{Example}[section]
+
+\newcommand{\tab}{\qquad}
+
+% mathematical object definition
+\newcommand{\term}[1]		{\mathrm{#1}}
+\newcommand{\var}[1]		{\mathrm{#1}}
+\newcommand{\type}[1]		{\mathrm{#1}}
+\newcommand{\kw}[1]		{\mathrm{#1}}
+\newcommand{\ite}[3]		{\kw{if} \: #1 \: \kw{then} \: #2 \: \kw{else} \: #3}
 
@@
 
-\subsubsection{Product Variability}
+Given a set of features $F$:
+\begin{itemize}
+	\item $P[F]$ is the set of propositional formulas over $F$
+	\item $sat:P[F] \to boolean$, \[ sat(p) = \begin{cases}
+			true\ \text{if p is propositionally satisfiable} \\
+			false\ \text{otherwise}
+		\end{cases} \]
+	\item $unsat:P[F] \to boolean$, \[ unsat(p) = \begin{cases}
+			true\ \text{if p is propositionally unsatisfiable} \\
+			false\ \text{otherwise}
+		\end{cases} \]
 
-\subsubsection{Features}
+\end{itemize}
 
-\subsubsection{Presence Conditions}
-Software Product Lines (SPLs) are built on top of both common and variable software artifacts.
-Common artifacts exist in all products of the product line, while variable artifacts exist only in a non-empty subset of valid products.
-That subset is defined using a \emph{Presence Condition}: a propositional formula over product line features.
-
-\subsubsection{Product Generation}
-
-\subsection{Software Analyses}
+A product line $P$ it a tuple $<F,\Phi,D,\phi>$ where:
+\begin{itemize}
+	\item $F$ is a set of features
+	\item $\Phi \in P[F]$ (the feature model) is a propositional formula over $F$
+	\item $D$ (the domain model) is a set of domain elements
+	\item $\phi : D \to P[F]$ is a total function mapping each domain elements to a propositional formula over $F$, known as a Presence Condition(PC)
+\end{itemize}
 
@@
 
+This is a summary of the TLC syntax and semantics from\cite{Pierce2002}.
+
+\subsubsection{Syntax}
+
+\begin{center}
+\begin{tabular}{ l r}
+$\term{t} ::=$                                           	&	\tab (terms)		\\
+\tab 	$\var{x}$						&	\tab variable		\\
+\tab	$\lambda \var{v}:\type{T} .\term{t}$	&	\tab abstraction		\\	
+\tab	$\term{t} \:  \term{t}$ 				&	\tab application		\\
+\\
+$\var{v} ::=$						&	\tab (values)		\\
+\tab	$\lambda \var{v}:\type{T} .\term{t}$	&	\tab abstraction	value\\
+\\
+$T ::=$							&	\tab (types)		\\
+\tab $...$							&	\tab primitive types	\\
+\tab	$T \to T$						&	\tab type of functions \\
+\\
+$\Gamma ::=$						&	\tab (contexts)		\\
+\tab 	$\Phi$						&	\tab empty context	\\
+\tab	$\Gamma, \var{x}:\type{T}$		&	\tab term variable binding	\\
+\\	
+\end{tabular}
+\end{center}
+
+\subsubsection{Semantics}
+
+\begin{prooftree}
+\AxiomC{$\term{t}_1 \to \term{t}_1'$}
+\RightLabel{(E-App1)}
+\UnaryInfC{$\term{t}_1 \: \term{t}_2 \to \term{t}_1' \:  \term{t}_2$}	
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\term{t}_2 \to \term{t}_2'$}
+\RightLabel{(E-App2)}
+\UnaryInfC{$\var{v}_1 \: \term{t}_2 \to \var{v}_1 \:  \term{t}_2'$}	
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{}
+\RightLabel{E-AppAbs}
+\UnaryInfC{$(\lambda \var{x} . \term{t}) \var{v}_2 \to [\var{x} \mapsto \var{v}_2] \term{t}$}
+\end{prooftree}
+
+\subsubsection{Typing}
+
+\begin{prooftree}
+\AxiomC{$\var{x}:\type{T} \in \Gamma$}
+\RightLabel{T-Var}
+\UnaryInfC{$\Gamma \vdash \var{x}:\type{T}$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\Gamma, \var{x}:\type{T_1} \vdash \term{t_2}:\type{T_2}$}
+\RightLabel{T-Abs}
+\UnaryInfC{$\Gamma \vdash \lambda \var{x}:\type{T_1}.\term{t_2} : \type{T_1} \to \type{T_2}$}
+\end{prooftree}
+
+\begin{prooftree}
+\AxiomC{$\Gamma \vdash \term{t_1}:\type{T_1} \to \type{T_2}$}
+\AxiomC{$\Gamma \vdash \term{t_2}:\type{T_1}$}
+\RightLabel{T-App}
+\BinaryInfC{$\Gamma \vdash \term{t_1} \: \term{t_2} : \type{T_2}$}
+\end{prooftree}
+
+\subsection{Functional Software Analyses}
+We assume that a Software Analysis is a TLC program $G$ that analyzes a single instance of product line $L$.
+
 \section{Lifting Typed Lambda Calculus}
 
-By lifting Typed Lambda Calculus (TLC), we mean providing syntactic transformation rules that lift any program written in TLC by compositionally lifting TLC syntactic constructs.
+The goal of this work is to automatically lift functional software analyses to work on product lines rather than individual product instances.
+Since we assume that functional analyses are written in TLC, then we just need to provide sound syntactic transformation rules that lift each of the TLC syntactic constructs.
 Those transformation rules can be applied to any software analysis written in TLC, resulting in an equivalent analysis that can be applied to a Software Product Line.
@@
 In the following description of the transformation rules we use Haskell syntax.
-This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as the construction of those constructs on top of TLC is standard \cite{Pierce2002}.
+This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as those constructs have standard definitions on top of TLC \cite{Pierce2002}.
 
-When lifting an analysis from products to product lines, we need to appropriately tag variable artifacts with their corresponding presence conditions.
+When lifting an analysis from products to product lines, we need to appropriately tag values with their corresponding presence conditions.
 We use the \emph{PresenceCondition} abstract data type for this purpose.
@@
 \subsection{Lifting Types}
-Given a type \textbf{t}, we lift it into type \textbf{Lifted t} which is:
+Given a type \textbf{t}, we lift it into type \textbf{Var t} which is:
 
 \begin{minted}{haskell}
-type Lifted t = [(t, PresenceCondition)]
+type Var t = [(Maybe t, PresenceCondition)]
 \end{minted}
@@
 \item \textbf{t} is a type parameter that can be instantiated with any type
-\item \textbf{(t, PresenceCondition)} is a pair of a value of type \textbf{t} together with a \textbf{PresenceCondition}
-\item \textbf{[(t, PresenceCondition)]} is a list of such pairs
-\item \textbf{Lifted t $\equiv$ [(t, PresenceCondition)]} is a parametric data type, instantiated by passing the type argument \textbf{t} to the type constructor \textbf{Lifted}
+\item \textbf{Maybe t} is  an Option type.
+Elements of this type are either \textbf{Nothing}, or \textbf{Just x}, where \textbf{x} is of type \textbf{t}.
+We use an option type because a value does not necessarily have to exist in all products of the product line.
+\item \textbf{(Maybe t, PresenceCondition)} is a pair of a value of type \textbf{t} together with a \textbf{PresenceCondition}
+\item \textbf{[(Maybe t, PresenceCondition)]} is a list of such pairs
+\item \textbf{Var t $\equiv$ [(Maybe t, PresenceCondition)]} is a parametric data type, instantiated by passing the type argument \textbf{t} to the type constructor \textbf{Var}
 \end{itemize}
@@
 
+Another invariant is that a \textbf{Var t} value covers the full product space:
+
+\newtheorem*{covInv*}{Full Coverage Invariant}
+\begin{covInv*}
+Given a lifted value $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\bigvee_{i} pc_i = \top $
+\end{covInv*}
+
+This invariant is not necessary for soundness, but it makes formal treatment of this framework more straightforward.
+It simply states that the presence conditions of a \textbf{Var t} value form a partition of the truth assignments over the feature set $F$.
+
 \subsection{Lifting Values}
 
-To lift a value of type \textbf{t}, all we need to is to provide a presence condition for it, and then encapsulate both into a value of type \textbf{Lifted t}.
+To lift a value of type \textbf{t}, all we need to is to provide a presence condition for it, and then encapsulate both into a value of type \textbf{Var t}.
 This is what the following \textbf{lift} operator does:
@@
 \begin{minted}{haskell}
-lift :: PresenceCondition -> t -> Lifted t
-lift pc x = [(x, pc)]
+lift :: PresenceCondition -> t -> Var t
+lift pc x = [(Just x,pc), (Nothing, neg pc)]
 \end{minted}
@@
 \begin{minted}{haskell}
-liftT :: t -> Lifted t
+liftT :: t -> Var t
 liftT x = lift True x
@@
 
-Assuming we have two features A and B, variables x and y of type \textbf{Lifted Int} and variable z of type \textbf{Lifted String} can have the following values: \\
+Assuming we have two features A and B, variables x and y of type \textbf{Var Int} and variable z of type \textbf{Var String} can have the following values: \\
 
-x = [(1,A),(-2, $\neg A \wedge B$),(3, $\neg A \wedge \neg B$)]
+x = [(Just 1,A),(Just (-2), $\neg A \wedge B$),(Just 3, $\neg A \wedge \neg B$)]
 
-y = [(5, $A \wedge \neg B$), (4,B), (-3, $\neg A \wedge \neg B$)]
+y = [(Just 5, $A \wedge \neg B$), (Just 4,B), (Just (-3), $\neg A \wedge \neg B$)]
 
-z = [("hello", $\neg A \wedge B$),("helloooo", A)] \\
-
-Both x and y satisfy the disjointness invariant.
-It is worth mentioning that a lifted value does not have to be defined in each product in the product line.
-For example, z is not defined in the product identified by the presence condition $(\neg A \wedge \neg B)$.
+z = [(Just "hello", $\neg A \wedge B$),(Just "helloooo", A), (Nothing, $\neg A \wedge \neg B$] \\
 
@@
 This is how data propagates through a functional program.
-Starting with the simplest case, where we have a lifted unary function of type \textbf{Lifted (a $\to$ b)} and an argument of type \textbf{Lifted a} we need to provide a way to apply that function to its argument resulting in a \textbf{Lifted b} value.
-Recall that a lifted value of type \textbf{Lifted t} is just a collection of values of type \textbf{t}, each with a corresponding presence condition, and the whole lifted value has to satisfy the disjointness Invariant.
+Starting with the simplest case, where we have a lifted unary function of type \textbf{Var (a $\to$ b)} and an argument of type \textbf{Var a} we need to provide a way to apply that function to its argument resulting in a \textbf{Var b} value.
+
+Recall that a lifted value of type \textbf{Var t} is just a collection of values of type \textbf{t}, each with a corresponding presence condition, and the whole lifted value has to satisfy the disjointness Invariant.
 So to apply a collection of functions to a collection of values, all we need to is to apply each of the functions to each of the values.
@@
 \begin{minted}{haskell}
-apply :: Lifted (a -> b) -> Lifted a -> Lifted b
-apply fn x = [(fnVal xVal, conj [fnPC,xPC]) 
-              | (fnVal,fnPC) <- fn, (xVal,xPC) <- x, sat(conj[fnPC,aPC])] 
+apply :: Var (a -> b) -> Var a -> Var b
+apply (Var fn) (Var x) =
+  Var [(case (fn', x') of
+    (Just fn'', Just x'') -> Just (fn'' x'')
+    (_,_) -> Nothing
+  , pc) 
+       | (fn', fnpc) <- fn,
+         (x', xpc) <- x,
+         let pc = conj[fnpc, xpc],
+         (sat pc)]
 \end{minted}
-
-Here we are using Haskell's list comprehension to build the lifted result of applying a lifted function to a single lifted argument.
-The function \textbf{conj} takes a list of presence conditions are returns their conjunction, and the function \textbf{sat} takes a presence condition (a symbolic proposition in general) and returns true if its argument is satisfiable, otherwise it returns false.
 
@@
 
-[(1,A),(2, $\neg A \wedge B$), (3, $\neg A\wedge \neg B$)].
+[(Just 1,A),(Just 2, $\neg A \wedge B$), (Just 3, $\neg A\wedge \neg B$)].
 
@@
 
-But since multi-parameter functions are actually Curried unary functions, \emph{intermediateVal} is a lifted unary function [((plus 1), A),((plus -2), $\neg A \wedge B$),((plus 3), $\neg A \wedge \neg B$)].
+But since multi-parameter functions are actually Curried unary functions, \emph{intermediateVal} is a lifted unary function [(Just (plus 1), A),(Just (plus -2), $\neg A \wedge B$),(Just (plus 3), $\neg A \wedge \neg B$)].
 
@@
 
-[(6, $A \wedge \neg B$), (5, $A \wedge B$), (2, $\neg A \wedge B$), (0, $\neg A \wedge \neg B$)] \\
+[(Just 6, $A \wedge \neg B$), (Just 5, $A \wedge B$), (Just 2, $\neg A \wedge B$), (Just 0, $\neg A \wedge \neg B$)] \\
 
-Remember that apply generates the cross product of its arguments (each of them a list of values labeled with presence conditions), and excludes the pairs where the conjunction of presence conditions is not satisfiable.
+Recall that apply generates the cross product of its arguments (each of them a list of values labeled with presence conditions), and excludes the pairs where the conjunction of presence conditions is not satisfiable.
 This is why finalVal has only four possible values, although the cross product initially had 9.
@@
 \begin{minted}{haskell}
-apply2 :: Lifted (a -> b -> c) -> Lifted a -> Lifted b -> Lifted c
+apply2 :: Var (a -> b -> c) -> 
+	  Var a -> Var b -> Var c
 apply2 fn a b = apply (apply fn a) b
@@
 \begin{minted}{haskell}
-apply3 :: Lifted (a -> b -> c -> d) -> Lifted a -> Lifted b -> Lifted c -> Lifted d
+apply3 :: Var (a -> b -> c -> d) -> 
+          Var a -> Var b -> Var c -> Var d
 apply3 fn a b c = apply (apply2 fn a b) c
-
-apply4 :: Lifted (a -> b -> c -> d -> e) -> 
-			Lifted a -> Lifted b -> Lifted c -> Lifted d -> Lifted e
-apply4 fn a b c d = apply (apply3 fn a b c) d
-
 ...
@@
 
-condLifted (liftedEq x y) (liftedPlus x y) (liftedMinus x y)
+condLifted (liftedEq x y) 
+           (liftedPlus x y) 
+           (liftedMinus x y)
 \end{minted}
 
-Please not that Haskell's lazy evaluation makes the de-sugared \emph{cond} function semantically equivalent to a conditional expression because only one of the \emph{then} and \emph{else} expressions is evaluated after the boolean condition is evaluated based on its truth value.
+Please note that Haskell's lazy evaluation makes the de-sugared \emph{cond} function semantically equivalent to a conditional expression because only one of the \emph{then} and \emph{else} expressions is evaluated after the boolean condition is evaluated based on its truth value.
 However, a language with strict evaluation semantics would evaluate both the \emph{then} and \emph{else} expressions before they are passed to \emph{cond}.
@@
 
-\subsubsection{The Disjointness Invariant and Function Application}
+%\subsubsection{Lifting Composite Types}
 
-Function application is the only way of generating new lifted values out of existing ones.
-To preserve the soundness of our lifting system, we need to make sure that function application preserves Invariant 1.
+%\subsubsection{Function Rewriting}
+
+%\subsection{Lifting Parametric Types}
+
+%\section{Formal Definitions and Theorems}
+
+%\subsection{Definitions}
+%\subsubsection{Product Line Values}
+%Given a type $t \in T$,  a value $\var{v}$ of type $\type{t}$, and the presence condition of $\var{v}$ is $\var{pc}$, then the pair $(\var{v}, \term{pc})$ is of the polymorphic type $\type{SPLVal} \: \type{t}$.
+
+%\begin{prooftree}
+%\AxiomC{$\type{t} \in T$}
+%\AxiomC{$\var{v}:\type{t}$}
+%\AxiomC{$\phi (\var{v}) = \term{pc}$}
+%\RightLabel{(T-SPLVal)}
+%\TrinaryInfC{$(\var{v}, \term{pc}) : \type{(SPLVal \: \type{t})}$}	
+%\end{prooftree}
+
+%\begin{minted}{haskell}
+%type SPLVal t = (t, PresenceCondition)
+%\end{minted}
+
+%\subsubsection{Product Line Variables}
+
+%Given a type $t \in T$, the set $\{(x_1, pc_1),...,(x_n, pc_n)\}$ is of type $(\type{SPLVar} \: \type{t})$, where:
+%\begin{itemize}
+%\item $\forall i, (x_i, pc_i) : (\type{SPLVal} \: \type{t})$ (Correctness)
+%\item $\forall (i \neq j), {unsat(\var{pc}_i \wedge \var{pc}_j)}$ (Disjointness)
+%\end{itemize}
+
+%\begin{prooftree}
+%\AxiomC{$\type{t} \in T$}
+%\AxiomC{$(\var{x_1}, \var{pc_1}), ... , (\var{x_n}, \var{pc_n}) : (\type{SPLVal} \: \type{t})$}
+%\AxiomC{$\forall (i \neq j), {unsat(\var{pc}_i \wedge \var{pc}_j)}$}
+%\RightLabel{(T-SPLVar)}
+%\TrinaryInfC{$\{(x_1, pc_1),...,(x_n, pc_n)\} : (\type{SPLVar} \: \type{t})$}	
+%\end{prooftree}
+
+%\begin{minted}{haskell}
+%type SPLVar t = [SPLVal t]
+%\end{minted}
+
+%\newtheorem*{prodValid*}{Product Validity}
+%\begin{prodValid*}
+%Given an SPLVar $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i: {pc_i \Rightarrow \Phi}$
+%\end{prodValid*}
+
+%\newtheorem*{correctness*}{Correctness}
+%\begin{correctness*}
+%Given an SPLVar $v'$ that corresponds to a single-product variable $v$ in $G$, $(x,pc) \in v' \Leftrightarrow \forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket v \rrbracket = x}$
+%\end{correctness*}
+
+%\newtheorem*{disjInv*}{Disjointness}
+%\begin{disjInv*}
+%Given an SPLVar $v = [(v_0, pc_0), ... (v_n, pc_n)]$, $\forall i \neq j: {unsat (pc_i \wedge pc_j)}$
+%\end{disjInv*}
+
+%\subsubsection{Function Definition}
+%Given a function $\var{f} : \type{a_1} \to ... \to \type{a_n} \to \type{b}$ where $\exists i, \type{a_i} \in T$, \\
+%we define $\var{f'} : \type{SPLVar} \: (\type{a_1} \to ... \to \type{a_n} \to \type{b})$, where $\var{f'} = \{(\var{f}, True)\}$
+
+%\begin{prooftree}
+%\AxiomC{$\var{f} : \type{a_1} \to ... \to \type{a_n} \to \type{b}$}
+%\AxiomC{$\exists i, \type{a_i} \in T$}
+%\RightLabel{(T-SPLVar-f)}
+%\BinaryInfC{$\var{f'} : \type{SPLVar} \: (\type{a_1} \to ... \to \type{a_n} \to \type{b}) = \{(\var{f}, True)\}$}	
+%\end{prooftree}
+
+%\begin{minted}{haskell}
+%f' :: SPLVar (a -> b)
+%f' = [(f, True)]
+%\end{minted}
+
+%Here the newly defined SPL function variable $f'$ is a singleton wrapper of the original function $f$, and making it applicable to all products.
+The SPLVar $f'$ obviously satisfies the disjointness condition, and because it is a fresh variable that does not exist in $G$ the correctness condition does not apply to it.
+
+%\subsubsection{Function Application}
+
+%Given a function $\var{f'} : \type{SPLVar} \: (\type{a} \to \type{b})$ 
+%and a variable $\var{x} : (\type{SPLVar} \: \type{a})$, 
+%the result of applying $\var{f'}$ to $\var{x}$ is the set 
+%$\{(\var{f''}(\var{x'}), \var{fpc} \wedge \var{xpc}) \: | \: (\var{f''}, \var{fpc}) \in \var{f'}, (\var{x'}, \var{xpc}) \in x, sat(\var{fpc} \wedge \var{xpc})\}$
+
+%\begin{prooftree}
+%\AxiomC{$\var{f'} : \type{SPLVar} \: (\type{a} \to \type{b})$}
+%\AxiomC{$\var{x} : (\type{SPLVar} \: \type{a})$}
+%\RightLabel{(E-SPL-apply)}
+%\BinaryInfC{$(\var{apply} \: \var{f'} \: \var{x}) = \{(\var{f''}(\var{x'}), \var{fpc} \wedge \var{xpc}) \: | \: (\var{f''}, \var{fpc}) \in \var{f'}, (\var{x'}, \var{xpc}) \in x, sat(\var{fpc} \wedge \var{xpc})\}$}
+%\end{prooftree}
+
+%\begin{minted}{haskell}
+%apply :: SPLVar (a -> b) -> SPLVar a -> SPLVar b
+%apply fn x = [(fnVal xVal, conj [fnPC,xPC]) 
+%              | (fnVal,fnPC) <- fn, (xVal,xPC) <- x, sat(conj[fnPC,aPC])] 
+%\end{minted}
+
+%Here we are using Haskell's list comprehension syntax to build the (SPLVar b) result of applying a lifted function to a single argument of type (SPLVar a).
+The function \textbf{conj} takes a list of presence conditions and returns their conjunction, and the function \textbf{sat} takes a presence condition (a symbolic proposition in general) and returns true if its argument is satisfiable, otherwise it returns false.
+
+%As in almost all functional programming languages, Currying is the standard mechanism to extend application of single-argument functions to multi-argument functions.
+We provide versions of apply for different arities:
+
+%\begin{minted}{haskell}
+%apply2 :: SPLVar (a -> b -> c) -> SPLVar a -> SPLVar b -> SPLVar c
+%apply2 fn a = apply (apply fn a)
+
+%apply3 :: SPLVar (a -> b -> c -> d) -> SPLVar a -> SPLVari b -> SPLVar c -> SPLVar d
+%apply3 fn a b = apply (apply2 fn a b)
+%\end{minted}
+
+\subsection{Theorems}
+
+\subsubsection{Application Preserves Disjointness}
 
@@
 \begin{th1}
-If fn and x satisfy Invariant 1, then the result of (apply fn x) also satisfies the invariant.
+If f and x satisfy the disjointness condition, then the result of (apply f x) also satisfies disjointness.
 \end{th1}
@@
 \begin{itemize}
-\item Assume that both fn and x satisfy the invariant while the result y = (apply fn x) does not
+\item Assume that both f and x satisfy disjointness while the result y = (apply f x) does not
 \item Then there exists at least one pair of presence conditions in y ($pc_i$ and $pc_j$) where sat($pc_i \wedge pc_j$) = true
-\item Since presence conditions $pc_i$ and $pc_j$ where constructed by apply as conjunctions of presence conditions from fn and x, and we know presence conditions of fn are disjoint (same for those of x), then there is no truth assignment that would satisfy the conjunction of any of the pairs of presence conditions in y.
+\item Each of $pc_i$ and $pc_j$ is a conjunction of presence conditions from f and x (definition of apply), then for arbitrary a, b, c and d: \\
+               $pc_i = fpc_a \wedge xpc_b$ \\
+               $pc_j = fpc_c \wedge xpc_d$
+\item Then $pc_i \wedge pc_j = pc_a \wedge xpc_b \wedge fpc_c \wedge xpc_d$
+\item For this conjunction to be satisfiable, we must have both sat($fpc_a \wedge xpc_b$) and sat($fpc_c \wedge xpc_d$), which contradicts the assumption that f and x both satisfy disjointness
+\item Then y = (apply f x) must satisfy disjointness too
 \end{itemize}
@@
 
-\subsubsection{Lifting Composite Types}
+\subsubsection{Application Preserves Correctness}
 
-\subsubsection{Function Rewriting}
+\newtheorem{th2}{Theorem}
+\begin{th2}
+If f' and x' satisfy the correctness condition with respect to original variables f and x respectively, then the result of (apply f' x') also satisfies correctness with respect to (f x).
+\end{th2}
 
-\subsection{Lifting Parametric Types}
+\begin{proof}
+Given v' = (apply f' x'), we need to prove that $(y, pc) \in v' \Leftrightarrow \forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket f(x) \rrbracket = y}$
+\begin{itemize}
+\item{$\Rightarrow$}
+	\begin{itemize}
+	\item Assume that $(y,pc) \in v'$ and for some arbitrary truth assignment $e$, $e \Rightarrow pc$
+	\item Then by definition of apply, there exists $(f'', fpc) \in f'$, $(x'', xpc) \in x$, such that y = f''(x'') and pc = $fpc \wedge xpc$
+	\item Since $e \Rightarrow pc$, then $e \Rightarrow fpc \wedge xpc$, which means that $e \Rightarrow fpc$ and $e \Rightarrow xpc$
+	\item By definition of $G(e)$ and the assumption that both f' and x' satisfy the correctness condition, $G(e) \llbracket f \rrbracket = f'$, and $G(e) \llbracket x \rrbracket = x'$
+	\item Then $G(e) \llbracket f(x) \rrbracket = y$
+	\end{itemize}
+\item{$\Leftarrow$}
+	\begin{itemize}
+	\item Assume $\forall e:{(e \Rightarrow pc) \Rightarrow G(e) \llbracket f(x) \rrbracket = y}$
+	\item Then given an arbitrary truth assignment $e$ (identifying a single product) where $e \Rightarrow pc$, $G(e) \llbracket f(x) \rrbracket = y$
+	\item Since f' and x' are both assumed to satisfy the disjointness and correctness conditions, then $(f,fpc) \in f'$ and $(x,xpc) \in x'$ where $pc = fpc \wedge xpc$
+	\item Then by definition of apply, $(y,pc) \in$ (apply f' x')
+	\end{itemize}
+\end{itemize}
 
-\section{Examples and Evaluation}
+\end{proof}
+
+\section{Evaluation}
 
@@
 
-\subsection{TODO: Some other analysis}
+%\subsection{TODO: Some other analysis}
 
-\section{Discussion}
-\subsection{Deep vs Shallow Lifting}
+%\section{Discussion}
+%\subsection{Deep vs Shallow Lifting}
 
 \section{Related Work}
+TODO
+
+% An example of a floating figure using the graphicx package.
+% Note that \label must occur AFTER (or within) \caption.
+% For figures, \caption should occur after the \includegraphics.
+% Note that IEEEtran v1.7 and later has special internal code that
+% is designed to preserve the operation of \label within \caption
+% even when the captionsoff option is in effect. However, because
+% of issues like this, it may be the safest practice to put all your
+% \label just after \caption rather than within \caption{}.
+%
+% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
+% option should be used if it is desired that the figures are to be
+% displayed while in draft mode.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation results for the network.}
+%\label{fig_sim}
+%\end{figure}
+
+% Note that the IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
+
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command,
+% and the \label for the overall figure must come after \caption.
+% \hfil is used as a separator to get equal spacing.
+% Watch out that the combined width of all the subfigures on a 
+% line do not exceed the text width or a line break will occur.
+%
+%\begin{figure*}[!t]
+%\centering
+%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
+%\label{fig_second_case}}
+%\caption{Simulation results for the network.}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat[]), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
+% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
+% labels, the optional argument to \subfloat must be present. If a
+% subcaption is not desired, just leave its contents blank,
+% e.g., \subfloat[].
+
+% An example of a floating table. Note that, for IEEE style tables, the
+% \caption command should come BEFORE the table and, given that table
+% captions serve much like titles, are usually capitalized except for words
+% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
+% and up, which are usually not capitalized unless they are the first or
+% last word of the caption. Table text will default to \footnotesize as
+% the IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
+%\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that the IEEE does not put floats in the very first column
+% - or typically anywhere on the first page for that matter. Also,
+% in-text middle ("here") positioning is typically not used, but it
+% is allowed and encouraged for Computer Society conferences (but
+% not Computer Society journals). Most IEEE journals/conferences use
+% top floats exclusively. 
+% Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the
+% \fnbelowfloat command of the stfloats package.
 
 \section{Conclusion}
+TODO
 
-\bibliography{/users/ramy/bib/library} 
+% conference papers do not normally have an appendix
+
+% use section* for acknowledgment
+%\section*{Acknowledgment}
+
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
+
+% references section
+
+% can use a bibliography generated by BibTeX as a .bbl file
+% BibTeX documentation can be easily obtained at:
+% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
+% The IEEEtran BibTeX style support page is at:
+% http://www.michaelshell.org/tex/ieeetran/bibtex/
+%\bibliographystyle{IEEEtran}
+% argument is your BibTeX string definitions and bibliography database(s)
+%\bibliography{IEEEabrv,../bib/paper}
+%
+% <OR> manually copy in the resultant .bbl file
+% set second argument of \begin to the number of references
+% (used to reserve space for the reference number labels box)
+
+\bibliography{paper} 
 \bibliographystyle{ieeetr}
 
+% that's all folks
 \end{document}

### 2017-04-10 SPL/src/Deep/VList.hs
Lifted lists - under development

@@ -17,8 +17,30 @@ type VList a = Var (VList' a)
 vNil = mkVarT VNil'
 
 vCons :: Var a -> VList a -> VList a 
-vCons x xs = (liftV2 VCons') x (mkVarT xs)
+vCons x@(Var x') xs = 
+  Var ([(Just (VCons' x'' (subst xs xpc)), xpc) | (Just x'',xpc) <- x'] ++ 
+       nothingEls)
+  where nothingPC = undefinedAt x
+        nothingXs = subst xs nothingPC
+        nothingEls= case nothingXs of
+                          Var xs' -> (map (\(x,pc) -> (x,conj[pc, nothingPC])) xs') 
 
+head' :: VList' a -> a 
+head' xs = case xs of
+              VCons' x xs'' -> x 
+              _ -> error ""
+
+vHead :: VList a -> Var a 
+vHead = liftV head'
+
+tail' :: VList' a -> VList a
+tail' xs = case xs of
+            VNil' -> error ""
+            VCons' x xs -> xs
+
+vTail :: VList a -> VList a
+vTail (Var xs) =
+  Var 
 -- test
 u :: Universe
 u = mkUniverse ["P", "Q", "R", "S"]

### 2017-04-10 SPL/src/SPL.hs

@@ -78,9 +78,23 @@ valIndex (Var v) x =
                             _        -> False) 
                         v
 
-subst :: Show t => Var t -> PresenceCondition -> Var t
+subst :: Var t -> PresenceCondition -> Var t
 subst (Var v) pc =
     Var (filter (\(_,pc') -> sat (conj [pc,pc'])) v)
+
+definedAt :: Var t -> PresenceCondition
+definedAt (Var xs) = disj(pcs)
+    where   pcs     = map snd pairs
+            pairs   = filter (\(x,pc) -> case x of
+                                            Just _  ->  True
+                                            Nothing ->  False) xs
+
+undefinedAt :: Var t -> PresenceCondition
+undefinedAt (Var xs) = disj(pcs)
+    where   pcs     = map snd pairs
+            pairs   = filter (\(x,pc) -> case x of 
+                                            Just _  ->  False
+                                            Nothing ->  True) xs
 
 restrict :: PresenceCondition -> Var t -> Var t
 restrict pc (Var v) =

### 2017-04-10 SPL/src/Shallow/VList.hs

@@ -4,8 +4,8 @@ import SPL
 
 type VList a = Var [a]
 
-ve :: Var [a]
-ve = mkVarT []
+vNil :: Var [a]
+vNil = mkVarT []
 
-vcons :: Var a -> Var [a] -> Var [a]
-vcons = liftV2 (:)
+vCons :: Var a -> Var [a] -> Var [a]
+vCons = liftV2 (:)

### 2017-04-11 SPL/src/Deep/VList.hs
Deep.VList head and tail

@@ -9,7 +9,7 @@ data List a =
 
 data VList' a =
     VNil'
-  | VCons' a (VList a)
+  | VCons' (Var a) (VList a)
   deriving Show
   
 type VList a = Var (VList' a)
@@ -17,21 +17,18 @@ type VList a = Var (VList' a)
 vNil = mkVarT VNil'
 
 vCons :: Var a -> VList a -> VList a 
-vCons x@(Var x') xs = 
-  Var ([(Just (VCons' x'' (subst xs xpc)), xpc) | (Just x'',xpc) <- x'] ++ 
-       nothingEls)
-  where nothingPC = undefinedAt x
-        nothingXs = subst xs nothingPC
-        nothingEls= case nothingXs of
-                          Var xs' -> (map (\(x,pc) -> (x,conj[pc, nothingPC])) xs') 
+vCons x@(Var x') xs@(Var xs') = union def undef 
+  where xdef = defSubst x
+        def = Var [(Just (VCons' hd (Var[p])), (definedAt hd)) | p@(xs'',xspc) <- xs', let hd = (restrict xspc xdef)]
+        undef = restrict (undefinedAt x) xs
 
-head' :: VList' a -> a 
+head' :: VList' a -> Var a 
 head' xs = case xs of
               VCons' x xs'' -> x 
               _ -> error ""
 
 vHead :: VList a -> Var a 
-vHead = liftV head'
+vHead xs = union2 $ (liftV head') xs
 
 tail' :: VList' a -> VList a
 tail' xs = case xs of
@@ -39,8 +36,8 @@ tail' xs = case xs of
             VCons' x xs -> xs
 
 vTail :: VList a -> VList a
-vTail (Var xs) =
-  Var 
+vTail xs = union2 $ (liftV tail') xs
+
 -- test
 u :: Universe
 u = mkUniverse ["P", "Q", "R", "S"]
@@ -68,3 +65,9 @@ y = mkVars [(-11, _p)]
 
 z :: Var Int
 z = mkVars [(6, p)]
+
+l0 = vNil
+l1 = vCons w l0
+l2 = vCons x l1
+l3 = vCons y l2
+l4 = vCons z l3

### 2017-04-11 SPL/src/SPL.hs

@@ -56,8 +56,9 @@ mkVarT v = Var [(Just v,T)]
 mkVars :: [(t,PresenceCondition)] -> Var t
 mkVars vs = let nothingPC = (neg . disj) (map snd vs) 
                 vs'       = map (\(v,pc) -> (Just v, pc)) vs
-            in  Var ((Nothing, nothingPC) : vs')
-
+            in  if (sat nothingPC)
+                then Var ((Nothing, nothingPC) : vs')
+                else Var vs'
 
 --mkVars vs = Var (map (\(v,pc) -> (Just v, pc)) vs)
 
@@ -98,11 +99,23 @@ undefinedAt (Var xs) = disj(pcs)
 
 restrict :: PresenceCondition -> Var t -> Var t
 restrict pc (Var v) =
-    Var $ map (\(x,pc') -> (x, conj[pc',pc])) v
+    Var $ filter (\(x,pc') -> sat pc') (map (\(x,pc') -> (x, conj[pc',pc])) v)
 
+defSubst :: Var t -> Var t 
+defSubst (Var v) = Var (filter (\(x,_) -> case x of
+                                            Just _  -> True
+                                            _       -> False) v)
+                                    
 union :: Var t -> Var t -> Var t
 union (Var a) (Var b) =
     Var (a ++ b)
+
+unions :: [Var t] -> Var t 
+unions xs = Var (foldr (++) [] (map (\(Var v) -> v) xs))
+
+union2 :: Var (Var t) -> Var t 
+union2 xs = unions (map (\(Just x,pc) -> (restrict pc x)) xs')
+    where (Var xs') = (defSubst xs)
 
 pairs :: [t] -> [(t,t)]
 pairs [] = []

### 2017-09-21 vcpp/app/Main.hs
--

@@ -12,6 +12,4 @@ main = do
     args <- getArgs
     file <- readFile (args!!0)
     Var v <- vcpp file
-    mapM_ (\(x,pc) -> case x of
-                        Just x' -> putStrLn ("#" ++ (show pc) ++ "#\n" ++ x' ++ "#---#")
-                        Nothing -> putStrLn ((show pc) ++ ": Nothing\n#---#")) v
+    mapM_ (\(x,pc) -> putStrLn ("#" ++ (show pc) ++ "#\n" ++ x ++ "#---#")) v

### 2017-09-21 vcpp/src/vcpp.hs

@@ -44,8 +44,7 @@ lineByLine (x:xs) e@(CPPEnv feats env) =
             return ((mkVars [(x,envProp), ("", neg envProp)]) |:| rest)
         -- #ifdef
         else if token0 == "#ifdef"
-        then let (feats', i) = queryOrUpdate feats token1
-                 p = Atom feats' i
+        then let (feats', p) = queryOrUpdate feats token1
                  env' = stackPush env p 
              in  do
                  rest <- lineByLine xs (CPPEnv feats' env')

### 2017-09-21 doc/SafetyCasesPL.tex
safety case PL document draft 1

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{hyperref}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Safety Cases for Software Product Lines}
%\author{The Author}
\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Introduction}
Software systems are everywhere, including safety critical domains. Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles. Safety analysis of this category of systems is an essential ingredient of the Software Development Lifecycle (SDLC). Safety cases are the most common artifacts resulting from safety analysis.

Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs). Variability is inherent in families of software systems. Variability can be due to different deployment environments, differences in hardware devices and peripherals, varying requirements across models of systems, and difference in performance characteristics.

Variability in an SPL is usually measured in terms of the number of features in a product line. The number of possible products that can be synthesized from an SPL grows exponentially with the number of features. As a result, enumerating all possible products and performing safety analysis at the product level is intractable.

Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular. This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.

\subsection{Goals}
This document aims at:
\begin{itemize}
\item Surveying existing techniques
\item Identifying gaps and pragmatic requirements that are not fulfilled by existing techniques
\item Identifying any underlying research questions
\end{itemize}

\subsection{Non-Goals}

The following are beyond the scope of this document:

\begin{itemize}
\item Coming up with new notations or techniques
\item Empirical assessment of existing techniques
\item Formalizing any of the discussed concepts
\end{itemize}

\subsection{Running Example}

Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}. This is a simply product line safety case that extends a single product ISO26262 safety case \cite{Dardar_2014}. The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.

\section{Existing Techniques}

\subsection{Software Product Line Definition}

Software Product Lines are composed of a common core (common across all product variants) and variable features. Different approaches have been used to define SPLs:

\subsubsection{Generative}

In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts. Examples of Generative Product Line approaches include Feature-Oriented Programming and Aspect-Oriented Programming.

\subsubsection{Annotative}

In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (presence conditions) are used to annotate different syntactic constructs belonging to different families of products. The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.

\subsubsection{Delta-oriented}

In Delta-Oriented Product Lines, a base product is fully defined, and variants are defined in terms of their differences with respect to the base product.

\subsection{Safety Cases}

\subsubsection{Goal Structuring Notation (GSN)}

Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn}
  \caption{GSN Elements}
\end{figure}

\begin{itemize}
\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
\end{itemize}

\subsubsection{Structured Assurance Case Metamodel (SACM)}

SACM \cite{SACM} is an Object Management Group (OMG) standard. SACM integrates well with other OMG metamodels (e.g. UML). It provides a much richer notation compared to GSN. The notation is composed of the following metamodels:

\begin{itemize}
\item An Argumentation Meta-model
\item An Evidence Meta-model
\end{itemize}

\section{Adding Variability}

\subsection{GSN Modules and Patterns}

\subsubsection{GSN Modules}

Modules are a GSN extension that facilitate modularity and reuse in safety cases. GSN elements (goals, evidences and strategies) can be packages into modules, and those modules can be reused in different contexts.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn-modules}
  \caption{GSN Module Constructs}
\end{figure}

\subsubsection{GSN Patterns}

GSN patterns are yet another GSN extension. A GSN pattern is a parameterized GSN item. A pattern can then be instantiated by binding a parameter to a concrete value.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn-patterns}
  \caption{GSN Pattern Constructs}
\end{figure}

As an example of a GSN pattern, a generic hazard avoidance pattern would be in the context of an identified hazard, aiming at assuring that the system is safe by addressing that hazard.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{gsn-pattern-example}
  \caption{GSN Pattern Example - Hazard mitigation}
\end{figure}

\subsubsection{GSN Product Line Example}

This is just a fragment of the safety case in \cite{Gallucci_2013}, demonstrating the use of GSN modules:

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{safetyCase}
  \caption{Safety Case Fragment}
\end{figure}

%3.2	Comparative Analysis
%
%GSN+
%SACM?
%GM
%Expressivity
%
%
%
%Syntactic Conciseness
%
%
%
%Well-defined Semantics?
%
%
%
%Semantic Consistency
%
%
%
%Traceability within a single product
%
%
%
%Traceability across products
%
%
%
%

\section{Discussion}

\subsection{Product Line Scenarios}

Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios yielding Product Lines.

\subsubsection{Product Evolution over Time}

Starting with a single product, variability might start exhibiting itself in lots of different ways. Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability. Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem. 
One way to manage ad-hoc variability is to evolve a single product with variability in a product line. This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.

\subsubsection{Product Component Reuse}

Again starting with a single component-oriented product, some components might be potentially reused in other products. Sometimes a component can be reused as is, without the need for any modification. In other cases the component might require minor modifications, resulting in a slightly different version.

This component-level variability can again evolve into a component-level Product Line. With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.

\subsubsection{Product-Line Engineering}

The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line. Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one. If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.

\subsection{Product Lines and Safety Cases}

Reasoning about Software Product Lines is not straightforward because of several reasons:

\begin{itemize}

\item The number of possible valid products of a product lines grows exponentially with the number of features. As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.

\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations. When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.

\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification).

\end{itemize}

\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}

Safety Analysis is one example of reasoning about a software systems. Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products. However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons.

In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:

\begin{itemize}

\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.

\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts).

\end{itemize}

\subsection{Product Line evolution impact analysis}

As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well. This is essential to maintain the validity of the safety arguments. However, the safety assurance process is expensive in terms of resources and is also time consuming and labor intensive. As a result, as much of existing evidences should be reused as possible, without jeopardizing the validity of the safety arguments. Two questions are relevant here:

\begin{itemize}

\item Q2: As an SPL evolves, which pieces of evidence are still valid, and can be reused as is?

\item Q3: Within a single piece of evidence, how much reuse can be achieved? For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific? How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences?

\end{itemize}

\bibliographystyle{abbrv}
\bibliography{SafetyCasesPL}

\end{document}

### 2017-09-21 doc/taxonomy.tex
safety case PL document draft 1

\documentclass{article}

%% macros
% reference macros
\newcommand{\Classen}{\cite{Classen:2010:MCL:1806799.1806850}}
\begin{document}

\title{A Taxonomy of Software Product Line Analyses}
\maketitle

\abstract
TODO

\section{Introduction}
TODO

\section{Definitions}

\section{Taxonomy}

\subsection{Model Checking}

\subsubsection{FTS}
A Featured-Transition System (FTS)\Classen is a modeling abstraction that extends a Labeled Transition System (LTS) to represent a family (product line) of models rather than a single model. 
\subsection{Data-Flow Analyses}
\subsection{Abstract Interpretation}
\subsection{Parsing}

\section{Discussion}
TODO

\section{Conclusion}
TODO

\bibliographystyle{ieeetr}
\bibliography{taxonomy} 
\end{document}

### 2017-09-22 doc/SafetyCases/SafetyCasesPL.tex
Moving the safety cases document into a separate directory

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{hyperref}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Safety Cases for Software Product Lines}
%\author{The Author}
\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
%
%
\section{Introduction}
%Software systems are everywhere, including safety critical domains. Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles. Safety analysis of this category of systems is an essential ingredient of the Software Development Lifecycle (SDLC). Safety cases are the most common artifacts resulting from safety analysis.
%
%Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs). Variability is inherent in families of software systems. Variability can be due to different deployment environments, differences in hardware devices and peripherals, varying requirements across models of systems, and difference in performance characteristics.
%
%Variability in an SPL is usually measured in terms of the number of features in a product line. The number of possible products that can be synthesized from an SPL grows exponentially with the number of features. As a result, enumerating all possible products and performing safety analysis at the product level is intractable. 
%
%Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular. This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.
%
\subsection{Goals}
%This document aims at:
%\begin{itemize}
%\item Surveying existing techniques
%\item Identifying gaps and pragmatic requirements that are not fulfilled by existing techniques
%\item Identifying any underlying research questions
%\end{itemize}
%
\subsection{Non-Goals} 
%
%The following are beyond the scope of this document:
%
%\begin{itemize}
%\item Coming up with new notations or techniques
%\item Empirical assessment of existing techniques
%\item Formalizing any of the discussed concepts
%\end{itemize}
%
%\subsection{Running Example}
%
%Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}. This is a simply product line safety case that extends a single product ISO26262 safety case \cite{Dardar_2014}. The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.
%
%\section{Existing Techniques}
%
\section{Software Product Lines (SPLs)}

\subsection{Basic Concepts}

Variability is commonplace in software systems. Managing variability across different software artifacts, and throughout the Software Development Life Cycle (SDLC) is always a challenge though. Software Product Line Engineering draws an analogy to industrial product lines, where different variants of a product are manufactured using a common set of parts and machinery. The goal of SPLs is to maximize reuse across different variants of a software product, while at the same time simplifying the extensibility and maintainability of the product line.

We start with some basic definitions, together with some simple examples illustrating them. The definitions and the object store example presented here are adapted from those in \cite{Thum}.

\begin{description}

\item[Software Product Line]
A Software Product Line (SPL) is a set of similar software products built from a common set of artifacts (e.g. design/specification models, code constructs, test cases, etc...).

\end{description}

For example, let us consider a product line of simple object stores. An object store can support the storage of a single object or multiple objects. Orthogonally, the store can provide some access control mechanism allowing only certain users to store objects. Store capacity (SingleObject vs MultiObject) and access control are just examples of possible variability points in an SPL.

\begin{description}

\item[Feature]
A feature is an externally visible property, aspect or quality of a software system.

\end{description}

In our object store SPL, SingleObject, MultiObject and access control are the 3 features within this product line.

An individual product can be generated from an SPL by selecting a set of features for that product. Not all product combinations are valid though. For example, it does not make sense for an object store to be SingleObject and MultiObject at the same time. Those two features are supposed to be mutually exclusive in our product line. This kind of constraints are captured by a \emph{Feature Model}:

\begin{description}

\item[Feature Model]
(also knows as Variability Model) is a specification of the valid combinations of features in an SPL. This is typically expressed as a boolean formula over features.

\end{description}

For example, a boolean formula expressing our mutual exclusion condition for SingleObject and MultiObject stores, together with an optional AccessControl feature, would be:

$(SingleObject \vee MultiObject) \wedge (\neg SingleObject \vee \neg MultiObject) $

\begin{description}

\item[Feature Diagram]
a hierarchical graphical representation of a Feature Model, where each non-root feature depends on its parent.

\end{description}

\begin{figure}
  \centering
    \includegraphics[width=0.75\textwidth]{FeatureDiagram}
  \caption{Feature Diagram Example}
  \label{fig:FeatureDiagram}
\end{figure}

For example, figure \ref{fig:FeatureDiagram} shows a feature diagram for our SPL.

\begin{description}

\item[Presence Condition]
A Presence Condition (PC) is a boolean formula over the set of features that does not conflict with the Feature Model.
\end{description}

For example, a product where we have the MultiObject and AccessControl features can be represented by the following PC: $ MultiObject \wedge AccessControl $

\subsection{SPL Engineering Approaches}

\subsubsection{Generative}

%In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts. Examples of Generative Product Line approaches include Feature-Oriented Programming and Aspect-Oriented Programming.
%
\subsubsection{Annotative}
%
%In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (presence conditions) are used to annotate different syntactic constructs belonging to different families of products. The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
%
\subsubsection{Delta-Oriented}
%
%In Delta-Oriented Product Lines, a base product is fully defined, and variants are defined in terms of their differences with respect to the base product. 
%
\section{Safety Cases}

\subsection{Goal Structuring Notation (GSN)}
%
%Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn}
%  \caption{GSN Elements}
%\end{figure}
%
%\begin{itemize}
%\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
%\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
%\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
%\end{itemize}
%
%\subsubsection{Structured Assurance Case Metamodel (SACM)}
%
%SACM \cite{SACM} is an Object Management Group (OMG) standard. SACM integrates well with other OMG metamodels (e.g. UML). It provides a much richer notation compared to GSN. The notation is composed of the following metamodels:
%
%\begin{itemize}
%\item An Argumentation Meta-model
%\item An Evidence Meta-model
%\end{itemize}
%
%\section{Adding Variability}
%
%\subsection{GSN Modules and Patterns}
%
%\subsubsection{GSN Modules}
%
%Modules are a GSN extension that facilitate modularity and reuse in safety cases. GSN elements (goals, evidences and strategies) can be packages into modules, and those modules can be reused in different contexts.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-modules}
%  \caption{GSN Module Constructs}
%\end{figure}
%
%\subsubsection{GSN Patterns}
%
%GSN patterns are yet another GSN extension. A GSN pattern is a parameterized GSN item. A pattern can then be instantiated by binding a parameter to a concrete value.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-patterns}
%  \caption{GSN Pattern Constructs}
%\end{figure}
%
%As an example of a GSN pattern, a generic hazard avoidance pattern would be in the context of an identified hazard, aiming at assuring that the system is safe by addressing that hazard.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{gsn-pattern-example}
%  \caption{GSN Pattern Example - Hazard mitigation}
%\end{figure}
%
%\subsubsection{GSN Product Line Example}
%
%This is just a fragment of the safety case in \cite{Gallucci_2013}, demonstrating the use of GSN modules:
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.75\textwidth]{safetyCase}
%  \caption{Safety Case Fragment}
%\end{figure}
%
%%3.2	Comparative Analysis
%%
%%GSN+
%%SACM?
%%GM
%%Expressivity
%%
%%
%%
%%Syntactic Conciseness
%%
%%
%%
%%Well-defined Semantics?
%%
%%
%%
%%Semantic Consistency
%%
%%
%%
%%Traceability within a single product
%%
%%
%%
%%Traceability across products
%%
%%
%%
%%
%

\subsection{Example}

\section{Discussion}
%
\subsection{Product Line Scenarios}
%
%Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios yielding Product Lines.
%
\subsubsection{Product Evolution over Time}
%
%Starting with a single product, variability might start exhibiting itself in lots of different ways. Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability. Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem. 
%One way to manage ad-hoc variability is to evolve a single product with variability in a product line. This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.
%
\subsubsection{Product Component Reuse}
%
%Again starting with a single component-oriented product, some components might be potentially reused in other products. Sometimes a component can be reused as is, without the need for any modification. In other cases the component might require minor modifications, resulting in a slightly different version. 
%
%This component-level variability can again evolve into a component-level Product Line. With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.
%
\subsubsection{Product-Line Engineering}
%
%The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line. Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one. If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.
%
\subsection{Product Lines and Safety Cases}
%
%Reasoning about Software Product Lines is not straightforward because of several reasons:
%
%\begin{itemize}
%
%\item The number of possible valid products of a product lines grows exponentially with the number of features. As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.
%
%\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations. When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.
%
%\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification). 
%
%\end{itemize}
%
\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
%
%Safety Analysis is one example of reasoning about a software systems. Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products. However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons. 
%
%In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
%
%\begin{itemize}
%
%\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
%
%\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts). 
%
%\end{itemize}
%
\subsection{Product Line evolution impact analysis}
%
%As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well. This is essential to maintain the validity of the safety arguments. However, the safety assurance process is expensive in terms of resources and is also time consuming and labor intensive. As a result, as much of existing evidences should be reused as possible, without jeopardizing the validity of the safety arguments. Two questions are relevant here:
%
%\begin{itemize}
%
%\item Q2: As an SPL evolves, which pieces of evidence are still valid, and can be reused as is?
%
%\item Q3: Within a single piece of evidence, how much reuse can be achieved? For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific? How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences? 
%
%\end{itemize}
%
\bibliographystyle{abbrv}
\bibliography{SafetyCasesPL}

\end{document}

### 2017-09-22 doc/SafetyCases/SafetyCasesPL.tex
feature-oriented, annotative and delta-oriented examples

@@
 \usepackage{hyperref}
+\usepackage{listings}
+\usepackage{color}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
@@
 
+SPLs are composed of features, and those features can be represented in various ways:
+
 \subsubsection{Generative}
 
-%In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
-Examples of Generative Product Line approaches include Feature-Oriented Programming and Aspect-Oriented Programming.
-%
+In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
+Examples of Generative Product Line approaches include Feature-Oriented Programming \cite{Prehofer1997} and Aspect-Oriented Programming \cite{Kiczales1997}.
+
+\begin{figure}
+  \centering
+    \includegraphics[width=0.75\textwidth]{FeatureOriented}
+  \caption{Feature-Oriented implementation of the object store Product Line}
+  \label{fig:FeatureOriented}
+\end{figure}
+
+Figure \ref{fig:FeatureOriented} shows a possible Feature-Oriented implementation of our object store Product Line.
+\emph{SingleStore} and \emph{MultiStore} are two separate implementations of a their corresponding features, and \emph{AccessControl} is implemented as a cross-cutting feature that refines any of the 2 \emph{Store} implementations.
+
+\begin{figure}
+  \centering
+    \includegraphics[width=0.75\textwidth]{FeatureOrientedGenerated}
+  \caption{The product generated out of the \emph{MultiStore} and \emph{AccessControl} features}
+  \label{fig:FeatureOrientedGenerated}
+\end{figure}
+
+A product can then be generated out of the individual features given a valid Presence Condition.
+For example, if we are to generate a product with the \emph{MultiStore} and \emph{AccessControl} features, the generator would automatically generate the code in figure \ref{fig:FeatureOrientedGenerated}.
+
 \subsubsection{Annotative}
-%
-%In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (presence conditions) are used to annotate different syntactic constructs belonging to different families of products.
+
+Generative approaches are only supported by a limited set of programming languages and tools.
+In mainstream languages, Annotative approaches are typically used, mainly because they don't depend on syntactic support for features, or compiler/runtime support for feature composition.
+
+In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (Presence Conditions) are used to annotate different syntactic constructs belonging to different families of products.
 The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
-%
+
+If we are to rewrite our object store Product Line using CPP macros, it would look like listing 
+ref{lst:Annotative}. A single feature (e.g. \emph{ACCESS\_CONTROL} is now scattered across the code base, and in some cases replicated (e.g. lines 13-15 and 24-26).
+
+\lstset{
+	basicstyle=\ttfamily,
+     keywordstyle=\color{blue}\ttfamily,
+     stringstyle=\color{red}\ttfamily,
+     commentstyle=\color{green}\ttfamily,
+     morecomment=[l][\color{magenta}]{\#},
+	numbers=left, 
+	tabsize=2,
+	captionpos=b, 
+	frame=shadowbox
+}
+
+\begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
+class Store {
+	#if defined (ACCESS_CONTROL)
+	private boolean sealed = false;
+	#endif
+
+	#if defined (SINGLE_STORE)
+	private Object value;
+	Object read() {
+		#if defined (ACCESS_CONTROL)
+ 		if (!sealed) {
+		#endif
+			return value;
+		#if defined (ACCESS_CONTROL)
+		} else { throw new RuntimeException("Access denied!"); }
+		#endif 
+	}
+	#elif defined (MULTI_STORE)
+	private LinkedList values = new LinkedList();
+	Object read() { 
+		#if defined (ACCESS_CONTROL)
+ 		if (!sealed) {
+		#endif
+			return values.getFirst(); 
+		#if defined (ACCESS_CONTROL)
+		} else { throw new RuntimeException("Access denied!"); }
+		#endif
+	}
+	Object readAll() { return values.toArray(); }
+	#endif
+
+	void set(Object nvalue) {
+		#if defined (ACCESS_CONTROL)
+ 		if (!sealed) {
+		#endif
+			#if defined (SINGLE_STORE)
+			value = nvalue;
+			#elif defined (MULTI_STORE)
+			values.addFirst(nvalue);
+			#endif
+		#if defined (ACCESS_CONTROL)
+		} else { throw new RuntimeException("Access denied!"); }
+		#endif
+	}
+}
+\end{lstlisting}
+
 \subsubsection{Delta-Oriented}
-%
-%In Delta-Oriented Product Lines, a base product is fully defined, and variants are defined in terms of their differences with respect to the base product. 
-%
+
+In Delta-Oriented \cite{Schaefer} Product Lines, a base product (referred to as the Core Module) is fully defined, and variants (deltas) are defined in terms of their differences with respect to the base product.
+
+Listing \ref{lst:Annotative} shows a Delta-Oriented implementation of our object store Product Line.
+The base product here is a simple single-object store.
+On top of that, delta for a multi-object store and an access controlled store are defined as deltas.
+
+\begin{lstlisting}[caption=Annotative example, label={lst:Annotative}]
+core SingleStore {
+	class Store {
+		private Object value;
+		Object read() { return value; }
+		void set(Object nvalue) { value = nvalue; }
+	} 
+}
+
+delta DMultiStore when MultiStore {
+	modifies class Store {
+		remove Object value;
+		add private LinkedList values = new LinkedList();
+
+		remove read();
+		add Object read() { return values.getFirst(); }
+
+		add Object[] readAll() { return values.toArray(); }
+
+		remove void set(Object);
+		add void set(Object nvalue) { values.addFirst(nvalue); }
+	}
+}
+
+delta DAccessControl when AccessControl {
+	modifies class Store {
+		add private boolean sealed = false;
+
+		remove read();
+		add Object read() {
+			if (!sealed) {
+				return Store.read();
+			} else { throw new RuntimeException("Access denied!"); }
+		}
+	}
+}
+\end{lstlisting}
+
 \section{Safety Cases}

### 2017-09-22 doc/vamos18/taxonomy.tex
basic definitions

\documentclass{article}

%% macros
% reference macros
\newcommand{\Classen}{\cite{Classen:2010:MCL:1806799.1806850}}
\begin{document}

\title{A Taxonomy of Software Product Line Analyses}
\maketitle

\abstract
TODO

\section{Introduction}
TODO

\section{Motivation}

\subsection{Compositionaility}

Compositionaility is a property of product-level analyses. An analysis is compositional if it is applied to a component of a system rather than the system as a whole. The granularity of that component varies from one analysis to another. In many programming languages, type checkers work at the level of a compilation unit or a module. Intra-procedural analyses on the other hand work at the level of a single procedure. Non-compositional analyses on the other hand include monolithic model checkers that build a state machine of the whole model with no decomposition into smaller units.

\subsection{Modularity}

In the context of this paper modularity is an attribute of Product Line analyses. By modular, we mean that the Product Line analysis uses the corresponding product-level analysis as a sub-module. This can be done in one of two ways:

A product-level analysis can be used as a black box. In such cases the Product Line analysis calls into the product level analysis and aggregates the results for the whole Product Line.

A product-level analysis can be used as a white box. Some automatic technique is used to lift the implementation of the product-level analysis into a Produt Line analysis. Since this process is automated, we still consider this approach modular.

Non-modular Product Line analyses on the other hand are written from scratch independently of the corresponding product level analysis.

\section{Taxonomy}

\subsection{Compositional and Modular}

\subsection{Compositional and Non-Modular}

\subsection{Non-Compositional and Modular}

\subsection{Non-Compositional and Non-Modular}

\section{Discussion}
TODO

\section{Conclusion}
TODO

\bibliographystyle{ieeetr}
\bibliography{taxonomy} 
\end{document}

### 2017-09-26 doc/SafetyCases/SafetyCasesPL.tex
safety case, questions and introduction

@@
 \maketitle
-%
-%
+
 \section{Introduction}
-%Software systems are everywhere, including safety critical domains.
+Software systems are everywhere, including safety critical domains.
 Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles.
@@
 Safety cases are the most common artifacts resulting from safety analysis.
-%
-%Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs).
+
+Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs).
 Variability is inherent in families of software systems.
 Variability can be due to different deployment environments, differences in hardware devices and peripherals, varying requirements across models of systems, and difference in performance characteristics.
-%
-%Variability in an SPL is usually measured in terms of the number of features in a product line.
+
+Variability in an SPL is usually measured in terms of the number of features in a product line.
 The number of possible products that can be synthesized from an SPL grows exponentially with the number of features.
 As a result, enumerating all possible products and performing safety analysis at the product level is intractable.
-%
-%Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular.
+
+Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular.
 This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.
-%
+
 \subsection{Goals}
-%This document aims at:
-%\begin{itemize}
-%\item Surveying existing techniques
-%\item Identifying gaps and pragmatic requirements that are not fulfilled by existing techniques
-%\item Identifying any underlying research questions
-%\end{itemize}
-%
+This document aims at:
+\begin{itemize}
+\item Introducing terminology for both Software Product Line and Safety Case concepts
+\item Demonstrating the concepts on simple examples
+\item Identifying any underlying pragmatic questions
+\end{itemize}
+
 \subsection{Non-Goals}
-%
-%The following are beyond the scope of this document:
-%
-%\begin{itemize}
-%\item Coming up with new notations or techniques
-%\item Empirical assessment of existing techniques
-%\item Formalizing any of the discussed concepts
-%\end{itemize}
-%
+
+The following are beyond the scope of this document:
+
+\begin{itemize}
+\item Coming up with new notations or techniques
+\item Empirical assessment of existing techniques
+\item Formalizing any of the discussed concepts
+\item Covering any hazard analysis techniques
+\end{itemize}
+
 %\subsection{Running Example}
@@
 
+A safety case is a structured argument demonstrating the safety of a system with respect to a set of identified hazards.
+Hazard analysis and safety case construction are typically done for individual products.
+Variability in SPLs naturally leads to variability in hazards and safety cases.
+However, standard hazard analysis and safety case techniques and notations do not accommodate for variability.
+
 \subsection{Goal Structuring Notation (GSN)}
-%
-%Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:
-%
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=0.75\textwidth]{gsn}
-%  \caption{GSN Elements}
-%\end{figure}
-%
-%\begin{itemize}
-%\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
-%\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
-%\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
-%\end{itemize}
-%
+
+Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{gsn}
+  \caption{GSN Elements}
+\end{figure}
+
+\begin{itemize}
+\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
+\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
+\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
+\end{itemize}
+
 %\subsubsection{Structured Assurance Case Metamodel (SACM)}
@@
 
+As a simple case study, we refer to the Fuel Level Display (FLD) system safety case presented in \cite{Dardar_2014}. This is an automotive embedded system with the following requirements:
+
+\begin{itemize}
+ \item To track the fuel level in different motor vehicle models
+ \item To displaying the fuel level on a gauge visible by the driver
+ \item Optionally, to set an alarm if the fuel level goes below a certain threshold 
+ \end{itemize}
+
+ This case study was extended into a Fuel Level Display Product Line (FLDPL) in \cite{Gallucci_2013}. Variability is introduced in two different dimensions:
+
+ \begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{FLDPLContext}
+  \caption{FLDPL Features Context}
+  \label{fig:FLDPLContext}
+\end{figure}
+
+ \begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{FLDPLFeatures}
+  \caption{FLDPL Functional Features}
+  \label{fig:FLDPLFeatures}
+\end{figure}
+
+ \begin{itemize}
+
+ \item Different vehicle types (trucks and buses) and different fuel types (liquid and gas) as depicted in the feature diagram in figure \ref{fig:FLDPLContext}.
+ \item Based on the combination of the features listed above, a low level fuel warning feature is either enabled or not, as shown in figure \ref{fig:FLDPLFeatures}.
+
+ \end{itemize}
+
 \section{Discussion}
-%
+
 \subsection{Product Line Scenarios}
-%
-%Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios yielding Product Lines.
-%
+
+Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios for building Product Lines.
+
 \subsubsection{Product Evolution over Time}
-%
-%Starting with a single product, variability might start exhibiting itself in lots of different ways.
+
+Starting with a single product, variability might start exhibiting itself in lots of different ways.
 Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability.
 Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem.
-%One way to manage ad-hoc variability is to evolve a single product with variability in a product line.
+
+One way to manage ad-hoc variability is to evolve a single product with variability in a product line.
 This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.
-%
+
 \subsubsection{Product Component Reuse}
-%
-%Again starting with a single component-oriented product, some components might be potentially reused in other products.
+
+Again starting with a single component-oriented product, some components might be potentially reused in other products.
 Sometimes a component can be reused as is, without the need for any modification.
 In other cases the component might require minor modifications, resulting in a slightly different version.
-%
-%This component-level variability can again evolve into a component-level Product Line.
+
+This component-level variability can again evolve into a component-level Product Line.
 With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.
-%
+
 \subsubsection{Product-Line Engineering}
-%
-%The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line.
+
+The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line.
 Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one.
 If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.
-%
+
 \subsection{Product Lines and Safety Cases}
-%
-%Reasoning about Software Product Lines is not straightforward because of several reasons:
-%
-%\begin{itemize}
-%
-%\item The number of possible valid products of a product lines grows exponentially with the number of features.
+
+Reasoning about Software Product Lines is not straightforward because of several reasons:
+
+\begin{itemize}
+
+\item The number of possible valid products of a product lines grows exponentially with the number of features.
 As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.
-%
-%\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations.
+
+\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations.
 When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.
-%
-%\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification).
-%
-%\end{itemize}
-%
+
+\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification).
+
+\end{itemize}
+
 \subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
-%
-%Safety Analysis is one example of reasoning about a software systems.
+
+Safety Analysis is one example of reasoning about a software systems.
 Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products.
 However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons.
-%
-%In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
-%
-%\begin{itemize}
-%
-%\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
-%
-%\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts).
-%
-%\end{itemize}
-%
+
+In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
+
+\begin{itemize}
+
+\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
+
+\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts).
+
+\end{itemize}
+
 \subsection{Product Line evolution impact analysis}
-%
-%As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well.
+
+As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well.
 This is essential to maintain the validity of the safety arguments.
@@
 Two questions are relevant here:
-%
-%\begin{itemize}
-%
-%\item Q2: As an SPL evolves, which pieces of evidence are still valid, and can be reused as is?
-%
-%\item Q3: Within a single piece of evidence, how much reuse can be achieved?
+
+\begin{itemize}
+
+\item Q2: As an SPL evolves, how can we determine which pieces of evidence are still valid, and can be reused as is, as opposed to pieces of evidence that are invalidated by the SPL changes?
+
+\item Q3: Within a single piece of evidence, how much reuse can be achieved?
 For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific?
 How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences?
-%
-%\end{itemize}
-%
+
+\end{itemize}
+
 \bibliographystyle{abbrv}

### 2017-10-04 doc/SafetyCases/SafetyCasesPL.tex
Definitions with FLDPL examples

@@
 
-%\subsection{Running Example}
-%
-%Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}.
+\section{Running Example}
+
+Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}.
 This is a simply product line safety case that extends a single product ISO26262 safety case \cite{Dardar_2014}.
 The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.
-%
-%\section{Existing Techniques}
-%
-\section{Software Product Lines (SPLs)}
 
-\subsection{Basic Concepts}
+The Fuel Level Display (FLD) system is an automotive component implementing two functionalities:
+
+\begin{enumerate}
+\item Displaying the fuel level of a fuel tank on the driver's dashboard
+\item Triggering a warning when the fuel level goes below a certain threshold (optional feature)
+\end{enumerate}
+
+The Fuel Level Display Product Line (FLDPL) captures four variants of FLDs based on the vehicle type and fuel type:
+
+\begin{enumerate}
+\item Truck with liquid fuel
+\item Bus with liquid fuel
+\item Truck with gas
+\item Bus with gas
+\end{enumerate}
+
+section{Software Product Lines (SPLs)}
 
@@
 
+\subsection{Software Product Line}
+
 \begin{description}
@@
 
-For example, let us consider a product line of simple object stores.
-An object store can support the storage of a single object or multiple objects.
-Orthogonally, the store can provide some access control mechanism allowing only certain users to store objects.
-Store capacity (SingleObject vs MultiObject) and access control are just examples of possible variability points in an SPL.
+For our running example, we have four variants of Fuel Level Display systems.
+In addition, each of those variants can come with or without the warning feature.
+This family of product variants has a lot in common, so it is more cost effective to build them as an SPL instead of building each product in isolation.
+
+\subsection{Features}
 
@@
 
-In our object store SPL, SingleObject, MultiObject and access control are the 3 features within this product line.
+At a high level, our FLDPL has two features: fuel level display and low fuel level warning.
 
-An individual product can be generated from an SPL by selecting a set of features for that product.
-Not all product combinations are valid though.
-For example, it does not make sense for an object store to be SingleObject and MultiObject at the same time.
-Those two features are supposed to be mutually exclusive in our product line.
-This kind of constraints are captured by a \emph{Feature Model}:
+Orthogonally, variation with respect to vehicle types and fuel types are also considered feature variants.
+
+So this is the FLDPL list of features:
+
+\begin{enumerate}
+\item FuelLevel
+\item FuelWarning (not present in trucks with gas)
+\item Truck
+\item Bus
+\item LiquidFuel
+\item Gas
+\end{enumerate}
+
+\subsection{Feature Model}
 
@@
 
-For example, a boolean formula expressing our mutual exclusion condition for SingleObject and MultiObject stores, together with an optional AccessControl feature, would be:
+For example, a boolean formula expressing the mutual exclusion condition for the Truck and Bus features:
 
-$(SingleObject \vee MultiObject) \wedge (\neg SingleObject \vee \neg MultiObject) $
+$(Truck \vee Bus) \wedge (\neg Truck \vee \neg Bus) $
+
+\subsection{Feature Diagram}
 
@@
 
+\subsection{Presence Condition}
+
 \begin{description}
@@
 
-For example, a product where we have the MultiObject and AccessControl features can be represented by the following PC: $ MultiObject \wedge AccessControl $
+For example, an FLD product for buses, with liquid fuel, with fuel level display and low fuel warning can be represented by the following PC:
 
-\subsection{SPL Engineering Approaches}
+Bus $\wedge$ Liquid $\wedge$ Estimation $\wedge$ Warning
+
+\section{SPL Engineering Approaches}
 
@@
 
-\subsubsection{Generative}
+\subsection{Generative}
 
@@
 
-\subsubsection{Annotative}
+\subsection{Annotative}
 
@@
 
-\subsubsection{Delta-Oriented}
+\subsection{Delta-Oriented}
 

### 2017-10-07 doc/SafetyCases/SafetyCasesPL.tex
Generative, annotative and delta-oriented pseudo-code

@@
 \item Bus with liquid fuel
-\item Truck with gas
+\item Truck with gas (doesn't include the fuel level warning feature)
 \item Bus with gas
@@
 
-section{Software Product Lines (SPLs)}
+For vehicles with gas, the fuel level is reported directly to the driver.
+However, liquid fuel level in the tank fluctuates due to vehicle movement.
+To smooth the liquid gas level reported to the driver, the level read from the tank sensor is passed into a filter.
+Two types of filters are used:
+
+\begin{enumerate}
+\item a Kalman filter is used in trucks
+\item a Low-Pass filter is used in buses
+\end{enumerate}
+
+\section{Software Product Lines (SPLs)}
 
@@
 \item Bus
-\item LiquidFuel
+\item Liquid
 \item Gas
@@
 \item[Presence Condition]
-A Presence Condition (PC) is a boolean formula over the set of features that does not conflict with the Feature Model.
+A Presence Condition (PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
 \end{description}
 
-For example, an FLD product for buses, with liquid fuel, with fuel level display and low fuel warning can be represented by the following PC:
+For example, the presence condition for the Low Fuel Level Warning feature is:
 
-Bus $\wedge$ Liquid $\wedge$ Estimation $\wedge$ Warning
+$\neg ($ Truck $\wedge$ Gas $)$
 
@@
 SPLs are composed of features, and those features can be represented in various ways:
-
-\subsection{Generative}
-
-In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
-Examples of Generative Product Line approaches include Feature-Oriented Programming \cite{Prehofer1997} and Aspect-Oriented Programming \cite{Kiczales1997}.
-
-\begin{figure}
-  \centering
-    \includegraphics[width=0.75\textwidth]{FeatureOriented}
-  \caption{Feature-Oriented implementation of the object store Product Line}
-  \label{fig:FeatureOriented}
-\end{figure}
-
-Figure \ref{fig:FeatureOriented} shows a possible Feature-Oriented implementation of our object store Product Line.
-\emph{SingleStore} and \emph{MultiStore} are two separate implementations of a their corresponding features, and \emph{AccessControl} is implemented as a cross-cutting feature that refines any of the 2 \emph{Store} implementations.
-
-\begin{figure}
-  \centering
-    \includegraphics[width=0.75\textwidth]{FeatureOrientedGenerated}
-  \caption{The product generated out of the \emph{MultiStore} and \emph{AccessControl} features}
-  \label{fig:FeatureOrientedGenerated}
-\end{figure}
-
-A product can then be generated out of the individual features given a valid Presence Condition.
-For example, if we are to generate a product with the \emph{MultiStore} and \emph{AccessControl} features, the generator would automatically generate the code in figure \ref{fig:FeatureOrientedGenerated}.
-
-\subsection{Annotative}
-
-Generative approaches are only supported by a limited set of programming languages and tools.
-In mainstream languages, Annotative approaches are typically used, mainly because they don't depend on syntactic support for features, or compiler/runtime support for feature composition.
-
-In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (Presence Conditions) are used to annotate different syntactic constructs belonging to different families of products.
-The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
-
-If we are to rewrite our object store Product Line using CPP macros, it would look like listing 
-ref{lst:Annotative}. A single feature (e.g. \emph{ACCESS\_CONTROL} is now scattered across the code base, and in some cases replicated (e.g. lines 13-15 and 24-26).
 
@@
 
+\subsection{Generative}
+
+In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
+Examples of Generative Product Line approaches include Feature-Oriented Programming \cite{Prehofer1997} and Aspect-Oriented Programming \cite{Kiczales1997}.
+
+\begin{lstlisting}[language=Java, caption=Generative example, label={lst:Generative}]
+// Feature module FLDGas
+class FLD {
+	void reportFuelLevel(float level) { 
+		dashboard.display(level); 
+	}
+}
+
+// Feature module FLDLiquidTruck
+class FLD {
+	void reportFuelLevel(float level) { 
+		dashboard.display(filterKalman(level)); 
+	}
+}
+
+// Feature module FLDLiquidBus
+class FLD {
+	void reportFuelLevel(float level) { 
+		dashboard.display(filterLowPass(level)); 
+	}
+}
+
+// Feature module LowGasWarning
+refines class FLD {
+	const float THRESHOLD;
+
+	void reportFuelLevel(float level) { 
+		if (level < THRESHOLD) { dashboard.triggerWarning()); }
+		Super.reportFuelLevel(level);
+	}
+} 
+\end{lstlisting}
+
+Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL.
+\emph{FLDGas}, \emph{FLDLiquidTruck} and \emph{FLDLiquidBus} are three separate implementations of the Fuel Level Display module.
+\emph{LowGasWarning} is implemented as a cross-cutting feature that refines any of the 3 \emph{FLD} implementations.
+
+\begin{lstlisting}[language=Java, caption=Generated Product, label={lst:Generated}]
+class FLD {
+	const float THRESHOLD;
+
+	void reportFuelLevel(float level) { 
+		if (level < THRESHOLD) { dashboard.triggerWarning()); }
+		dashboard.display(filterKalman(level);
+	}
+} 
+\end{lstlisting}
+
+Given the feature expression (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.
+
+\subsection{Annotative}
+
+Generative approaches are only supported by a limited set of programming languages and tools.
+In mainstream languages, Annotative approaches are typically used, mainly because they don't depend on syntactic support for features, or compiler/runtime support for feature composition.
+
+In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (Presence Conditions) are used to annotate different syntactic constructs belonging to different families of products.
+The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
+
+If we are to rewrite FLDPL using CPP macros, it would look like listing 
+ref{lst:Annotative}. A single feature (e.g. \emph{LOW\_GAS\_WARNING} is now scattered across the code base, and in some cases replicated.
+
 \begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
-class Store {
-	#if defined (ACCESS_CONTROL)
-	private boolean sealed = false;
+class FLD {
+	#if defined (LOW_GAS_WARNING)
+	const float THRESHOLD;
 	#endif
 
-	#if defined (SINGLE_STORE)
-	private Object value;
-	Object read() {
-		#if defined (ACCESS_CONTROL)
- 		if (!sealed) {
-		#endif
-			return value;
-		#if defined (ACCESS_CONTROL)
-		} else { throw new RuntimeException("Access denied!"); }
-		#endif 
-	}
-	#elif defined (MULTI_STORE)
-	private LinkedList values = new LinkedList();
-	Object read() { 
-		#if defined (ACCESS_CONTROL)
- 		if (!sealed) {
-		#endif
-			return values.getFirst(); 
-		#if defined (ACCESS_CONTROL)
-		} else { throw new RuntimeException("Access denied!"); }
-		#endif
-	}
-	Object readAll() { return values.toArray(); }
+	void reportFuelLevel(float level) {
+		#if defined (LOW_GAS_WARNING)
+		if (level < THRESHOLD) { dashboard.triggerWarning(); }
 		#endif
 
-	void set(Object nvalue) {
-		#if defined (ACCESS_CONTROL)
- 		if (!sealed) {
+		dashboard.display(
+			#if defined (FLD_GAS)
+			level
 			#endif
-			#if defined (SINGLE_STORE)
-			value = nvalue;
-			#elif defined (MULTI_STORE)
-			values.addFirst(nvalue);
+
+			#if defined (FLD_LIQUID_TRUCK)
+			filterKalman(level);
 			#endif
-		#if defined (ACCESS_CONTROL)
-		} else { throw new RuntimeException("Access denied!"); }
+
+			#if defined (FLD_LIQUID_BUS)
+			filterLowPass(level);
 			#endif
+		);
 	}
@@
 
-Listing \ref{lst:Annotative} shows a Delta-Oriented implementation of our object store Product Line.
-The base product here is a simple single-object store.
-On top of that, delta for a multi-object store and an access controlled store are defined as deltas.
+Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL. The base product here is the Gas FLD. On top of that, deltas for other variants are implemented.
 
-\begin{lstlisting}[caption=Annotative example, label={lst:Annotative}]
-core SingleStore {
-	class Store {
-		private Object value;
-		Object read() { return value; }
-		void set(Object nvalue) { value = nvalue; }
+\begin{lstlisting}[caption=Delta-Oriented example, label={lst:Delta}]
+core FLDGas {
+	class FLD {
+		void reportFuelLevel(float level) { 
+			dashboard.display(level); 
+		}
 	} 
@@
 
-delta DMultiStore when MultiStore {
-	modifies class Store {
-		remove Object value;
-		add private LinkedList values = new LinkedList();
-
-		remove read();
-		add Object read() { return values.getFirst(); }
-
-		add Object[] readAll() { return values.toArray(); }
-
-		remove void set(Object);
-		add void set(Object nvalue) { values.addFirst(nvalue); }
+delta DLiquidTruck when LiquidTruck {
+	modifies class FLD {
+		remove reportFuelLevel;
+		add void reportFuelLevel(float level) { 
+			dashboard.display(filterKalman(level)); 
 		}
@@
 
-delta DAccessControl when AccessControl {
-	modifies class Store {
-		add private boolean sealed = false;
+delta DLiquidBus when LiquidBus {
+	modifies class FLD {
+		remove reportFuelLevel;
+		add void reportFuelLevel(float level) { 
+			dashboard.display(filterLowPass(level)); 
+		}
+}
 
-		remove read();
-		add Object read() {
-			if (!sealed) {
-				return Store.read();
-			} else { throw new RuntimeException("Access denied!"); }
+delta DLowGasWarning when LowGasWarning {
+	modifies class FLD {
+		add const float THRESHOLD;
+
+		remove reportFuelLevel;
+		add void reportFuelLevel(float level) {
+			if (level < THRESHOLD) { dashboard.triggerWarning(); }
+			FLD.reportFuelLevel(level);
 		}

### 2017-10-07 doc/vamos18/taxonomy.tex
motivation paragraph

@@
-\documentclass{article}
+\documentclass[11pt]{article}
 
@@
 % reference macros
-\newcommand{\Classen}{\cite{Classen:2010:MCL:1806799.1806850}}
+%\newcommand{\Classen}{\cite{Classen:2010:MCL:1806799.1806850}}
 \begin{document}
@@
 \section{Motivation}
+Compositionality of analyses directly impacts the degree of variability the analysis needs to deal with.
+Since presence conditions are propositional formulas, propositional reasoning is an essential part of any Product Line analysis.
+The number of propositional variables (which reflects the degree of variability) in formulas is an important factor when it comes to picking the propositional reasoning component used in the analysis.
+For example, with hundreds or thousands of variables, SAT solvers are usually used.
+On the other hand, if the number of variables is much lower (10-20), Binary Decision Digarams (BDDs) tend to perform much better.
+The difference between both reasoning techniques was demonstrated in the difference in performance between Typechef \cite{} and TODO(the other parser) \cite{}.
 
@@
 
-\bibliographystyle{ieeetr}
-\bibliography{taxonomy} 
+%\bibliographystyle{ieeetr}
+%\bibliography{taxonomy} 
 \end{document}

### 2017-10-11 doc/SafetyCases/SafetyCasesPL.tex
safety case example

@@
 \item Introducing terminology for both Software Product Line and Safety Case concepts
-\item Demonstrating the concepts on simple examples
-\item Identifying any underlying pragmatic questions
+\item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
+\item To come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
 \end{itemize}
@@
 
-As a simple case study, we refer to the Fuel Level Display (FLD) system safety case presented in \cite{Dardar_2014}. This is an automotive embedded system with the following requirements:
-
-\begin{itemize}
- \item To track the fuel level in different motor vehicle models
- \item To displaying the fuel level on a gauge visible by the driver
- \item Optionally, to set an alarm if the fuel level goes below a certain threshold 
- \end{itemize}
-
- This case study was extended into a Fuel Level Display Product Line (FLDPL) in \cite{Gallucci_2013}. Variability is introduced in two different dimensions:
+For a specific Fuel Level Display (FLD) product, safety hazards are identified, and then those hazards are structurally addressed in a safety case.
+For example, a snippet of the FLD safety case addressing Kalman filter hazards is presented in figure \ref{fig:SafetyCaseExample}.
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{FLDPLContext}
-  \caption{FLDPL Features Context}
-  \label{fig:FLDPLContext}
+  \includegraphics[width=0.75\textwidth]{SafetyCaseExample}
+  \caption{Kalman filter safety case example}
+  \label{fig:SafetyCaseExample}
 \end{figure}
 
- \begin{figure}
-  \centering
-  \includegraphics[width=0.75\textwidth]{FLDPLFeatures}
-  \caption{FLDPL Functional Features}
-  \label{fig:FLDPLFeatures}
-\end{figure}
-
- \begin{itemize}
-
- \item Different vehicle types (trucks and buses) and different fuel types (liquid and gas) as depicted in the feature diagram in figure \ref{fig:FLDPLContext}.
- \item Based on the combination of the features listed above, a low level fuel warning feature is either enabled or not, as shown in figure \ref{fig:FLDPLFeatures}.
-
- \end{itemize}
-
-\section{Discussion}
-
-\subsection{Product Line Scenarios}
+\section{Product Line Engineering Scenarios}
 
@@
 
-\subsubsection{Product Evolution over Time}
+\subsection{Product Evolution over Time}
 
@@
 
-\subsubsection{Product Component Reuse}
+\subsection{Product Component Reuse}
 
@@
 
-\subsubsection{Product-Line Engineering}
+\subsection{Explicit Product-Line Engineering}
 
@@
 
-\subsection{Product Lines and Safety Cases}
-
-Reasoning about Software Product Lines is not straightforward because of several reasons:
-
-\begin{itemize}
-
-\item The number of possible valid products of a product lines grows exponentially with the number of features.
+%\subsection{Product Lines and Safety Cases}
+%
+%Reasoning about Software Product Lines is not straightforward because of several reasons:
+%
+%\begin{itemize}
+%
+%\item The number of possible valid products of a product lines grows exponentially with the number of features.
 As a result, enumerating all possible valid products becomes intractable, and thus existing product-level reasoning techniques don't directly scale to product lines.
-
-\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations.
+%
+%\item A Feature Model is an integral part of a Product Line, and its role is to define the valid set of feature combinations.
 When reasoning about a Product Line, the Feature Model has to be taken into consideration to avoid including invalid products in any argument.
-
-\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification).
-
-\end{itemize}
-
-\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
-
-Safety Analysis is one example of reasoning about a software systems.
+%
+%\item Whenever an argument is made regarding a Product Line, it is essential to specify whether this argument applies to all valid products (universal quantification over products), or only to some (existential quantification).
+%
+%\end{itemize}
+%
+%\subsection{Q1: Product-Line of Safety Cases or a Safety Case of a Product Line}
+%
+%Safety Analysis is one example of reasoning about a software systems.
 Techniques like Hazard Analysis, Fault Tree Analysis and Safety Cases have been well studied and applied to software products.
 However, when it comes to Product Lines, those techniques do not directly scale due to the previously mentioned reasons.
-
-In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
-
-\begin{itemize}
-
-\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
-
-\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts).
-
-\end{itemize}
-
-\subsection{Product Line evolution impact analysis}
-
-As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well.
+%
+%In addition, one fundamental question pertains to the granularity to which Safety Analysis is performed. We can think of 2 different approaches here:
+%
+%\begin{itemize}
+%
+%\item A Safety Case of a Product Line: here safety analysis as a process is applied to a Software Product Line, resulting in a single Safety Case covering the whole Product Line.
+%
+%\item A Product Line of Safety Cases: since a Safety Case itself is a Software Artifact (albeit a semi-formal one at best), we can have a Product Line of Safety Cases (pretty much like Product Lines of Use Case Diagrams, Class Diagrams or Source Code artifacts).
+%
+%\end{itemize}
+%
+%\subsection{Product Line evolution impact analysis}
+%
+%As an SPL evolves over time (with changing requirements or more variability), existing safety cases need to evolve as well.
 This is essential to maintain the validity of the safety arguments.
@@
 Two questions are relevant here:
-
-\begin{itemize}
-
-\item Q2: As an SPL evolves, how can we determine which pieces of evidence are still valid, and can be reused as is, as opposed to pieces of evidence that are invalidated by the SPL changes?
-
-\item Q3: Within a single piece of evidence, how much reuse can be achieved?
+%
+%\begin{itemize}
+%
+%\item Q2: As an SPL evolves, how can we determine which pieces of evidence are still valid, and can be reused as is, as opposed to pieces of evidence that are invalidated by the SPL changes?
+%
+%\item Q3: Within a single piece of evidence, how much reuse can be achieved?
 For example, if the results of a test suite and considered a piece of evidence, how much of the test cases are common across variants and how much are variant-specific?
 How easy is it to draw a correspondence between features, goals/sub-goals and evidences/sub-evidences?
-
-\end{itemize}
+%
+%\end{itemize}
 

### 2017-10-17 doc/SafetyCases/SafetyCasesPL.tex
minor document updates

@@
 
-\title{Safety Cases for Software Product Lines}
-%\author{The Author}
-\date{}                                           % Activate to display a given date or no date
+\title{Safety Cases for Software Product Lines: Terminology and Example}
+\author{Ramy Shahin \\ University of Toronto \\ rshahin@cs.toronto.edu \and
+Marsha Chechik \\ University of Toronto \\ chechik@cs.toronto.edu}
+\date{\today}                                           % Activate to display a given date or no date
 
@@
 \section{Introduction}
-Software systems are everywhere, including safety critical domains.
+%Software systems are everywhere, including safety critical domains.
 Software is inherently embedded into medical devices, power plant controllers, airplane controllers, drones, and soon self-driving vehicles.
@@
 
-Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs).
+%Orthogonally, most of the aforementioned examples of safety critical software systems are engineered as Software Product Lines (SPLs).
 Variability is inherent in families of software systems.
@@
 
-Variability in an SPL is usually measured in terms of the number of features in a product line.
+%Variability in an SPL is usually measured in terms of the number of features in a product line.
 The number of possible products that can be synthesized from an SPL grows exponentially with the number of features.
@@
 
-Several attempts have been made to leverage the high degree of commonality across products of an SPL to minimize the cost of product line analysis in general, and safety analysis in particular.
+Several attempts have been made to leverage the high degree of commonality across products of a Software Product Line (SPL) to minimize the cost of product line analysis in general, and safety analysis in particular.
 This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.
@@
 \item Demonstrating the Product Line and Safety Case concepts on a simple exemplar
-\item To come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
+\item To try to come to a mutual agreement on concepts, terminology and the Product Line safety engineering at GM
 \end{itemize}
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item Displaying the fuel level of a fuel tank on the driver's dashboard
 \item Triggering a warning when the fuel level goes below a certain threshold (optional feature)
-\end{enumerate}
+\end{itemize}
 
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item Truck with liquid fuel
@@
 \item Bus with gas
-\end{enumerate}
+\end{itemize}
 
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item a Kalman filter is used in trucks
 \item a Low-Pass filter is used in buses
-\end{enumerate}
+\end{itemize}
 
@@
 
-Variability is commonplace in software systems.
+%Variability is commonplace in software systems.
 Managing variability across different software artifacts, and throughout the Software Development Life Cycle (SDLC) is always a challenge though.
 Software Product Line Engineering draws an analogy to industrial product lines, where different variants of a product are manufactured using a common set of parts and machinery.
+
 The goal of SPLs is to maximize reuse across different variants of a software product, while at the same time simplifying the extensibility and maintainability of the product line.
@@
 In addition, each of those variants can come with or without the warning feature.
-This family of product variants has a lot in common, so it is more cost effective to build them as an SPL instead of building each product in isolation.
+This family of product variants has a lot in common, so it makes sense to build them as an SPL instead of building each product in isolation.
 
@@
 
-Orthogonally, variation with respect to vehicle types and fuel types are also considered feature variants.
+Orthogonally, there is variation with respect to vehicle types and fuel types.
 
-So this is the FLDPL list of features:
+The FLDPL list of features is:
 
@@
 
-$(Truck \vee Bus) \wedge (\neg Truck \vee \neg Bus) $
+(Truck $\vee$ Bus) $\wedge$ ($\neg$ Truck $\vee \neg$ Bus)
 
@@
     \includegraphics[width=0.75\textwidth]{FeatureDiagram}
-  \caption{Feature Diagram Example}
+  \caption{Feature Diagram of the FLDPL}
   \label{fig:FeatureDiagram}
@@
 
-For example, figure \ref{fig:FeatureDiagram} shows a feature diagram for our SPL.
+For example, figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL.
+In this diagram, the usage context is composed of two mandatory features: Vehicle type and Fuel type.
+Vehicle type can be either Truck or Bus (mutually exclusive options).
+Similarly, Fuel type can be either Liquid or Gas.
+In addition, the FLD system has one mandatory feature (Fuel Level Estimation) and one optional feature (Low Fuel Level Warning).
+An annotation (O1) is added to the diagram to specify the set of products where the low fuel level warning feature is present.
 
@@
 $\neg ($ Truck $\wedge$ Gas $)$
+
+\subsection{Product Configuration}
+
+\begin{description}
+
+\item[Product Configuration]
+A product configuration is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
+\end{description}
+
+For example, the configuration of an FLD for a Truck using Gas as fuel would be (Truck $\wedge$ Gas $\wedge$ FuelLevel).
+
+\subsection{150\% Representation}
+
+\begin{description}
+
+\item[150\% Representation]
+The 150\% representation of an SPL is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
+\end{description}
+
+Since in the vast majority of cases all those syntactic constructs can not all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.
+
+Examples of 150\% representations can be found in section 4.
 
@@
 
-\begin{lstlisting}[language=Java, caption=Generative example, label={lst:Generative}]
+\begin{lstlisting}[language=Java, caption=Generative FLDPL, label={lst:Generative}, float]
 // Feature module FLDGas
@@
 Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL.
-\emph{FLDGas}, \emph{FLDLiquidTruck} and \emph{FLDLiquidBus} are three separate implementations of the Fuel Level Display module.
-\emph{LowGasWarning} is implemented as a cross-cutting feature that refines any of the 3 \emph{FLD} implementations.
+FLDGas, FLDLiquidTruck and FLDLiquidBus are three separate implementations of the Fuel Level Display module.
+LowGasWarning is implemented as an orthogonal feature that refines any of the 3 FLD implementations.
 
-\begin{lstlisting}[language=Java, caption=Generated Product, label={lst:Generated}]
+\begin{lstlisting}[language=Java, caption=Generated FLD Product, label={lst:Generated}, float]
 class FLD {
@@
 
-Given the feature expression (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.
+Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.
 
@@
 
-If we are to rewrite FLDPL using CPP macros, it would look like listing 
-ref{lst:Annotative}. A single feature (e.g. \emph{LOW\_GAS\_WARNING} is now scattered across the code base, and in some cases replicated.
+If we are to rewrite FLDPL using CPP macros, it would look like listing \ref{lst:Annotative}.
+A single feature implementation (e.g.
+LOW\_GAS\_WARNING is now scattered across the code base, and in some cases replicated.
+This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.
 
-\begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
+\begin{lstlisting}[language=Java, caption=Annotative FLDPL, label={lst:Annotative}, float]
 class FLD {
@@
 
-\begin{lstlisting}[caption=Delta-Oriented example, label={lst:Delta}]
+\begin{lstlisting}[caption=Delta-Oriented FLDPL, label={lst:Delta}, float]
 core FLDGas {
@@
 
-Goal Structuring Notation is used to layout logical arguments connecting safety goals to pieces of evidence satisfying them. A safety case in GSN is typically composed of 3 kinds of artifacts:
+Goal Structuring Notation \cite{Kelly} can be used to layout logical arguments connecting safety goals to pieces of evidence satisfying them.
+As seen in figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:
 
@@
   \caption{GSN Elements}
+  \label{fig:gsn}
 \end{figure}
@@
 
+Here the robustness of the estimation algorithm in the context of using the Kalman filter is a goal (G\_6).
+This goal is refined into a more specific goal: the steadiness of the output of the Kalman filter (G\_7).
+This goal is further refined into 3 sub-goals (G\_8, G\_9, G\_10).
+Goal G\_8 s fulfilled by the results of standard deviation analysis (evidence E\_1).
+In addition, goals G\_9 and G\_10 are fulfilled by simulation results (evidence E\_2).
+
  \begin{figure}
@@
 
-\section{Product Line Engineering Scenarios}
+\section{Questions}
 
-Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios for building Product Lines.
+To make sure all parties agree on the context of this project, and that research work is aligned with GM practical problems, we have a few questions:
 
-\subsection{Product Evolution over Time}
+\begin{enumerate}
 
-Starting with a single product, variability might start exhibiting itself in lots of different ways.
+\item Is the example used in this document (FLDPL) representative of the SPLs developed at GM?
+
+\item What is the product line engineering approach used by GM engineers (annotative, generative or delta-oriented)?
+
+\item What is the safety case notation used by GM safety engineers?
+
+\item How is safety case variability currently maintained by GM safety engineers?
+
+\item How are hazards, safety case goals, pieces of evidence managed across different products in a product line by GM safety engineers?
+
+\end{enumerate}
+
+%\section{Product Line Engineering Scenarios}
+%
+%Software Product Lines can come into existence in several ways. This section briefly outline engineering scenarios for building Product Lines.
+%
+%\subsection{Product Evolution over Time}
+%
+%Starting with a single product, variability might start exhibiting itself in lots of different ways.
 Different deployment environments, different modes of operations, minor requirement variations from different stakeholders, and varying workloads are all examples of product-level variability.
 Over time, the maintenance of software artifacts with too much ad-hoc variability becomes a real problem.
-
-One way to manage ad-hoc variability is to evolve a single product with variability in a product line.
+%
+%One way to manage ad-hoc variability is to evolve a single product with variability in a product line.
 This process usually starts with factoring out a common core (common to all variants), and variant-specific artifacts.
-
-\subsection{Product Component Reuse}
-
-Again starting with a single component-oriented product, some components might be potentially reused in other products.
+%
+%\subsection{Product Component Reuse}
+%
+%Again starting with a single component-oriented product, some components might be potentially reused in other products.
 Sometimes a component can be reused as is, without the need for any modification.
 In other cases the component might require minor modifications, resulting in a slightly different version.
-
-This component-level variability can again evolve into a component-level Product Line.
+%
+%This component-level variability can again evolve into a component-level Product Line.
 With more and more components being reused in this fashion, a component Product Line library becomes the skeleton of a Software Product Line.
-
-\subsection{Explicit Product-Line Engineering}
-
-The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line.
+%
+%\subsection{Explicit Product-Line Engineering}
+%
+%The previously outlined scenarios start with a single product, and with more and more variability introduced the product evolves into a product line.
 Sometimes on the other hand the designers of a system are aware from day one that they are building a family or products, not just one.
 If at design time they draw a clear boundary between a common core and a set of varying artifacts, they are actually engineering a Product Line.
-
+%
 %\subsection{Product Lines and Safety Cases}

### 2017-10-19 doc/SafetyCases/SafetyCasesPL.tex
minor document updates

@@
 Several attempts have been made to leverage the high degree of commonality across products of a Software Product Line (SPL) to minimize the cost of product line analysis in general, and safety analysis in particular.
-This document explores previous attempts and fundamental questions at the crossroads of Safety Analysis and Software Product Lines.
+The goals of this document are:
 
-\subsection{Goals}
-This document aims at:
 \begin{itemize}
@@
 \end{itemize}
-
-\subsection{Non-Goals}
 
@@
 Throughout this document we use a Fuel Level Display Product Line (FLDPL) example from \cite{Gallucci_2013}.
-This is a simply product line safety case that extends a single product ISO26262 safety case \cite{Dardar_2014}.
 The diagrams in this document also come from \cite{Gallucci_2013} unless otherwise specified.
@@
 
-We start with some basic definitions, together with some simple examples illustrating them. The definitions and the object store example presented here are adapted from those in \cite{Thum}.
+We start with some basic definitions, together with some simple examples illustrating them.
 
@@
 \item[Software Product Line]
-A Software Product Line (SPL) is a set of similar software products built from a common set of artifacts (e.g. design/specification models, code constructs, test cases, etc...).
+(SPL) is a set of similar software products built from a common set of artifacts (e.g., design/specification models, code constructs, test cases, etc...).
 
@@
 \item[Feature]
-A feature is an externally visible property, aspect or quality of a software system.
+is an externally visible property, aspect or quality of a software system.
 
@@
 \item[Feature Diagram]
-a hierarchical graphical representation of a Feature Model, where each non-root feature depends on its parent.
+is a hierarchical graphical representation of a Feature Model, where each non-root feature depends on its parent.
 
@@
 
-For example, figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL.
+For example, Figure \ref{fig:FeatureDiagram} shows a feature diagram of the FLDPL.
 In this diagram, the usage context is composed of two mandatory features: Vehicle type and Fuel type.
@@
 \item[Presence Condition]
-A Presence Condition (PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
+(PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
 \end{description}
@@
 \item[Product Configuration]
-A product configuration is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
+is a boolean formula over the set of features of an SPL that specifies the set of features present in a product.
 \end{description}
@@
 \item[150\% Representation]
-The 150\% representation of an SPL is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
+is a syntactic representation with syntactic constructs belonging to all the features of the SPL. 
 \end{description}
 
-Since in the vast majority of cases all those syntactic constructs can not all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.
+Since in the vast majority of cases all those syntactic constructs cannot all coexist in a product (given feature model constraints), this representation is referred to as 150\%, as opposed to the 100\% representation of a single product.
 
-Examples of 150\% representations can be found in section 4.
+Examples of 150\% representations can be found in Section 4.
 
@@
      captionpos=b, 
-	frame=shadowbox
+     frame=shadowbox,
+     language=Java,
+     morekeywords={if, endif, defined, refines, core, delta, modifies, remove, add, when}
 }
@@
 
-In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
-Examples of Generative Product Line approaches include Feature-Oriented Programming \cite{Prehofer1997} and Aspect-Oriented Programming \cite{Kiczales1997}.
-
-\begin{lstlisting}[language=Java, caption=Generative FLDPL, label={lst:Generative}, float]
+\begin{lstlisting}[caption=Generative FLDPL, label={lst:Generative}, float]
 // Feature module FLDGas
@@
 
-Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL.
-FLDGas, FLDLiquidTruck and FLDLiquidBus are three separate implementations of the Fuel Level Display module.
-LowGasWarning is implemented as an orthogonal feature that refines any of the 3 FLD implementations.
-
-\begin{lstlisting}[language=Java, caption=Generated FLD Product, label={lst:Generated}, float]
+\begin{lstlisting}[caption=Generated FLD Product, label={lst:Generated}, float]
 class FLD {
@@
 
-Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.
+In Generative Product Lines, each feature is designed and implemented separately, and then given a product configuration, a product is automatically generated out of the separate feature artifacts.
+Examples of Generative Product Line approaches include Feature-Oriented Programming \cite{Prehofer1997} and Aspect-Oriented Programming \cite{Kiczales1997}.
+
+Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL.
+FLDGas, FLDLiquidTruck and FLDLiquidBus are three separate implementations of the Fuel Level Display module.
+LowGasWarning is implemented as an orthogonal feature that refines any of the 3 FLD implementations.
+
+Given the product configuration (Liquid $\wedge$ Truck $\wedge$ FuelWarning), a code generator would automatically produce the product in Listing \ref{lst:Generated}.
 
@@
 
-Generative approaches are only supported by a limited set of programming languages and tools.
-In mainstream languages, Annotative approaches are typically used, mainly because they don't depend on syntactic support for features, or compiler/runtime support for feature composition.
-
-In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (Presence Conditions) are used to annotate different syntactic constructs belonging to different families of products.
-The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
-
-If we are to rewrite FLDPL using CPP macros, it would look like listing \ref{lst:Annotative}.
-A single feature implementation (e.g.
-LOW\_GAS\_WARNING is now scattered across the code base, and in some cases replicated.
-This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.
-
-\begin{lstlisting}[language=Java, caption=Annotative FLDPL, label={lst:Annotative}, float]
+\begin{lstlisting}[caption=Annotative FLDPL, label={lst:Annotative}, float]
 class FLD {
@@
 
+Generative approaches are only supported by a limited set of programming languages and tools.
+In mainstream languages, Annotative approaches are typically used, mainly because they don't depend on syntactic support for features, or compiler/runtime support for feature composition.
+
+In Annotative Product Lines, a single set of artifacts is maintained, and feature expressions (Presence Conditions) are used to annotate different syntactic constructs belonging to different families of products.
+The most commonly used Annotative technique is C Pre-Processor (CPP) macro annotations surrounding fragments of source code.
+
+If we are to rewrite FLDPL using CPP macros, it would look like Listing \ref{lst:Annotative}.
+A single feature implementation (e.g., LOW\_GAS\_WARNING) is now scattered across the code base (lines 3 and 8).
+This code snippet is an example of a 150\% representation, where a product configuration (in terms of macro definitions) excludes some of the syntactic constructs, leaving only those belonging to a specific product.
+
 \subsection{Delta-Oriented}
-
-In Delta-Oriented \cite{Schaefer} Product Lines, a base product (referred to as the Core Module) is fully defined, and variants (deltas) are defined in terms of their differences with respect to the base product.
-
-Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL. The base product here is the Gas FLD. On top of that, deltas for other variants are implemented.
 
@@
 
+In Delta-Oriented \cite{Schaefer} Product Lines, a base product (referred to as the Core Module) is fully defined, and variants (deltas) are defined in terms of their differences with respect to the base product.
+
+Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL.
+The base product here is the Gas FLD (lines 1-7).
+On top of that, deltas for other variants are implemented.
+For example, the DLiquidTruck delta (lines 9-15) modifies the FLD class implemented in the base product by removing the reportFuelLevel method and adding another one that passes the fuel level through a Kalman filter.
+Similarly, DLiquidBus (lines 17-23) and DLowGasWarning (lines 25-35) are deltas for their corresponding features.
+
 \section{Safety Cases}
@@
 Goal Structuring Notation \cite{Kelly} can be used to layout logical arguments connecting safety goals to pieces of evidence satisfying them.
-As seen in figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:
+As seen in Figure \ref{fig:gsn}, a safety case in GSN is typically composed of 3 kinds of artifacts:
 
@@
 \begin{itemize}
-\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques
-\item Solutions(Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports
-\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences
+\item Goals: These are the safety goals (requirements) identified as a result of hazard analysis techniques.
+\item Solutions (Evidences): These are pieces of evidence that a risk/hazard is mitigated. Evidences can be test case results, formal verification outputs, or manual investigation reports.
+\item Strategies: a strategy provides the logical argument connecting safety goals to their corresponding mitigating evidences.
 \end{itemize}
@@
 %SACM \cite{SACM} is an Object Management Group (OMG) standard.
-SACM integrates well with other OMG metamodels (e.g.
-UML).
+SACM integrates well with other OMG metamodels (e.g., UML).
 It provides a much richer notation compared to GSN.
@@
 For a specific Fuel Level Display (FLD) product, safety hazards are identified, and then those hazards are structurally addressed in a safety case.
-For example, a snippet of the FLD safety case addressing Kalman filter hazards is presented in figure \ref{fig:SafetyCaseExample}.
+For example, a snippet of the FLD safety case addressing Kalman filter hazards is presented in Figure \ref{fig:SafetyCaseExample}.
 
@@
 This goal is further refined into 3 sub-goals (G\_8, G\_9, G\_10).
-Goal G\_8 s fulfilled by the results of standard deviation analysis (evidence E\_1).
-In addition, goals G\_9 and G\_10 are fulfilled by simulation results (evidence E\_2).
+Results of standard deviation analysis (evidence E\_1) address Goal G\_8.
+In addition,  simulation results (evidence E\_2) address goals G\_9 and G\_10.
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{SafetyCaseExample}
+  \includegraphics[width=1.0\textwidth]{SafetyCaseExample}
   \caption{Kalman filter safety case example}
@@
 
-\item What is the safety case notation used by GM safety engineers?
+\item What are the safety case constructs used by GM safety engineers? Is there a standard notation (e.g., GSN) used?
 
-\item How is safety case variability currently maintained by GM safety engineers?
+\item How is safety case variability currently managed by GM safety engineers?
 
-\item How are hazards, safety case goals, pieces of evidence managed across different products in a product line by GM safety engineers?
+\item How are hazards, safety case goals and pieces of evidence managed across different products in a product line by GM safety engineers?
+
+\item How is variability traced throughout the end-to-end Software Lifecycle at GM? In particular, how does safety case variability map to variability in design and code artifacts?
 

### 2017-11-10 doc/vamos18/taxonomy.tex
taxonomy doc minor changes

@@
-\documentclass[11pt]{article}
+\documentclass[sigplan]{acmart}
 
@@
 \title{A Taxonomy of Software Product Line Analyses}
+
+\author{Ramy Shahin}
+\affiliation{%
+  \institution{University of Toronto}
+}
+\email{ramy.shahin@cs.toronto.edu}
+
+\author{Marsha Chechik}
+\affiliation{%
+  \institution{University of Toronto}
+}
+\email{marsha.chechik@cs.toronto.edu}
+
+\begin{abstract}
+TODO
+\end{abstract}
+
 \maketitle
-
-\abstract
-TODO
 
@@
 \section{Motivation}
-Compositionality of analyses directly impacts the degree of variability the analysis needs to deal with.
+Product line analyses have been surveyed and classified according to different criteria (e.g., \cite{Thum}). 
+Modularity of analyses directly impacts the degree of variability the analysis needs to deal with.
 Since presence conditions are propositional formulas, propositional reasoning is an essential part of any Product Line analysis.
@@
 On the other hand, if the number of variables is much lower (10-20), Binary Decision Digarams (BDDs) tend to perform much better.
-The difference between both reasoning techniques was demonstrated in the difference in performance between Typechef \cite{} and TODO(the other parser) \cite{}.
+The difference between both reasoning techniques was demonstrated in the difference in performance between Typechef \cite{Kastner} and SuperC \cite{Gazzillo}.
 
-\subsection{Compositionaility}
+\subsection{Modularity}
 
-Compositionaility is a property of product-level analyses.
-An analysis is compositional if it is applied to a component of a system rather than the system as a whole.
+Modularity is a property of product-level analyses. An analysis can be either modular or monolithic. 
+\begin{description}
+\item [Modular Analysis]
+is applied to units (components) of a system rather than the system as a whole. The results of unit-by-unit analysis are then aggregated to provide the analysis result for the whole system.
+\end{description}
+
 The granularity of that component varies from one analysis to another.
@@
 Intra-procedural analyses on the other hand work at the level of a single procedure.
-Non-compositional analyses on the other hand include monolithic model checkers that build a state machine of the whole model with no decomposition into smaller units.
 
-\subsection{Modularity}
+\begin{description}
+\item [Monolithic Analysis]
+is applied to the system as a whole. 
+\end{description}
 
-In the context of this paper modularity is an attribute of Product Line analyses.
-By modular, we mean that the Product Line analysis uses the corresponding product-level analysis as a sub-module.
-This can be done in one of two ways:
+Examples of monolithic analyses include monolithic model checkers that build a state machine of the whole model with no decomposition into smaller units.
 
-A product-level analysis can be used as a black box. In such cases the Product Line analysis calls into the product level analysis and aggregates the results for the whole Product Line.
+\subsection{Derivability}
 
-A product-level analysis can be used as a white box.
+In the context of this paper derivability is an attribute of Product Line analyses.
+
+\begin{description}
+\item [Derived Analysis]
+is a  Product Line analysis that is built on top of its corresponding product-level analysis. 
+\end{description}
+
+An analysis can be derived in one of two ways:
+
+\begin{itemize}
+
+\item A product-level analysis can be used as a black box. In such cases the Product Line analysis calls into the product level analysis and aggregates the results for the whole Product Line.
+
+\item A product-level analysis can be used as a white box.
 Some automatic technique is used to lift the implementation of the product-level analysis into a Produt Line analysis.
 Since this process is automated, we still consider this approach modular.
+\end{itemize}
 
-Non-modular Product Line analyses on the other hand are written from scratch independently of the corresponding product level analysis.
+\begin{description}
+
+\item [Rewritten Analysis]
+is a Product Line analysis written from scratch independently of its corresponding product level analysis.
+\end{description}
 
@@
 
-\subsection{Compositional and Modular}
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{| l | l | l |}
+\hline
+	& Modular & Monolithic \\ \hline
+Derived & Hello & World \\ \hline
+Rewritten & Hello & World \\ \hline
+\end{tabular}
+\end{center}
+\caption{A taxonomy of Software Product Line analyses}
+\label{table:taxonomy}
+\end{table}
 
-\subsection{Compositional and Non-Modular}
+\subsection{Modular and Derived}
 
-\subsection{Non-Compositional and Modular}
+\subsection{Modular and Rewritten}
 
-\subsection{Non-Compositional and Non-Modular}
+\subsection{Monolithic and Derived}
+
+\subsection{Monolithic and Rewritten}
 
@@
 
-%\bibliographystyle{ieeetr}
-%\bibliography{taxonomy} 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{taxonomy} 
 \end{document}

### 2017-12-09 doc/quals/quals.tex
Quals paper outline

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage[T1]{fontenc}

\usepackage[utf8]{inputenc}

\usepackage{setspace}
\doublespacing

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Software Product Line Analysis Techniques}
\author{Ramy Shahin \\ rshahin@cs.toronto.edu}
\date{\today}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\begin{abstract}
TODO
\end{abstract}

\section{Introduction}
TODO

\section{Featured Transition Systems}
\cite{Classen:2013}

\section{SPLLift}
\cite{Bodden:2013}

\section{Scalable Analysis of Variable Software}
\cite{Liebig:2013}

\section{Product Line Type Checking}

In \cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq\cite{Bertot:2010} proof assistant. Featherweight Java (FJ) is a simplistic subset of Java, leaving out constructs such as assignment (mutation) and interfaces. CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE. This type checker is further extended to fully cover Java constructs.

A product line is considered well-typed only if all the products that can be generated out of it are well-typed (typing is preserved by product generation). The type checker here takes the source code of a product line as input, and checks all variants for type errors. In addition to typing preservation, a second goal is backward compatibility. No new syntactic constructs were introduced, which allowed the use of existing tools.

One fundamental limitation of this work is that it severely restricts the allowed forms of variability. First, only disciplined annotations are supported. Although this does not restrict the expressivity of variability, it limits the applicability of this type checker to existing product lines that use undisciplined annotations. More importantly though, variability annotation is only supported within elements of syntactic sequences. For example, a sequence of methods (potentially empty) is defined. Each element (method) within that sequence can be annotated individually. However, variability is not supported for singleton syntactic constructs. This means that we cannot vary the subtype of a class or the return type of a method across products.

This restricted form of variability serves two purposes. First it fulfills the backward compatibility goal, because the 150\% representation (without the preprocessor macros) is still syntacticly correct. Moreover, it turns the variability type checking problem into reachability of definitions. For example, a method definition annotated by a presence condition has to be reachable by all the method calls across all product variants.

The type checker has been evaluated on 5 Java product lines, ranging from 9 to 42 features. Some type errors were found by the checker in some product variants. The performance (in terms of time taken to type check the product line) was evaluated compared to approximate per-product performance, and on these 5 product lines processing time grew much slower than the number of variants. It is worth noting though that these product lines are relatively small, both in terms of number of lines of code and number of features.

\section{SuperC}
\cite{Gazzillo:2012}

\section{Variability-Aware Abstract Interpretation}
\cite{Midtgaard:2015}

\section{Strategies for Product-line Verification}
\cite{Apel:2013}

\bibliographystyle{abbrv}
\bibliography{quals} 
\end{document}

### 2017-12-09 doc/SafetyCases/SafetyCasesPL.tex
safety cases presentation

@@
 \usepackage{color}
+\usepackage[dvipsnames]{xcolor}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
@@
      stringstyle=\color{red}\ttfamily,
-     commentstyle=\color{green}\ttfamily,
+     commentstyle=\color{Gray}\ttfamily,
      morecomment=[l][\color{magenta}]{\#},

### 2017-12-09 doc/vamos18/taxonomy.tex
taxonomy outline

@@
-\documentclass[sigplan]{acmart}
+\documentclass[11pt, english]{article}
+\usepackage{babel}
 
@@
 
-\author{Ramy Shahin}
-\affiliation{%
-  \institution{University of Toronto}
-}
-\email{ramy.shahin@cs.toronto.edu}
+\author{Ramy Shahin \\
+  University of Toronto \\
+ ramy.shahin@cs.toronto.edu
+ \and
+  Marsha Chechik \\
+  University of Toronto \\
+ marsha.chechik@cs.toronto.edu}
 
-\author{Marsha Chechik}
-\affiliation{%
-  \institution{University of Toronto}
-}
-\email{marsha.chechik@cs.toronto.edu}
+\maketitle
 
@@
 
-\maketitle
-
 \section{Introduction}
@@
 \section{Motivation}
-Product line analyses have been surveyed and classified according to different criteria (e.g., \cite{Thum}). 
+Product line analyses have been surveyed and classified according to different criteria (e.g., \cite{Thum:2014}). 
 Modularity of analyses directly impacts the degree of variability the analysis needs to deal with.
@@
 For example, with hundreds or thousands of variables, SAT solvers are usually used.
-On the other hand, if the number of variables is much lower (10-20), Binary Decision Digarams (BDDs) tend to perform much better.
-The difference between both reasoning techniques was demonstrated in the difference in performance between Typechef \cite{Kastner} and SuperC \cite{Gazzillo}.
+On the other hand, if the number of variables is much lower (10-20), Binary Decision Diagrams (BDDs) tend to perform much better.
+The difference between both reasoning techniques was demonstrated in the difference in performance between Typechef \cite{Kastner:2011} and SuperC \cite{Gazzillo:2012}.
 
@@
 \item A product-level analysis can be used as a white box.
-Some automatic technique is used to lift the implementation of the product-level analysis into a Produt Line analysis.
+Some automatic technique is used to lift the implementation of the product-level analysis into a Product Line analysis.
 Since this process is automated, we still consider this approach modular.
@@
 
-\begin{table}[t]
+\begin{table*}[t]
 \begin{center}
@@
 	& Modular & Monolithic \\ \hline
-Derived & Hello & World \\ \hline
-Rewritten & Hello & World \\ \hline
+	Derived & 
+		SPL\textsuperscript{LIFT} \cite{Bodden:2013} & 
+		FTS \cite{Classen:2010} \\ \hline
+	Rewritten & 
+		\begin{tabular}{@{}c@{}}
+			TypeChef \cite{Kastner:2011} \\ 
+			SuperC \cite{Gazzillo:2012}
+		\end{tabular} &
+			Model Transformations \cite{Salay:2014} \\ \hline
 \end{tabular}
@@
 \label{table:taxonomy}
-\end{table}
+\end{table*}
 
@@
 
+\subsubsection{SPL\textsuperscript{LIFT}}
+
 \subsection{Modular and Rewritten}
+
+\subsubsection{Feature Transition Systems}
 
@@
 
+\subsubsection{TypeChef}
+
+\subsubsection{SuperC}
+
 \subsection{Monolithic and Rewritten}
+
+\subsubsection{Model Transformations}
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{abbrv}
 \bibliography{taxonomy} 

### 2017-12-24 doc/quals/quals.tex
update

@@
 \author{Ramy Shahin \\ rshahin@cs.toronto.edu}
-\date{\today}                                           % Activate to display a given date or no date
+\date{December 2, 2017}                                           % Activate to display a given date or no date
 
@@
 
+In \cite{Classen:2013}, Labeled Transition Systems (LTSs) are extended with feature expressions into Featured Transition Systems (FTSs).
+FTSs can be used to encode variability-aware models, which can be checked by the SNIP model checker.
+
+The paper provides a FTS-based formalism, together with model checking algorithms, complexity analysis (with proofs) and the featured LTL (fLTL) logic.
+Several state space reduction optimizations are also explained.
+The empirical evaluation highlights the speedup due to the optimized algorithm.
+
+One shortcoming of this paper is that empirical evaluation is done on only one model (the mine pump controller model).
+This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties.
+With only one model (moreover, a small one), the evaluation results cannot be safely generalized.
+
 \section{SPLLift}
 \cite{Bodden:2013}
+
+In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS framework) is extended to annotative product lines.
+In IFDS, the data flow analysis problem is encoded as a graph reachability problem.
+Variability due to feature expressions is encod
 

### 2018-01-02 doc/quals/quals.tex
Apel's ICSE'13 paper

@@
 \author{Ramy Shahin \\ rshahin@cs.toronto.edu}
-\date{December 2, 2017}                                           % Activate to display a given date or no date
+\date{\today}                                           % Activate to display a given date or no date
 
@@
 \section{Strategies for Product-line Verification}
-\cite{Apel:2013}
+
+In \cite{Apel:2013}, product-based, sample-based and family-based model-checking strategies of SPLs are compared.
+Product-based (generating all products and checking each one of them individually) is used as a baseline.
+Sample-based strategies only generate a sample of products (single-wise, pair-wise and triple-wise in terms of valid feature combinations).
+Finally, family-based checking checks the whole SPL in one shot, treating variability points as conditionals.
+
+In this work the sampling strategies used are too primitive.
+Systems defects that are due to the interaction of more than 3 different features (which are not rare) will not be detected by the strategies presented in this paper, simply because this work only samples up to triple-wise feature combinations.
+
+Their family-based approach is more interesting.
+Features are encoded as boolean variables, and feature expressions as Binary Decision Digarams (BDDs).
+Variability within a source-code function is mapped into multiple implementations, with an additional dispatcher function choosing which implementation to call based on feature expressions.
+This simply reduces variability into conditionals, allowing existing model checkers to be used as is.
+
+One caveat that is not explicitly mentioned in the paper is that this family-based approach is not applicable to annotative SPLs.
+The case studies presented are all generative (each feature is implemented separately and then composed into an SPL).
+Different function implementations together with the dispatcher function are generated as a part of the SPL generation.
+However, annotative SPLs are less strict when it comes to variability points (they don't all map to conditionals).
+Late Splitting and Early Joining can't be directly implemented in CPP-based SPLs for example.
+
+The performance of the different strategies is compared in terms of their speedup with respect to the product-based baseline.
+In product-based and sample-based strategies, model checking is terminated once a defect is found in a product.
+However, in the family-based strategy it is not possible to terminate checking when a defect is found, simply because that defect does not necessarily exist in all products.
+This is a threat to validity of the performance numbers, not mentioned in the paper.
+The family-based strategy effectively tries to finds all defects in each product in the SPL, while product-based and strategy-based strategies only find the first defect (if exists).
+Comparing those two different metrics is thus misleading.
 

### 2018-01-03 doc/quals/quals.tex
Liebig's FSE'13 paper

@@
 Variability due to feature expressions is encod
-
-\section{Scalable Analysis of Variable Software}
-\cite{Liebig:2013}
 
@@
 
+\section{Scalable Analysis of Variable Software}
+In \cite{Liebig:2013}, variability-aware (family-based) static analyses are compared to some sample-based techniques.
+Two variability-aware C language analyses are implemented and evaluated: type checking and liveness analysis.
+Three relatively big open source systems were used for evaluation: the Linux kernel, the Busybox codebase, and the OpenSSL library.
+
+In addition to focusing on static analysis as opposed to model checking, there are two main differences between this work and that of \cite{Apel:2013}.
+First, the variability-aware analyses are applied to annotative SPLs (using the CPP to incorporate variability).
+Because annotative SPLs and more permissive than generative SPLs in terms of syntactic variability, some building blocks had to be purposely designed to be variability-aware.
+Those include variability-aware Abstract Syntax Trees (ASTs), Control Flow Graphs (CFGs) and symbol tables.
+In addition, a dedicated variability-aware parser (incorporating CPP logic) is used.
+
+The second difference with respect to \cite{Apel:2013} is that the sampling techniques used in the comparison are more elaborate.
+In addition to pair-wise sampling (covered in \cite{Apel:2013}), single configuration, random and coverage-based sampling were also evaluated.
+
+Similar to \cite{Apel:2013}, the main principles guiding the design of the variability-aware analyses are late splitting and early joining.
+These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperformaing brute-force per-product analysis by orders of magnitude.
+A third principle implicit in \cite{Apel:2013} but explicitly spelled out here is local variability representation.
+This is the mechanism of maintaining variability within intermediate analysis values.
+
+The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples.
+In addition, in some cases variability-aware analyses also outperform coverage-based analysis.
+More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e.
+all valid products), while all the sampling techniques are incomplete in that regard.
+
 \bibliographystyle{abbrv}

### 2018-01-06 doc/quals/quals.tex
SuperC

@@
 
-\section{SuperC}
-\cite{Gazzillo:2012}
-
 \section{Variability-Aware Abstract Interpretation}
@@
 
+\section{SuperC}
+In \cite{Gazzillo:2012}, a variability-aware C-language parser and preprocessor are presented and evaluated against a similar C parser implemented as a part of the TypeChef project \cite{Kastner:2011}.
+SuperC includes a lexer, a preprocessor and a Fork Merge LR (FMLR) parser.
+Several optimizations have been implemented to lower the number of forked subparsers, and consequently speedup parsing.
+On the x86 Linux kernel codebase, SuperC is 3.4 to 3.8 times faster than TypeChef.
+
+The idea behind FMLR parsing is to fork subparsers (initially sharing the state of their parent) on each variability point.
+A naive implementation would result in a number of subparsers exponential in the number of variability features.
+To prevent this exponential explosion, SuperC implements 4 optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces.
+With those optimizations, for almost all the Linux kernel compilation units the number of subparsers is less than 20.
+
+In addition to the optimizations, SuperC uses Binary Decision Diagrams (BDDs) to encode and test the satisfiability of presence conditions, while TypeChef uses a SAT-solver.
+Presence conditions are conjuncted on the nesting of static conditional compilation directives, and disjuncted on merges.
+With a few (tens) of propositional variables, conjunction and disjunction of BDDs is typically more efficient than applying the same operations to CNF-encoded boolean formulas (processed by SAT-solvers).
+As a result, BDDs also contribute to SuperC's speedup compared to TypeChef.
+
+Another advantage of SuperC compared to TypeChef is that the SuperC preprocessor and parser take into account the possible interactions between different preprocessing and parsing rules of the C language.
+TypeChef misses some of those interactions (and also some non-standard preprocessing directives), which is the reason why it fails to parse thousands of the Linux kernel source files.
+SuperC on the other hand parses the whole codebase.
+
 \bibliographystyle{abbrv}

### 2018-01-08 doc/quals/quals.tex
SPLLift

@@
 \section{SPLLift}
-\cite{Bodden:2013}
 
-In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS framework) is extended to annotative product lines.
-In IFDS, the data flow analysis problem is encoded as a graph reachability problem.
-Variability due to feature expressions is encod
+In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS \cite{Reps:1995} framework) is extended to annotative product lines.
+In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
+The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
+This fits well within the IFDS framework because data flow properties are boolean expressions.
+
+This lifting approach scales well with the number of features.
+Complexity grows polynomially with the number of statements in an SPL, rather than exponentially in the number of features.
+In addition, the implementation takes the feature model of the SPL into consideration, pruning statements with invalid presence conditions.
+This eliminates the need to do excessive computations that are of no value to the SPL analysis.
+
+This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Unititialized Variables.
+4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
+Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies.
+One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.
 

### 2018-01-09 doc/quals/quals.tex
Abstract Interpretation

@@
 
-\begin{abstract}
-TODO
-\end{abstract}
+%\begin{abstract}
+%TODO
+%\end{abstract}
 
@@
 This eliminates the need to do excessive computations that are of no value to the SPL analysis.
+Presence conditions are encoded as BDDs, which further optimizes their satisfiability checks.
 
-This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Unititialized Variables.
+This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables.
 4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
@@
 \section{Variability-Aware Abstract Interpretation}
-\cite{Midtgaard:2015}
+
+In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses to SPLs is presented.
+The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level.
+Formal proofs for the correctness of the lifting approach are also presented.
+
+The lifting approach provides systematic liftings of the abstract domains, Galois connections, and corresponding analyses by the use of combinators.
+It is worth noting that this approach is systematic but not automated.
+Language-specific theorems need to be proven (3 of them have been proven in the paper), in addition to a language-agnostic framework theorem that has been proven as well.
+
+The paper also presents some optimizations that would trade analysis precision for speed.
+The idea is to allow for variational abstraction, which would further abstract variability.
+Further optimization techniques involve sharing analysis intermediate values while keeping track of presence conditions (either in CNF/DNF formulas or BDDs).
+In addition, a technique for extracting data-flow equations (both the product-level equations and the lifted equations) is explained.
+
+This work was evaluated on a Reaching Definitions analysis for Java programs.
+Three Java SPLs were used: GPL, Prevayler and BerkeleyDB.
+The speedup due to the different optimizations for the lifting approach (with respect to brute force) is exponential with respect to the number of SPL configurations.
 
@@
 Their family-based approach is more interesting.
-Features are encoded as boolean variables, and feature expressions as Binary Decision Digarams (BDDs).
+Features are encoded as boolean variables, and feature expressions as Binary Decision Diagrams (BDDs).
 Variability within a source-code function is mapped into multiple implementations, with an additional dispatcher function choosing which implementation to call based on feature expressions.
@@
 Similar to \cite{Apel:2013}, the main principles guiding the design of the variability-aware analyses are late splitting and early joining.
-These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperformaing brute-force per-product analysis by orders of magnitude.
+These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperforming brute-force per-product analysis by orders of magnitude.
 A third principle implicit in \cite{Apel:2013} but explicitly spelled out here is local variability representation.
@@
 SuperC includes a lexer, a preprocessor and a Fork Merge LR (FMLR) parser.
-Several optimizations have been implemented to lower the number of forked subparsers, and consequently speedup parsing.
+Several optimizations have been implemented to lower the number of forked sub-parsers, and consequently speedup parsing.
 On the x86 Linux kernel codebase, SuperC is 3.4 to 3.8 times faster than TypeChef.
 
-The idea behind FMLR parsing is to fork subparsers (initially sharing the state of their parent) on each variability point.
-A naive implementation would result in a number of subparsers exponential in the number of variability features.
+The idea behind FMLR parsing is to fork sub-parsers (initially sharing the state of their parent) on each variability point.
+A naive implementation would result in a number of sub-parsers exponential in the number of variability features.
 To prevent this exponential explosion, SuperC implements 4 optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces.
-With those optimizations, for almost all the Linux kernel compilation units the number of subparsers is less than 20.
+With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.
 

### 2018-01-12 doc/quals/quals.tex
Abstract Interpretation - take 2

@@
 
-In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses to SPLs is presented.
+\newcommand{\configSpace}{\mathbb{K}}
+\newcommand{\featSpace}{\mathbb{F}}
+\newcommand{\lang}{\mathcal{L}}
+
+In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented.
 The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level.
-Formal proofs for the correctness of the lifting approach are also presented.
+Formal proofs of the soundness of the lifting approach are also presented.
 
-The lifting approach provides systematic liftings of the abstract domains, Galois connections, and corresponding analyses by the use of combinators.
-It is worth noting that this approach is systematic but not automated.
-Language-specific theorems need to be proven (3 of them have been proven in the paper), in addition to a language-agnostic framework theorem that has been proven as well.
+Given a programming language $\lang$, starting with an abstract intepretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$.
+If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.
 
-The paper also presents some optimizations that would trade analysis precision for speed.
-The idea is to allow for variational abstraction, which would further abstract variability.
-Further optimization techniques involve sharing analysis intermediate values while keeping track of presence conditions (either in CNF/DNF formulas or BDDs).
-In addition, a technique for extracting data-flow equations (both the product-level equations and the lifted equations) is explained.
+The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length.
+$|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable.
+Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features.
+Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.
 
-This work was evaluated on a Reaching Definitions analysis for Java programs.
-Three Java SPLs were used: GPL, Prevayler and BerkeleyDB.
-The speedup due to the different optimizations for the lifting approach (with respect to brute force) is exponential with respect to the number of SPL configurations.
+\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations.
+This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each.
+The shorter the tuple, the faster the analysis since it will need to process less intermediate values.
+Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.
+
+\emph{Variational abstraction} pushes abstraction up to the configuration space.
+Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation.
+Pretty much like all abstraction techniques, this trades accuracy for performance.
+\emph{Variational abstraction} is implemented the same as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs.
+Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where collapses the configuration space into a single value).
+Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.
+
+This lifting approach was evaluated on a single analysis (reaching definitions).
+Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used.
+However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB).
+The configuration space of each of those methods varies between 8 and 106 configurations.
+
+The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away).
+Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration.
+The speedup due to lifting with sharing also grows exponentially, but at a slower rate.
+Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).
 

### 2018-01-14 doc/quals/quals.tex
FTS take 2

@@
 \section{Featured Transition Systems}
-\cite{Classen:2013}
 
-In \cite{Classen:2013}, Labeled Transition Systems (LTSs) are extended with feature expressions into Featured Transition Systems (FTSs).
-FTSs can be used to encode variability-aware models, which can be checked by the SNIP model checker.
+Classen et al.\cite{Classen:2013} address the problem of model checking variability-aware systems.
+The paper introduces the Featured Transition System (FTS) formalism of models, the feature Linear Temporal Logic (fLTL) for expressing temporal logic properties, and the SNIP model checker that checks fLTL properties against FTS models.
 
-The paper provides a FTS-based formalism, together with model checking algorithms, complexity analysis (with proofs) and the featured LTL (fLTL) logic.
-Several state space reduction optimizations are also explained.
-The empirical evaluation highlights the speedup due to the optimized algorithm.
+An FTS extends a Labeled Transition System (LTS) with a total function $\gamma$ mapping state transitions to presence conditions.
+This way behavior due to different model variants can be guarded by the presence conditions of those variants.
+In addition, an FTS is nonmonotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.
+
+Parallel composition of FTSs is also defined similarly to that of LTSs.
+When two FTSs are parallely composed, independent transition actions are asynchronously interleaved.
+However, shared actions are synchronized and their corresponding presence conditions are conjoined.
+
+Properties over FTSs are specified in fLTL, which associates presence conditions to LTL formulas.
+Now a variability-aware model checker needs to check an fLTL property against an FTS model.
+The checker returns true if the property holds in the set of products defined by the property's presence condition.
+If the property is violated in any of those products, a set of counterexample/presence condition pairs is returned.
+Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.
+
+The FTS model checking algorithm extends the state reachability algorithm for LTSs with the identification of the sets of products in which each state is reachable from an initial state.
+Sets ot products are defined in terms of feature expressions, which are encoded as BDDs.
+The FD of the model is used to exclude invalid products.
+Since only a single counterexample is needed whenever a product violates the property, the search can be pruned for products where a counterexample has already been found.
+
+The SNIP model checker implements FTS model checking.
+Its input language is fPromela, which extends Promela (used in the SPIN model checker) by feature expression guard annotation over statements.
+It also implements variability-aware vacuity detection and deadlock detection.
 

### 2018-01-20 doc/quals/quals.tex
Intro

@@
 \section{Introduction}
-TODO
+A Software Product Line (SPL) is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases).
+An SPL is composed of a set of features.
+A feature is an externally visible property, aspect or quality of a software system.
+An instance (product) of an SPL is composed of a subset of the SPL features.
+Not all combinations of features are valid though.
+A Feature Model (FM) - also known as a Variability Model (VM) - is a specification of the valid combinations of SPL features.
+An FM can be expressed as a boolean formula over features, or graphically as a Feature Diagram (FD).
+
+When building an SPL, software analysis tools are still needed to find bugs, generate test cases, collect metrics or track progress.
+A plethora of tools already exists, covering many different categories of analyses.
+However, these tools work on the artifacts of a single product.
+Because an SPL encompasses a superset of features existing in a single product, those tools are not capable of identifying which artifacts belong to which set of products.
+Consequently, those tools cannot be used as is on an SPL.
+A naive solution is to generate all possible products of an SPL and run tools on each one of them.
+This is generally intractable though, because the number of products grows exponentially with the number of SPL features.
+
+This position paper covers some research work attempting to solve this problem.
+One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques\cite{Apel:2013}.
+Another extends this work to more elaborate sampling techniques and more general categories of analysis\cite{Liebig:2013}.
+At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs\cite{Midtgaard:2015}.
+At the other end, some papers provide pragmatic implementations of source code syntax checkers\cite{Gazzillo:2012}, type checkers\cite{Kastner:2012}, data flow analyses\cite{Bodden:2013}, and model checkers\cite{Classen:2013}.
+
+\section{Variability-Aware Abstract Interpretation}
+
+\newcommand{\configSpace}{\mathbb{K}}
+\newcommand{\featSpace}{\mathbb{F}}
+\newcommand{\lang}{\mathcal{L}}
+
+In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented.
+The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level.
+Formal proofs of the soundness of the lifting approach are also presented.
+
+Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$.
+If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.
+
+The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length.
+$|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable.
+Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features.
+Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.
+
+\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations.
+This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each.
+The shorter the tuple, the faster the analysis since it will need to process less intermediate values.
+Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.
+
+\emph{Variational abstraction} pushes abstraction up to the configuration space.
+Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation.
+Pretty much like all abstraction techniques, this trades accuracy for performance.
+\emph{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs.
+Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where the configuration space is collapsed into a single value).
+Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.
+
+This lifting approach was evaluated on a single analysis (reaching definitions).
+Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used.
+However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB).
+The configuration space of each of those methods varies between 8 and 106 configurations.
+
+The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away).
+Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration.
+The speedup due to lifting with sharing also grows exponentially, but at a slower rate.
+Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).
 
@@
 This way behavior due to different model variants can be guarded by the presence conditions of those variants.
-In addition, an FTS is nonmonotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.
+In addition, an FTS is non-monotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.
 
 Parallel composition of FTSs is also defined similarly to that of LTSs.
-When two FTSs are parallely composed, independent transition actions are asynchronously interleaved.
+When two FTSs are composed in parallel, independent transition actions are asynchronously interleaved.
 However, shared actions are synchronized and their corresponding presence conditions are conjoined.
@@
 With only one model (moreover, a small one), the evaluation results cannot be safely generalized.
-
-\section{SPLLift}
-
-In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS \cite{Reps:1995} framework) is extended to annotative product lines.
-In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
-The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
-This fits well within the IFDS framework because data flow properties are boolean expressions.
-
-This lifting approach scales well with the number of features.
-Complexity grows polynomially with the number of statements in an SPL, rather than exponentially in the number of features.
-In addition, the implementation takes the feature model of the SPL into consideration, pruning statements with invalid presence conditions.
-This eliminates the need to do excessive computations that are of no value to the SPL analysis.
-Presence conditions are encoded as BDDs, which further optimizes their satisfiability checks.
-
-This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables.
-4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
-Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies.
-One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.
 
@@
 It is worth noting though that these product lines are relatively small, both in terms of number of lines of code and number of features.
-
-\section{Variability-Aware Abstract Interpretation}
-
-\newcommand{\configSpace}{\mathbb{K}}
-\newcommand{\featSpace}{\mathbb{F}}
-\newcommand{\lang}{\mathcal{L}}
-
-In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented.
-The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level.
-Formal proofs of the soundness of the lifting approach are also presented.
-
-Given a programming language $\lang$, starting with an abstract intepretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$.
-If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.
-
-The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length.
-$|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable.
-Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features.
-Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.
-
-\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations.
-This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each.
-The shorter the tuple, the faster the analysis since it will need to process less intermediate values.
-Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.
-
-\emph{Variational abstraction} pushes abstraction up to the configuration space.
-Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation.
-Pretty much like all abstraction techniques, this trades accuracy for performance.
-\emph{Variational abstraction} is implemented the same as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs.
-Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where collapses the configuration space into a single value).
-Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.
-
-This lifting approach was evaluated on a single analysis (reaching definitions).
-Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used.
-However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB).
-The configuration space of each of those methods varies between 8 and 106 configurations.
-
-The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away).
-Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration.
-The speedup due to lifting with sharing also grows exponentially, but at a slower rate.
-Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).
 
@@
 
+\section{SPLLift}
+
+In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS \cite{Reps:1995} framework) is extended to annotative product lines.
+In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
+The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
+This fits well within the IFDS framework because data flow properties are boolean expressions.
+
+This lifting approach scales well with the number of features.
+Complexity grows polynomially with the number of statements in an SPL, rather than exponentially in the number of features.
+In addition, the implementation takes the feature model of the SPL into consideration, pruning statements with invalid presence conditions.
+This eliminates the need to do excessive computations that are of no value to the SPL analysis.
+Presence conditions are encoded as BDDs, which further optimizes their satisfiability checks.
+
+This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables.
+4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
+Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies.
+One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.
+
 \section{SuperC}

### 2018-03-05 doc/quals/quals.tex
Conclusion

@@
 
+% macros
+\newcommand{\term}[1] {\emph{#1}}
+
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
@@
 \section{Introduction}
-A Software Product Line (SPL) is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases).
+A \term{Software Product Line (SPL)} is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases).
 An SPL is composed of a set of features.
-A feature is an externally visible property, aspect or quality of a software system.
-An instance (product) of an SPL is composed of a subset of the SPL features.
+A \term{feature} is an externally visible property, aspect or quality of a software system.
+An \term{instance} (product) of an SPL is composed of a subset of the SPL features.
 Not all combinations of features are valid though.
-A Feature Model (FM) - also known as a Variability Model (VM) - is a specification of the valid combinations of SPL features.
-An FM can be expressed as a boolean formula over features, or graphically as a Feature Diagram (FD).
+A \term{Feature Model (FM)} - also known as a \term{Variability Model (VM)} - is a specification of the valid combinations of SPL features.
+An FM can be expressed as a boolean formula over features, or graphically as a \term{Feature Diagram (FD)}.
 
@@
 
-This position paper covers some research work attempting to solve this problem.
-One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques\cite{Apel:2013}.
-Another extends this work to more elaborate sampling techniques and more general categories of analysis\cite{Liebig:2013}.
-At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs\cite{Midtgaard:2015}.
-At the other end, some papers provide pragmatic implementations of source code syntax checkers\cite{Gazzillo:2012}, type checkers\cite{Kastner:2012}, data flow analyses\cite{Bodden:2013}, and model checkers\cite{Classen:2013}.
+This report covers some research work attempting to solve this problem.
+One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques~\cite{Apel:2013}.
+Another extends this work to more elaborate sampling techniques and more general categories of analysis~\cite{Liebig:2013}.
+At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs~\cite{Midtgaard:2015}.
+At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}.
+
+\section{Strategies for Product-line Verification}
+
+Apel et al.~\cite{Apel:2013} compare product-based, sample-based and family-based model-checking strategies of SPLs.
+Product-based (generating all products and checking each one of them individually) is used as a baseline.
+Sample-based strategies only generate a sample of products (single-wise, pair-wise and triple-wise in terms of valid feature combinations).
+Finally, family-based checking checks the whole SPL in one shot, treating variability points as conditionals.
+
+This work only samples up to triple-wise feature combinations.
+As a result, system defects that are due to the interaction of more than three different features (which are not rare) cannot be detected by those sampling strategies.
+
+Their family-based approach is more interesting.
+Features are encoded as boolean variables, and feature expressions as Binary Decision Diagrams (BDDs).
+Variability within a source-code function is mapped into multiple implementations, with an additional dispatcher function choosing which implementation to call based on feature expressions.
+This simply reduces variability into conditionals, allowing existing model checkers to be used as is.
+
+One caveat that is not explicitly mentioned in the paper is that this family-based approach is not applicable to annotative SPLs.
+The case studies presented are all generative (each feature is implemented separately and then composed into an SPL).
+Different function implementations together with the dispatcher function are generated as a part of the SPL generation.
+However, annotative SPLs are less strict when it comes to variability points (they don't all map to conditionals).
+Late Splitting and Early Joining cannot be directly implemented in C Pre-Processor (CPP)-based SPLs for example.
+
+The performance of the different strategies is compared in terms of their speedup with respect to the product-based baseline.
+In product-based and sample-based strategies, model checking is terminated once a defect is found in a product.
+However, in the family-based strategy it is not possible to terminate checking when a defect is found, simply because that defect does not necessarily exist in all products.
+This is a threat to validity of the performance numbers, not mentioned in the paper.
+The family-based strategy effectively tries to find all defects in each product in the SPL, while product-based and sampling-based strategies only find the first defect (if one exists).
+Comparing those two different metrics is thus misleading.
+
+\section{Scalable Analysis of Variable Software}
+Liebig et al.~\cite{Liebig:2013} compare variability-aware (family-based) static analyses to some product sampling techniques.
+Two variability-aware C language analyses are implemented and evaluated: type checking and liveness analysis.
+Three relatively big open source systems were used for evaluation: the Linux kernel, the Busybox codebase, and the OpenSSL library.
+
+In addition to focusing on static analysis as opposed to model checking, there are two main differences between this work and that of Apel et al.~\cite{Apel:2013}.
+First, the variability-aware analyses are applied to annotative SPLs (using the CPP to annotate variability).
+Because annotative SPLs are more permissive than generative SPLs in terms of syntactic variability, some building blocks had to be purposely designed to be variability-aware.
+Those include variability-aware Abstract Syntax Trees (ASTs), Control Flow Graphs (CFGs) and symbol tables.
+In addition, a dedicated variability-aware parser (including CPP logic) is used.
+
+The second difference is that the sampling techniques used in the comparison are more elaborate.
+In addition to pair-wise sampling~\cite{Apel:2013}, single configuration, random and coverage-based sampling were also evaluated.
+
+The main principles guiding the design of the variability-aware analyses are late splitting, early joining and local variability representation.
+These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperforming brute-force per-product analysis by orders of magnitude.
+Local variability representation efficiently maintains variability within intermediate analysis values.
+Each of the three principles appear in most of the papers covered in this report, but this paper clearly and explicitly identify and explain them.
+
+The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples.
+In addition, in some cases variability-aware analyses also outperform coverage-based analysis.
+More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e.
+all valid products), while all the sampling techniques are incomplete in that regard.
 
@@
 
-In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented.
+Midtgaard et al.~\cite{Midtgaard:2015} present a systematic approach to lifting abstract interpretation analyses from single programs to SPLs.
 The paper provides a rigorous formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level.
@@
 
-Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$.
-If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.
+Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \term{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$.
+If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \term{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.
 
@@
 Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features.
-Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.
+Two optimizations are proposed in the paper: \term{value sharing} and \term{variational abstraction}.
 
-\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations.
+\term{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations.
 This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each.
@@
 
-\emph{Variational abstraction} pushes abstraction up to the configuration space.
+\term{Variational abstraction} pushes abstraction up to the configuration space.
 Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation.
 Pretty much like all abstraction techniques, this trades accuracy for performance.
-\emph{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs.
+\term{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs.
 Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where the configuration space is collapsed into a single value).
-Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.
+Since Galois connections compose (the composition of two Galois connections is a Galois connection), \term{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.
 
@@
 Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used.
-However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB).
-The configuration space of each of those methods varies between 8 and 106 configurations.
+However, the evaluation numbers presented are not aggregated across the three codebases, but rather for only four source code methods (two from GPL, one from Prevayler and one from BerkeleyDB).
+The configuration space of each of those methods varies between eight and 106 configurations.
 
@@
 The speedup due to lifting with sharing also grows exponentially, but at a slower rate.
-Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).
-
-\section{Featured Transition Systems}
-
-Classen et al.\cite{Classen:2013} address the problem of model checking variability-aware systems.
-The paper introduces the Featured Transition System (FTS) formalism of models, the feature Linear Temporal Logic (fLTL) for expressing temporal logic properties, and the SNIP model checker that checks fLTL properties against FTS models.
-
-An FTS extends a Labeled Transition System (LTS) with a total function $\gamma$ mapping state transitions to presence conditions.
-This way behavior due to different model variants can be guarded by the presence conditions of those variants.
-In addition, an FTS is non-monotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.
-
-Parallel composition of FTSs is also defined similarly to that of LTSs.
-When two FTSs are composed in parallel, independent transition actions are asynchronously interleaved.
-However, shared actions are synchronized and their corresponding presence conditions are conjoined.
-
-Properties over FTSs are specified in fLTL, which associates presence conditions to LTL formulas.
-Now a variability-aware model checker needs to check an fLTL property against an FTS model.
-The checker returns true if the property holds in the set of products defined by the property's presence condition.
-If the property is violated in any of those products, a set of counterexample/presence condition pairs is returned.
-Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.
-
-The FTS model checking algorithm extends the state reachability algorithm for LTSs with the identification of the sets of products in which each state is reachable from an initial state.
-Sets ot products are defined in terms of feature expressions, which are encoded as BDDs.
-The FD of the model is used to exclude invalid products.
-Since only a single counterexample is needed whenever a product violates the property, the search can be pruned for products where a counterexample has already been found.
-
-The SNIP model checker implements FTS model checking.
-Its input language is fPromela, which extends Promela (used in the SPIN model checker) by feature expression guard annotation over statements.
-It also implements variability-aware vacuity detection and deadlock detection.
-
-One shortcoming of this paper is that empirical evaluation is done on only one model (the mine pump controller model).
-This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties.
-With only one model (moreover, a small one), the evaluation results cannot be safely generalized.
+Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only four source code methods).
 
@@
 
-In \cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq\cite{Bertot:2010} proof assistant.
-Featherweight Java (FJ) is a simplistic subset of Java, leaving out constructs such as assignment (mutation) and interfaces.
+K\"{a}stner et a.~\cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq~\cite{Bertot:2010} proof assistant.
+Featherweight Java (FJ) is a functional subset of Java, leaving out constructs such as assignment (mutation) and interfaces.
 CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE.
@@
 
-\section{Strategies for Product-line Verification}
-
-In \cite{Apel:2013}, product-based, sample-based and family-based model-checking strategies of SPLs are compared.
-Product-based (generating all products and checking each one of them individually) is used as a baseline.
-Sample-based strategies only generate a sample of products (single-wise, pair-wise and triple-wise in terms of valid feature combinations).
-Finally, family-based checking checks the whole SPL in one shot, treating variability points as conditionals.
-
-In this work the sampling strategies used are too primitive.
-Systems defects that are due to the interaction of more than 3 different features (which are not rare) will not be detected by the strategies presented in this paper, simply because this work only samples up to triple-wise feature combinations.
-
-Their family-based approach is more interesting.
-Features are encoded as boolean variables, and feature expressions as Binary Decision Diagrams (BDDs).
-Variability within a source-code function is mapped into multiple implementations, with an additional dispatcher function choosing which implementation to call based on feature expressions.
-This simply reduces variability into conditionals, allowing existing model checkers to be used as is.
-
-One caveat that is not explicitly mentioned in the paper is that this family-based approach is not applicable to annotative SPLs.
-The case studies presented are all generative (each feature is implemented separately and then composed into an SPL).
-Different function implementations together with the dispatcher function are generated as a part of the SPL generation.
-However, annotative SPLs are less strict when it comes to variability points (they don't all map to conditionals).
-Late Splitting and Early Joining can't be directly implemented in CPP-based SPLs for example.
-
-The performance of the different strategies is compared in terms of their speedup with respect to the product-based baseline.
-In product-based and sample-based strategies, model checking is terminated once a defect is found in a product.
-However, in the family-based strategy it is not possible to terminate checking when a defect is found, simply because that defect does not necessarily exist in all products.
-This is a threat to validity of the performance numbers, not mentioned in the paper.
-The family-based strategy effectively tries to finds all defects in each product in the SPL, while product-based and strategy-based strategies only find the first defect (if exists).
-Comparing those two different metrics is thus misleading.
-
-\section{Scalable Analysis of Variable Software}
-In \cite{Liebig:2013}, variability-aware (family-based) static analyses are compared to some sample-based techniques.
-Two variability-aware C language analyses are implemented and evaluated: type checking and liveness analysis.
-Three relatively big open source systems were used for evaluation: the Linux kernel, the Busybox codebase, and the OpenSSL library.
-
-In addition to focusing on static analysis as opposed to model checking, there are two main differences between this work and that of \cite{Apel:2013}.
-First, the variability-aware analyses are applied to annotative SPLs (using the CPP to incorporate variability).
-Because annotative SPLs and more permissive than generative SPLs in terms of syntactic variability, some building blocks had to be purposely designed to be variability-aware.
-Those include variability-aware Abstract Syntax Trees (ASTs), Control Flow Graphs (CFGs) and symbol tables.
-In addition, a dedicated variability-aware parser (incorporating CPP logic) is used.
-
-The second difference with respect to \cite{Apel:2013} is that the sampling techniques used in the comparison are more elaborate.
-In addition to pair-wise sampling (covered in \cite{Apel:2013}), single configuration, random and coverage-based sampling were also evaluated.
-
-Similar to \cite{Apel:2013}, the main principles guiding the design of the variability-aware analyses are late splitting and early joining.
-These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperforming brute-force per-product analysis by orders of magnitude.
-A third principle implicit in \cite{Apel:2013} but explicitly spelled out here is local variability representation.
-This is the mechanism of maintaining variability within intermediate analysis values.
-
-The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples.
-In addition, in some cases variability-aware analyses also outperform coverage-based analysis.
-More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e.
-all valid products), while all the sampling techniques are incomplete in that regard.
-
-\section{SPLLift}
-
-In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS \cite{Reps:1995} framework) is extended to annotative product lines.
-In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
-The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
-This fits well within the IFDS framework because data flow properties are boolean expressions.
-
-This lifting approach scales well with the number of features.
-Complexity grows polynomially with the number of statements in an SPL, rather than exponentially in the number of features.
-In addition, the implementation takes the feature model of the SPL into consideration, pruning statements with invalid presence conditions.
-This eliminates the need to do excessive computations that are of no value to the SPL analysis.
-Presence conditions are encoded as BDDs, which further optimizes their satisfiability checks.
-
-This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables.
-4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
-Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies.
-One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.
-
 \section{SuperC}
-In \cite{Gazzillo:2012}, a variability-aware C-language parser and preprocessor are presented and evaluated against a similar C parser implemented as a part of the TypeChef project \cite{Kastner:2011}.
-SuperC includes a lexer, a preprocessor and a Fork Merge LR (FMLR) parser.
+Gazzillo and Grimm~\cite{Gazzillo:2012} present a variability-aware C-language parser and preprocessor, and evaluate it against a similar C parser implemented as a part of the TypeChef project~\cite{Kastner:2011, Kastner:2012}.
+SuperC includes a lexer, a preprocessor and a \term{Fork Merge LR (FMLR)} parser.
 Several optimizations have been implemented to lower the number of forked sub-parsers, and consequently speedup parsing.
@@
 A naive implementation would result in a number of sub-parsers exponential in the number of variability features.
-To prevent this exponential explosion, SuperC implements 4 optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces.
+To prevent this exponential explosion, SuperC implements four optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces.
 With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.
@@
 
+\section{SPLLift}
+
+Bodden et al.~\cite{Bodden:2013} extend inter-procedural data flow analysis (based on the IFDS~\cite{Reps:1995} framework) to annotative product lines.
+In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
+The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
+This fits well within the IFDS framework because data flow properties are boolean expressions.
+
+This lifting approach scales well with the number of features.
+Complexity grows polynomially with the number of statements in an SPL, rather than exponentially in the number of features.
+In addition, the implementation takes the feature model of the SPL into consideration, pruning statements with invalid presence conditions.
+This eliminates the need to do excessive computations that are of no value to the SPL analysis.
+Presence conditions are encoded as BDDs, which further optimizes their satisfiability checks.
+
+This work was evaluated on three lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables.
+Four real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08.
+Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies.
+One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.
+
+\section{Featured Transition Systems}
+
+Classen et al.~\cite{Classen:2013} address the problem of model checking variability-aware systems.
+The paper introduces the Featured Transition System (FTS) formalism of models, the feature Linear Temporal Logic (fLTL) for expressing temporal logic properties, and the SNIP model checker that checks fLTL properties against FTS models.
+
+An FTS extends a Labeled Transition System (LTS) with a total function $\gamma$ mapping state transitions to presence conditions.
+This way behavior due to different model variants can be guarded by the presence conditions of those variants.
+In addition, an FTS is non-monotonic in the sense that existing behavior can be removed by a variant by assigning the negation of the presence condition of that variant to the transitions associated with the behavior.
+
+Parallel composition of FTSs is also defined similarly to that of LTSs.
+When two FTSs are composed in parallel, independent transition actions are asynchronously interleaved.
+However, shared actions are synchronized and their corresponding presence conditions are conjoined.
+
+Properties over FTSs are specified in fLTL, which associates presence conditions to LTL formulas.
+Now a variability-aware model checker needs to check an fLTL property against an FTS model.
+The checker returns true if the property holds in the set of products defined by the property's presence condition.
+If the property is violated in any of those products, a set of counterexample/presence condition pairs is returned.
+Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.
+
+The FTS model checking algorithm extends the state reachability algorithm for LTSs with the identification of the sets of products in which each state is reachable from an initial state.
+Sets of products are defined in terms of feature expressions, which are encoded as BDDs.
+The FD of the model is used to exclude invalid products.
+Since only a single counterexample is needed whenever a product violates the property, the search can be pruned for products where a counterexample has already been found.
+
+The SNIP model checker implements FTS model checking.
+Its input language is fPromela, which extends Promela (used in the SPIN model checker) by feature expression guard annotation over statements.
+It also implements variability-aware vacuity detection and deadlock detection.
+
+One shortcoming of this paper is that empirical evaluation is done on only one model (the mine pump controller model).
+This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties.
+With only one model (moreover, a small one), the evaluation results cannot be safely generalized.
+
+\section{Conclusion}
+
+\begin{table}[h!]
+\begin{tabular}{ |l|l|l|l|}
+\hline
+Paper & Analysis & Limitation \\
+\hline
+Apel et al.~\cite{Apel:2013} & Model Checking & Generative SPLs only  \\
+Liebig et al.~\cite{Liebig:2013} & Static Analysis & Hand-crafted analyses and infrastructure  \\
+Midtgaard et al.~\cite{Midtgaard:2015} & Abstract Interpretation & Systematic but not automated \\
+K\"{a}stner et a.~\cite{Kastner:2012} & Type Checking & Limited source code variability \\
+Gazzillo and Grimm~\cite{Gazzillo:2012} & Parsing (Syntax Checking) & Fully hand-crafted \\
+Bodden et al.~\cite{Bodden:2013} & Data-flow Analysis & IFDS analyses only \\
+Classen et al.~\cite{Classen:2013} & Model Checking & Hand-crafted \\
+\hline
+\end{tabular}
+\caption{Different SPL analysis techniques covered in this report.}
+\label{tbl1}
+\end{table}
+
+Family-based analysis of SPLs covers all valid product line instances, and in many cases outperforms sampling techniques~\cite{Apel:2013, Liebig:2013}.
+Different family-based analysis techniques have been covered in this report.
+Table \ref{tbl1} summarizes the papers covered.
+
+Late splitting and early joining are two principles that underly many of the techniques presented.
+These two principles maximize reuse of intermediate results calculated as a part of the analysis process.
+Local variability representation is another principle that manages intermediate value variability.
+These principles together with some analysis specific design decisions can effectively lead to an efficient family-based analysis.
 \bibliographystyle{abbrv}

### 2018-03-05 doc/quals/quals.tex
moving table to intro

@@
 At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}.
+Table \ref{tbl1} summarizes the papers covered, with limitations explained throughout this report.
+
+\begin{table}[h!]
+\begin{tabular}{ |l|l|l|l|}
+\hline
+Paper & Analysis & Limitations \\
+\hline
+Apel et al.~\cite{Apel:2013} & Model Checking & Generative SPLs only  \\
+Liebig et al.~\cite{Liebig:2013} & Static Analysis & Hand-crafted analyses and infrastructure  \\
+Midtgaard et al.~\cite{Midtgaard:2015} & Abstract Interpretation & Systematic but not automated \\
+K\"{a}stner et a.~\cite{Kastner:2012} & Type Checking & Limited source code variability \\
+Gazzillo and Grimm~\cite{Gazzillo:2012} & Parsing (Syntax Checking) & Fully hand-crafted \\
+Bodden et al.~\cite{Bodden:2013} & Data-flow Analysis & IFDS analyses only \\
+Classen et al.~\cite{Classen:2013} & Model Checking & Hand-crafted \\
+\hline
+\end{tabular}
+\caption{Different SPL analysis techniques covered in this report.}
+\label{tbl1}
+\end{table}
 
@@
 
-\begin{table}[h!]
-\begin{tabular}{ |l|l|l|l|}
-\hline
-Paper & Analysis & Limitation \\
-\hline
-Apel et al.~\cite{Apel:2013} & Model Checking & Generative SPLs only  \\
-Liebig et al.~\cite{Liebig:2013} & Static Analysis & Hand-crafted analyses and infrastructure  \\
-Midtgaard et al.~\cite{Midtgaard:2015} & Abstract Interpretation & Systematic but not automated \\
-K\"{a}stner et a.~\cite{Kastner:2012} & Type Checking & Limited source code variability \\
-Gazzillo and Grimm~\cite{Gazzillo:2012} & Parsing (Syntax Checking) & Fully hand-crafted \\
-Bodden et al.~\cite{Bodden:2013} & Data-flow Analysis & IFDS analyses only \\
-Classen et al.~\cite{Classen:2013} & Model Checking & Hand-crafted \\
-\hline
-\end{tabular}
-\caption{Different SPL analysis techniques covered in this report.}
-\label{tbl1}
-\end{table}
-
 Family-based analysis of SPLs covers all valid product line instances, and in many cases outperforms sampling techniques~\cite{Apel:2013, Liebig:2013}.
 Different family-based analysis techniques have been covered in this report.
-Table \ref{tbl1} summarizes the papers covered.
 
@@
 Local variability representation is another principle that manages intermediate value variability.
-These principles together with some analysis specific design decisions can effectively lead to an efficient family-based analysis.
+These principles together with some analysis-specific design decisions can effectively lead to an efficient family-based analysis.
 \bibliographystyle{abbrv}

### 2018-03-07 doc/quals/background.tex
final version of the quals document

\documentclass[11pt]{article}
\begin{document}

My name is Ramy Shahin, a second year PhD student supervised by Marsha Chechik. I had my M.Sc. degree in Computer Science from Michigan State University in 2001, and since then I had worked as a Software Engineer for several companies, including Microsoft, IBM and Amazon.

My research focuses on analysis techniques for Software Product Lines (SPLs). In particular, I have been studying previous attempts at lifting product-level analyses to SPLs. My goal is to automate this kind of lifting (or at least do it systematically).

One attempt I had in this direction is lifting lambda-calculus programs to SPLs. The assumption here is that an analysis is a program written in Typed Lambda Calculus (TLC), and with some type-directed code rewriting rules a lifted version of the analysis can be automatically generated. Examples include type checkers, static analyses and model checkers. The goal is to have an automatically lifted analysis with time complexity polynomial in the SPL size as opposed to naive liftings that have exponential time complexity in the number of features.

Lifting TLC would allow us to lift any program written in it, not just analyses. Program and model transformations in particular can be lifted to SPLs as well. In particular, we are working with General Motors (GM) to see how their Assurance Case analyses and transformations can be lifted to their SPLs.

\end{document}

### 2018-03-07 doc/quals/quals.tex
final version of the quals document

@@
 Another extends this work to more elaborate sampling techniques and more general categories of analysis~\cite{Liebig:2013}.
-At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs~\cite{Midtgaard:2015}.
+At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation analyses to SPLs~\cite{Midtgaard:2015}.
 At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}.
@@
 The evaluation presented in the paper justifies the hypotheses of the authors: the implemented variability-aware analyses under-perform single configuration analyses, but outperform the analysis of pair-wise samples.
-In addition, in some cases variability-aware analyses also outperform coverage-based analysis.
+In addition, in some cases variability-aware analyses also outperform coverage-based sampling.
 More importantly, variability-aware analyses are complete in terms of covering all valid feature combinations (i.e.
@@
 The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away).
-Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration.
-The speedup due to lifting with sharing also grows exponentially, but at a slower rate.
-Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only four source code methods).
+Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction is exponential with respect to the number of configurations (on the expense of accuracy), and almost identical to the average analysis performance of a single configuration.
+The speedup due to lifting with sharing is also exponential, but at a slower rate.
+Generalizing from those performance numbers is risky though because of the highly restricted sample space (a single analysis on only four source code methods).
 
@@
 
-K\"{a}stner et a.~\cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq~\cite{Bertot:2010} proof assistant.
+K\"{a}stner et al.~\cite{Kastner:2012} formalize and certify a Colored Featherweight Java (CFJ) type checker using the Coq~\cite{Bertot:2010} proof assistant.
 Featherweight Java (FJ) is a functional subset of Java, leaving out constructs such as assignment (mutation) and interfaces.
 CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE.
-This type checker is further extended to fully cover Java constructs.
+This type checker is then extended to fully cover Java constructs.
 
-A product line is considered well-typed only if all the products that can be generated out of it are well-typed (typing is preserved by product generation).
+An SPL is considered well-typed only if all the products that can be generated out of it are well-typed (typing is preserved by product generation).
 The type checker here takes the source code of a product line as input, and checks all variants for type errors.
-In addition to typing preservation, a second goal is backward compatibility.
-No new syntactic constructs were introduced, which allowed the use of existing tools.
+In addition to typing preservation, a secondary goal is backward compatibility with existing tools.
+To achieve the backward compatibility goal, no new syntactic constructs were introduced.
 
@@
 More importantly though, variability annotation is only supported within elements of syntactic sequences.
-For example, a sequence of methods (potentially empty) is defined.
-Each element (method) within that sequence can be annotated individually.
+For example, when a sequence of methods (potentially empty) is defined, each element (method) within that sequence can be annotated individually.
 However, variability is not supported for singleton syntactic constructs.
-This means that we cannot vary the subtype of a class or the return type of a method across products.
+This means that we cannot vary the subtype of a class or the return type of a method (among many other things) across products.
 
@@
 Some type errors were found by the checker in some product variants.
-The performance (in terms of time taken to type check the product line) was evaluated compared to approximate per-product performance, and on these 5 product lines processing time grew much slower than the number of variants.
-It is worth noting though that these product lines are relatively small, both in terms of number of lines of code and number of features.
+Performance (in terms of time taken to type check the product line) was evaluated compared to approximate per-product performance, and on those 5 product lines processing time grew much slower than the number of variants.
+It is worth noting though that those product lines are relatively small, both in terms of number of lines of code and number of features.
 
@@
 To prevent this exponential explosion, SuperC implements four optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces.
-With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.
+With those optimizations, for almost each of the Linux kernel compilation units the number of sub-parsers is less than 20.
 
@@
 
-\section{SPLLift}
+\newcommand{\spllift}{SPL\textsuperscript{Lift}}
+
+\section{\spllift}
 
@@
 In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems.
-The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions.
+The main idea behind lifting IFDS analyses is to guard data-flow properties of program statements by presence conditions.
 This fits well within the IFDS framework because data flow properties are boolean expressions.
@@
 The checker returns true if the property holds in the set of products defined by the property's presence condition.
-If the property is violated in any of those products, a set of counterexample/presence condition pairs is returned.
+If the property is violated in any of those products, a set of (counterexample, presence condition) pairs is returned.
 Each counterexample covers a set of products in which the property is violated, and each product in which the property is violated  has to be covered by a counterexample.
@@
 The SNIP model checker implements FTS model checking.
-Its input language is fPromela, which extends Promela (used in the SPIN model checker) by feature expression guard annotation over statements.
+Its input language is fPromela, which extends Promela (used in the SPIN model checker) with feature expression guard annotation over statements.
 It also implements variability-aware vacuity detection and deadlock detection.
@@
 This model is small compared to realistic models (only 457 FTS states), and the evaluation is done checking only 6 properties.
-With only one model (moreover, a small one), the evaluation results cannot be safely generalized.
 
@@
 Local variability representation is another principle that manages intermediate value variability.
-These principles together with some analysis-specific design decisions can effectively lead to an efficient family-based analysis.
+These principles together, in addition to some analysis-specific design decisions, can effectively lead to an efficient family-based analysis of an SPL.
 \bibliographystyle{abbrv}

### 2018-03-23 doc/quals/RelatedWork.tex
Related work section

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}

\begin{document}

\section{Related Work}

Apel et al.~\cite{Apel:2013} compare the performance of family-based SPL analyses to that of single-wise, pairwise and triple-wise sampling-based strategies. In addition to being complete with respect to covering all valid products, family-based analysis was shown to outperform triple-wise sampling-based strategies. This paper explicitly generated code for different variants (at the source code function level), with dispatch functions calling different variants based on feature expressions. In contrast, Liebig et al.~\cite{Liebig:2013} implemented family-based analyses for type checking and liveness analysis. They compared their analyses to more elaborate sampling strategies (e.g. coverage-based sampling), and again family-based analyses outperformed sampling.

Several attempts have been made to lift product-level analyses to SPLs. Midtgaard et al.~\cite{Midtgaard:2015} provide a thorough theoretical treatment of systematically (but not automatically) lifting abstract interpretation analyses to SPLs. They also touch upon the idea of abstracting variability to improve performance on the expense of accuracy. K\"{a}stner et al.~\cite{Kastner:2012} provide a variability-aware type checker for Feather-weight Java, with correctness proofs in Coq. In this paper they only support very limited forms of variability though. Gazzillo and Grimm~\cite{Gazzillo:2012} designed an efficient variability-aware C-language Fork-Merge-LR (FMLR) parser. It highly outperforms other attempts (e.g. the TypeChef C-language parser~\cite{Kastner:2011}), but all the optimizations were hand-crafted. Bodden et al.~\cite{Bodden:2013} provide a lifting of interprocedural IFDS~\cite{Reps:1995} data-flow analyses to SPLs. They exploit the fact that these analyses reduce to graph reachability, and annotate graph edges with presence conditions. Similarly, Classen et al.~\cite{Classen:2013} extend Labeled Transition Systems (LTSs) to Featured Transition Systems (FTSs) by annotating state transitions with presence conditions. In addition, they extended Linear Temporal Logic (LTL) with features (fLTL), and similarly the Promela model specification language with features (fPromela).

\bibliographystyle{abbrv}
\bibliography{quals}

\end{document}

### 2018-03-24 scala/spl/src/main/scala/spl/SPL.scala
lazy vals

@@ -5,25 +5,33 @@ import prop._
 object SPL {
   type PresenceCondition = Prop
 
-  type Val[T] = (T, PresenceCondition)
+  class Val[T] (
+    _v : => T,
+    _pc : PresenceCondition){
+    lazy val v :T = _v;
+    val pc :PresenceCondition = _pc;
+  }
 
-  type Var[T] = List[Val[T]]
+  type Var[T] = List[Val[T]];
 
-  def mkVarT[T](a : T) : Var[T] = List((a, True()))
+  def mkVarT[T](a : T) : Var[T] = List(new Val(a, True()))
 
   def normalize[A](xs : Var[A]) : Var[A] = {
     xs.map(x => {
-      val vs = xs.filter(v => x != v && x._1 == v._1)
+      val vs = xs.filter(v => x != v && x.v == v.pc)
       if (vs == Nil) x
       else {
         val h = vs.head
-        (x._1, Or(x._2, h._2))
+        new Val(x.v, Or(x.pc, h.pc))
       }
     })
   }
 
+  //def lifted_cond[T](c: Var[Boolean])(lazy a: Var[T])(lazy b: Var[T]) = Var[T] = {
+//
+  //}
   def apply[A,B](f : Var[A => B], a : Var[A]) : Var[B] = {
-    val vs = for (_f <- f; _a <- a; pc = And(_f._2, _a._2); if pc.isSAT()) yield (_f._1(_a._1), pc)
+    val vs = for (_f <- f; _a <- a; pc = And(_f.pc, _a.pc); if pc.isSAT()) yield new Val(_f.v(_a.v), pc)
     //normalize(vs)
     return vs
   }

### 2018-03-24 scala/spl/src/test/scala/spl/TSPL.scala

@@ -21,6 +21,9 @@ object Visitor {
   val lifted_map = SPL.mkVarT(map _)
   val lifted_foo = SPL.mkVarT(foo _)
 
+  //def liftedMap(f: SPL.Var[Int => Int])(xs: SPL.Var[List[Int]]) : SPL.Var[List[Int]] = {
+//
+//  }
   def visit(xs : SPL.Var[List[Int]]) : Int = {
     SPL.apply2(lifted_map, lifted_foo, xs)
     val ret = count
@@ -43,9 +46,9 @@ class SPLSuite extends FunSuite {
   val _pq = And(_p, q)
   val _p_q = And(_p, _q)
 
-  val x = List((7, pq), (-3, p_q), (-8, _pq), (0, _p_q))
-  val y = List((5, p), (-11, _p))
-  val z = List((6, p))
+  val x = List(new SPL.Val(7, pq), new SPL.Val(-3, p_q), new SPL.Val(-8, _pq), new SPL.Val(0, _p_q))
+  val y = List(new SPL.Val(5, p), new SPL.Val(-11, _p))
+  val z = List(new SPL.Val(6, p))
 
   val a = SPL.mkVarT(1)
   val b = SPL.mkVarT(2)
@@ -58,13 +61,13 @@ class SPLSuite extends FunSuite {
   test("plus") {
     val lifted_plus = SPL.mkVarT(plus _)
     val result = SPL.apply2(lifted_plus, x, y)
-    assert(result == List((12, pq), (2, p_q), (-19,_pq), (-11,_p_q)))
+    assert(result == List(new SPL.Val(12, pq), new SPL.Val(2, p_q), new SPL.Val(-19,_pq), new SPL.Val(-11,_p_q)))
   }
 
   test("cons") {
     val lifted_cons = SPL.mkVarT(cons _)
 
-    val l0 = SPL.mkVarT(Nil)
+    val l0 = SPL.mkVarT(List[Int]())
     val l1 = SPL.apply2(lifted_cons, a, l0)
     val l2 = SPL.apply2(lifted_cons, b, l1)
     val l3 = SPL.apply2(lifted_cons, c, l2)

### 2018-05-22 doc/paper.tex
rewinding

@@
-%% bare_conf.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
+\documentclass[10pt,conference]{llncs}
 
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-\documentclass[10pt,conference]{IEEEtran}
-% Some Computer Society conferences also require the compsoc mode option,
-% but others use the standard conference format.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/pkg/ifpdf
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of the IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 5.0 (2009-03-20) and later if using hyperref.sty.
-% The latest version can be obtained at:
-% http://www.ctan.org/pkg/cite
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at: 
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics.
-%
-% Note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/amsmath
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/pkg/algorithms
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/pkg/algorithmicx
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/array
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/pkg/subfig
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure.
-% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
-% corrections already built into the system in which case a warning will
-% be issued if an attempt is made to load fixltx2e.sty as it is no longer
-% needed.
-% The latest version and documentation can be found at:
-% http://www.ctan.org/pkg/fixltx2e
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/stfloats
-% Do not use the stfloats baselinefloat ability as the IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that the IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/pkg/dblfloatfix
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/url
-% Basically, \url{my_url_here}.
-
+\usepackage{graphicx}
 \usepackage[utf8]{inputenc}
 \usepackage[english]{babel}
-\usepackage{amsmath}
-\usepackage{amsthm}
+%\usepackage{amsmath}
+%\usepackage{amsthm}
 \usepackage{bussproofs}
@@
 
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+\newcommand\UrlFont{\color{blue}\rmfamily}
 
@@
 % affiliations
-\author{\IEEEauthorblockN{Ramy Shahin}
-\IEEEauthorblockA{Computer Science Department\\
+\author{{Ramy Shahin\inst{1}} \and
+	    {Marsha Chechik\inst{1}}}
+
+\institute{Computer Science Department\\
 	University of Toronto\\
-Email: ramy.shahin@mail.utoronto.ca}
-\and
-\IEEEauthorblockN{Marsha Chechik}
-\IEEEauthorblockA{Computer Science Department\\
-University of Toronto\\
-Email: chechik@cs.toronto.edu}}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
+	\email\{rshahin,chechik\}@cs.toronto.edu}
 
@@
 % no keywords
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
 
@@
 
-This is a summary of the TLC syntax and semantics from\cite{Pierce2002}.
+This is a summary of the TLC syntax and semantics from~\cite{Pierce2002}.
 
@@
 In the following description of the transformation rules we use Haskell syntax.
-This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as those constructs have standard definitions on top of TLC \cite{Pierce2002}.
+This way we can use several functional constructs (parametric types, lists, pairs) without loss of generality, as those constructs have standard definitions on top of TLC~\cite{Pierce2002}.
 
@@
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that the IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
-% labels, the optional argument to \subfloat must be present. If a
-% subcaption is not desired, just leave its contents blank,
-% e.g., \subfloat[].
-
-% An example of a floating table. Note that, for IEEE style tables, the
-% \caption command should come BEFORE the table and, given that table
-% captions serve much like titles, are usually capitalized except for words
-% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
-% and up, which are usually not capitalized unless they are the first or
-% last word of the caption. Table text will default to \footnotesize as
-% the IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that the IEEE does not put floats in the very first column
-% - or typically anywhere on the first page for that matter. Also,
-% in-text middle ("here") positioning is typically not used, but it
-% is allowed and encouraged for Computer Society conferences (but
-% not Computer Society journals). Most IEEE journals/conferences use
-% top floats exclusively. 
-% Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the
-% \fnbelowfloat command of the stfloats package.
-
 \section{Conclusion}
@@
 
-% conference papers do not normally have an appendix
-
-% use section* for acknowledgment
-%\section*{Acknowledgment}
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-
-\bibliography{paper} 
-\bibliographystyle{ieeetr}
+\bibliography{taxonomy} 
+\bibliographystyle{splnc04}
 

### 2018-05-22 haskell/SPL/src/PropBDD.hs

@@ -8,21 +8,21 @@ import Cudd.Cudd
 manager = cuddInit
 
 type Prop = DDNode
-type Universe = [(String,Prop)]
+type Universe = [Prop]
 
 mkUniverse :: [String] -> Universe
-mkUniverse xs = map (\(s,r) -> (s, ithVar manager r)) (zip xs [0..])
+mkUniverse xs = map (\(s,r) -> (ithVar manager r)) (zip xs [0..])
 
 queryOrUpdate :: Universe -> String -> (Universe, Prop)
 queryOrUpdate [] s = (u', p)
     where   u' = mkUniverse [s]
-            p  = snd $ head u' 
+            p  = head u' 
 
-queryOrUpdate u@((s',p'):u') s =
-    if s == s' 
-    then (u,p')
-    else ((s',p'):u'', p'')
-         where (u'',p'') = queryOrUpdate u' s
+--queryOrUpdate u@((s',p'):u') s =
+--    if s == s' 
+--    then (u,p')
+--    else ((s',p'):u'', p'')
+--         where (u'',p'') = queryOrUpdate u' s
          
 tt = readOne manager
 ff = readLogicZero manager

### 2018-05-22 haskell/SPL/test/SPL_prop.hs

@@ -4,7 +4,7 @@ module SPL_prop where
 
 import Test.QuickCheck.All
 import SPL
-import Prop
+import PropBDD
 import Shallow.VList as S
 import Deep.VList as D
 import Control.Applicative
